### 2019-01-28 paper/paper.tex
paper.tex: outline

% $Id$
\documentclass[onecolumn,prd,nofootinbib]{revtex4}
\input{refsheetLinks.tex}
\begin{document}

\title{Accelerating parameter inference with graphics processing units}
\author{D. Wysocki}
\author{L. Fang}
\author{R. O'Shaughnessy}
\begin{abstract}
Gravitational wave Bayesian parameter inference involves repeated comparisons of GW data to generic candidate predictions.
Even with algorithmically efficient methods like RIFT or reduced-order quadrature, the time needed to perform these
calculations and overall computational cost can be significant compared to the minutes to hours needed to achieve the
goals of low-latency multimessenger astronomy.  By translating some elements of the RIFT algorithm to operate on
graphics processing units (GPU), we demonstrate substantial performance improvements, enabling dramatically reduced
overall cost and latency. 
\end{abstract}
\maketitle

\section{Introduction}

\section{Methods}

* Review ILE

* Rewrite in matrix form

* GPU code

* Refactoring code to use the efficient GPU loops

** discuss tradeoffs: how to balance overall runtime.  When the cost of the waveform integration matters.

\section{Results}

* Comparison: agreement.  Sky localization example

* Benchmarking performance (new speed for BBHs; also show BNS with tides)

\section{Conclusions}

* Enable low-latency parameter inference in basically real time for GW candidate sources, even for relatively costly
models with full physics

* Highly portable, generalizable to other codes

\bibstyle{unsrt}
\bibliography{paperexport}
\end{document}

### 2019-01-28 paper/paper.tex
paper: updating

@@
 % $Id$
-\documentclass[onecolumn,prd,nofootinbib]{revtex4}
-\input{refsheetLinks.tex}
+\documentclass[twocolumn,prd,nofootinbib]{revtex4}
+
+\newcommand\ForInternalReference[1]{}
+%\newcommand\SkipForEarlyCirculation[1]{}
+\newcommand\SkipForEarlyCirculation[1]{#1}
+%\newcommand\SkipBoxing[1]{}
+\newcommand\SkipBoxing[1]{#1}
+\newcommand\SkipPP[1]{}
+\usepackage{verbatim}
+\usepackage{graphicx}
+\usepackage{dcolumn}
+\usepackage{bm}
+\usepackage{color}
+\usepackage{xspace}
+\usepackage{url}
+\usepackage{amsmath}
+%\usepackage{adjustbox}
+\usepackage{float}
+\usepackage{multirow}
+%
+%
+\usepackage{times}
+%
+%
+%
+\newcommand\optional[1]{}
+
+%
+\newcommand\E[1]{\left\langle #1\right\rangle}
+\newcommand\qmstate[1]{\left|#1\right \rangle}
+\newcommand\qmstateKet[1]{\left\langle#1\right|}
+\newcommand\qmstateproduct[2]{\left\langle#1|#2\right\rangle}
+\newcommand\qmoperatorelement[3]{\left\langle#1\left|#2\right|#3\right\rangle}
+\newcommand\qmoperator[1]{{\bf #1}}
+%
+\newcommand\Y[1]{{{}_{#1}Y}}
+
+\newcommand\lnL{ \ln {\cal L}}
+\newcommand\lnLmarg{ \ln{\cal L}_{\rm marg}{}}
+\newcommand\unit[1]{{\rm #1}}
+
+\newcommand\rapidPEOrig{rapid\_pe1}
+\newcommand\ILE{ILE}
+\newcommand\editremark[1]{{\color{red} #1}}
+%
+%
+%
+\usepackage{color}
+\definecolor{amber}{rgb}{1.0, 0.75, 0.0}
+\definecolor{orange}{rgb}{1.0, 0.5, 0.0}
+\definecolor{amaranth}{rgb}{0.9, 0.17, 0.31}
+\def\fixme#1{\textcolor{red}{#1}}
+\newcommand{\Richard}[1]{ {\color{blue}{#1}}}
+\newcommand{\ros}[1]{ {\color{blue}{#1}}}
+%
+
+%
+
+%
+%
+%
+%
+\graphicspath{{./figures/}}
+\newcommand{\mc}{{\cal M}}
+\newcommand{\Ms}{M_{\odot}}
+\newcommand\LambdaTilde{\widetilde{\Lambda}}
+\newcommand\DeltaLambdaTilde{\delta \widetilde{\Lambda}}
+%
+\def\ltsima{$\; \buildrel < \over \sim \;$}
+\def\simlt{\lower.5ex\hbox{\ltsima}}
+\def\gtsima{$\; \buildrel > \over \sim \;$}
+\def\simgt{\lower.5ex\hbox{\gtsima}}
+
+\newcommand\prx{Phys.~Rev.~X}
+\def\aj{Astronomical Journal}                 %
+\def\apj{Astrophysical Journal}                %
+\def\apjl{Astrophysical Journal}             %
+\def\pasj{PASJ}
+\def\apjs{ApJS}              %
+\def\mnras{MNRAS}            %
+\def\prd{Phys.~Rev.~D}       %
+\def\prl{Phys.~Rev.~Lett}    %
+\def\cqg{Class.~Quant.~Grav.~}%
+\def\araa{ARA\&A}             %
+\def\nat{Nature}              %
+\def\aap{A\&A}                %
+\def\aapr{A\&A~Rev.~}    %
+\def\pasp{PASP}    %
+\def\sovast{Soviet Ast.}
+%
+%
+
+\newcommand{\IMRPD}{\textsc{IMRPhenomD}\xspace}
+\newcommand{\IMRPDT}{\textsc{IMRPhenomD\_NRTidal}\xspace}
+\newcommand{\IMRP}{\textsc{IMRPhenomPv2}\xspace}
+\newcommand{\SEOBP}{\textsc{SEOBNRv3}\xspace}
+\newcommand{\SEOBA}{\textsc{SEOBNRv4}\xspace}
+\newcommand{\SEOBAROM}{\textsc{SEOBNRv4\_ROM}\xspace}
+\newcommand{\NRSur}{NRSur7dq2\xspace}
+\newcommand{\TEOB}{SEOBNRv4T\xspace}
+\newcommand{\Resum}{TEOBResumS\xspace}
+\newcommand\RIFT{RIFT}
+\newcommand{\Taylor}{TaylorF2\xspace}
+\newcommand\PaperDetection{\underline{LVC-detect}\cite{DiscoveryPaper}}
+\newcommand\PaperPE{\underline{LVC-PE}\cite{PEPaper}}
+\newcommand\PaperTestGR{\underline{LVC-TestGR}\cite{TestingGRPaper}}
+\newcommand\PaperPENRMethods{\underline{PE+NR-Methods}\cite{gwastro-PENR-Methods-Lange}}
+\newcommand\PaperAstro{\underline{LVC-Astro}\cite{AstroPaper}}
+\newcommand\PaperBurst{\underline{LVC-Burst}\cite{BurstPaper}}
+\newcommand\PaperRates{\underline{LVC-Rates}\cite{RatesPaper}}
+\newcommand\PaperStochastic{\underline{LVC-Stochastic}}
+\newcommand\PaperSEOBNRvthree{\underline{LVC-SEOBNRv3}\cite{SEOBv3Paper}}
+
 \begin{document}
@@
 \author{R. O'Shaughnessy}
+\author(??Jake - if he does runs??}
 \begin{abstract}
@@
 \section{Introduction}
+The Advanced LIGO  \cite{2015CQGra..32g4001T}  and Virgo \cite{gw-detectors-Virgo-original-preferred}  ground-based gravitational wave (GW) detectors have
+identified several coalescing compact binaries
+\cite{DiscoveryPaper,LIGO-O1-BBH,2017PhRvL.118v1101A,LIGO-GW170814,LIGO-GW170608,LIGO-GW170817-bns}.  
+%
+The properties of the sources responsible have been inferred via Bayesian comparison of  each source with candidate
+gravitational ave signals
+\cite{DiscoveryPaper,LIGO-O1-BBH,2017PhRvL.118v1101A,LIGO-GW170814,LIGO-GW170608,LIGO-GW170817-bns}
+\editremark{citations: real events, also LI and RIFT} \cite{gwastro-PE-AlternativeArchitectures,gwastro-PENR-RIFT}.  
+Many more are expected as observatories reach design sensitivity \cite{2016LRR....19....1A}.  
+Both to handle the rapid pace of discovery and particularly to enable coordinated multimessenger followup, GW
+observatories should be able to reconstruct the full source parameters of coalescing binaries as fast as possible
+\cite{LIGO-whitepaper-2015} \editremark{update}.    
+However, particularly when using the best-available waveform models, these calculations can be very costly.  In
+practice, detailed inferences even using simplified waveforms take weeks for binary neutron stars, and can take months
+for costly time-domain effective-one-body models which incorporate the effects of precession.
+
+% BORROWED FROM ROM: edit
+Several strategies have been
+developed to reduce the 
+%overall computational cost and turnaround time of 
+computational cost of parameter estimation \cite{gwastro-pe-Brandon-STF2,gwastro-PE-AlternativeArchitectures,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,2016PhRvD..94d4031S,2017arXiv170302062V}.
+% The computational cost of parameter estimation can be reduced 
+Approaches that have appeared in the literature include
+%by 
+generating the approximate solutions more quickly \cite{gwastro-mergers-PE-ReducedOrder-2013,2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C,gwastro-mergers-IMRPhenomP,gwastro-SpinTaylorF2-2013};
+%by 
+interpolating some combination
+of the waveform or likelihood \cite{gwastro-approx-ROMNR-Blackman2015,2013PhRvD..87d4008C,2013PhRvD..87l2002S,2014CQGra..31s5010P,2014PhRvD..90d4074S,gwastro-PE-AlternativeArchitectures,cole2014likelihood,2012MNRAS.421..169G};  or
+%by 
+adopting a sparse representation to reduce the computational cost of data
+handling \cite{antil2013two,gwastro-mergers-PE-ReducedOrder-2013,2016PhRvD..94d4031S,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,gwastro-PE-AlternativeArchitectures}.
+% 2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C
+ Some methods, however, achieve rapid turnaround through simplifying approximations.  
+%
+The RIFT strategy described in \editremark{citations - include ROM} eschews these simplifications, performing
+embarassingly-parallel inferences even for costly models.   However, the method as previously implemented still had a
+significant net  computational cost and noticable startup time, limiting its diverse potential applications.
+
+In this paper we describe a substantial improvement to the RIFT approach, enabled by translating its low-level
+algorithms to operate on graphics processing units (GPUs).
+
+This paper is organized as follows.
+In Section \ref{Methods}, we review the underlying marginalized likelihood calculations used by RIFT; their updated GPU
+implementation; and the operating settings which enable substantially increased performance.
+In Section \ref{sec:demo}, we quantify the improved performance of our  GPU-accelerated code, and its agreement with the
+original CPU-only version.
 
@@
 
-* Review ILE
+\subsection{Parameter inference with the RIFT likelihood}
+% From RIFT paper
+\ILE{}  -- a specific algorithm to ``integrate over extrinsic parameters'' -- provides a straightforward and efficient mechanism to compare any specific candidate gravitational wave source with
+real or synthetic data   \cite{gwastro-PE-AlternativeArchitectures,NRPaper,2017PhRvD..96j4041L,2017CQGra..34n4002O},
+by marginalizing the likelihood of the data over the seven coordinates characterizing the spacetime coordinates and
+orientation of the binary relative to the earth.  
+Specifically the likelihood of the data given gaussian noise has the form  (up to normalization)
+\begin{equation}
+\label{eq:lnL}
+\ln {\cal L}(\bm{\lambda} ;\theta )=-\frac{1}{2}\sum\limits_{k}\langle h_{k}(\bm{\lambda} ,\theta )-d_{k} |h_{k}(\bm{\lambda} ,\theta )-d_{k}\rangle _{k}-\langle d_{k}|d_{k}\rangle _{k},
+\end{equation}
+where $h_{k}$ are the predicted response of the k$^{th}$ detector due to a source with parameters ($\bm{\lambda}$, $\theta$) and
+$d_{k}$ are the detector data in each instrument k; $\bm{\lambda}$ denotes the combination of redshifted mass $M_{z}$ and the
+remaining parameters needed to uniquely specify the binary's dynamics; $\theta$ represents the
+seven extrinsic parameters (4 spacetime coordinates for the coalescence event and 3 Euler angles for the binary's
+orientation relative to the Earth); and $\langle a|b\rangle_{k}\equiv
+\int_{-\infty}^{\infty}2df\tilde{a}(f)^{*}\tilde{b}(f)/S_{h,k}(|f|)$ is an inner product implied by the k$^{th}$ detector's
+noise power spectrum $S_{h,k}(f)$. 
+%In all calculations, we adopt the fiducial O1 noise power spectra associated with data near GW150914 \cite{DiscoveryPaper}.
+In practice we adopt both  low- and high- frequency cutoffs $f_{\rm max},f_{\rm min}$ so all inner products are modified to
+\begin{equation}
+\label{eq:overlap}
+\langle a|b\rangle_{k}\equiv 2 \int_{|f|>f_{\rm min},|f|<f_{\rm max}}df\frac{[\tilde{a}(f)]^{*}\tilde{b}(f)}{S_{h,k}(|f|)}.
+\end{equation}
+The joint posterior probability of $\bm{\lambda} ,\theta$ follows from Bayes' theorem:
+\begin{equation}
+p_{\rm post}(\bm{\lambda} ,\theta)=\frac{ {\cal L}(\bm{\lambda} ,\theta)p(\theta)p(\bm{\lambda})}{\int d\bm{\lambda} d\theta {\cal L}(\bm{\lambda} ,\theta)p(\bm{\lambda})p(\theta)},
+\end{equation}
+where $p(\theta)$ and $p(\bm{\lambda})$ are priors on the (independent) variables $\theta ,\bm{\lambda}$. For each $\bm{\lambda}$, we evaluate the marginalized likelihood
+\begin{equation}
+ {\cal L}_{\rm marg}\equiv\int  {\cal L}(\bm{\lambda} ,\theta )p(\theta )d\theta
+\end{equation}
+via direct Monte Carlo integration, where $p(\theta)$ is uniform in 4-volume and source orientation.  
+The integral in time is performed at a low level, by direct quadrature. 
+For the remaining dimensions, to evaluate the likelihood in regions of high importance, we use an adaptive Monte Carlo as described in
+\cite{gwastro-PE-AlternativeArchitectures}.    As described in Pankow et al, this marginalized likelihood can be evaluated efficiently
+because, having generated the dynamics and outgoing radiation in all possible directions once and for all for fixed
+$\mathbf{\lambda}$, the likelihood can be evaluated as a function of $\theta$ at very low computational cost.
 
-* Rewrite in matrix form
+% From ROM paper
+A dimensionless, complex gravitational-wave
+strain
+%% ~\footnote{\scott{\sout{We use a distance-independent dimensionless strain $R h/M$, where
+%% $R$ is the distance from the binary's center-of-mass
+%% and $M$ is the total mass. In this paper we choose units so that $c=G=1$.}
+%% [Remove this? I think it conflicts with other conventions/definitions 
+%% used throughout the paper.]}}
+\begin{align} \label{eq:strain}
+h(t,\vartheta,\phi;\lambda) =  h_+(t,\vartheta,\phi;\lambda) - 
+                                i h_\times (t,\vartheta,\phi;\lambda) \, ,
+\end{align}
+can be expressed in terms of its two fundamental polarizations $h_+$ and $h_\times$.
+Here, $t$ denotes time, $\vartheta$ and $\phi$ are the polar and azimuthal angles
+for the direction of gravitational wave propagation away from the source. 
+The complex gravitational-wave strain can be written in terms of
+spin-weighted spherical harmonics $\Y{-2}_{\ell m} \left(\vartheta, \phi \right)$ as 
+\begin{align} \label{eq:strain_mode}
+h(t,\vartheta,\phi;\lambda) = 
+\sum_{\ell=2}^{\infty} \sum_{m=-\ell}^{\ell} \frac{D_{\rm ref}}{D} h^{\ell m}(t;\lambda) \Y{-2}_{\ell m} \left(\vartheta, \phi \right) \, ,
+\end{align}
+where the sum includes all available harmonic modes $h^{\ell m}(t;\pmb{\lambda})$ made available by the model;  where
+$D_{\rm ref}$ is a fiducial reference distance; and where $D$, the luminosity distance to the  source, is one of the
+extrinsic parameters.
 
-* GPU code
+Following \citet{gwastro-PE-AlternativeArchitectures}, we substitute expression~\eqref{eq:strain_mode} 
+for $h_{\ell m}$ into the expression $h_k(t) =F_{+,k} h_+(t_k) +
+  F_{\times,k}h_\times(t_k)$ for the detector response $h_k$, 
+where $t_k=t_c - \vec{x}_k \cdot \hat{n}$ is the arrival time at the $k$th detector (at position $\vec{x}_k$)
+for a plane wave propagating along $\hat{n}$ \cite{gwastro-PE-AlternativeArchitectures}.
+We then substitute these expressions for $h_k$ into the likelihood function~\eqref{eq:loglikelihood}
+thereby generating~\cite{gwastro-PE-AlternativeArchitectures}
+\begin{widetext}
+\begin{align}
+\ln {\cal L}(\lambda, \theta) 
+&= (D_{\rm ref}/D) \text{Re} \sum_k \sum_{\ell m}(F_k \Y{-2}_{\ell m})^* Q_{k,lm}(\lambda,t_k)\nonumber \\
+&   -\frac{(D_{\rm ref}/D)^2}{4}\sum_k \sum_{\ell m \ell' m'}
+\left[
+{
+|F_k|^2 [\Y{-2}_{\ell m}]^*\Y{-2}_{\ell'm'} U_{k,\ell m,\ell' m'}(\lambda)
+}
+% \right. \nonumber \\ & \left.
+ {
++  \text{Re} \left( F_k^2 \Y{-2}_{\ell m} \Y{-2}_{\ell'm'} V_{k,\ell m,\ell'm'} \right)
+}
+\right]
+\label{eq:def:lnL:Decomposed}
+\end{align}
+\end{widetext}
+%% \noindent \scott{[FOR RICHARD: is eq 9/10 missing objects? Looks like $t_c$ an $F_k$ just appeared. On the other hand,
+%%     putting them into earlier equations will complicate the discussion and is awkward.]}  
+where 
+ where $F_k = F_{+,k} - i F_{\times,k}$ are the
+complex-valued detector
+response functions of the $k$th detector \cite{gwastro-PE-AlternativeArchitectures} and 
+the quantities $Q,U,V$ depend on $h$ and the data as
+\begin{subequations}
+%\label{eq:ComputeRhoViaInnerProductMatrix}
+\label{eq:QUV}
+\begin{align}
+Q_{k,\ell m}(\lambda,t_k) &\equiv \qmstateproduct{h_{\ell m}(\lambda,t_k)}{d}_k \nonumber\\
+&= 2 \int_{|f|>f_{\rm low}} \frac{df}{S_{n,k}(|f|)} e^{2\pi i f t_k} \tilde{h}_{\ell m}^*(\lambda;f) \tilde{d}(f)\ , \\
+{ U_{k,\ell m,\ell' m'}(\lambda)}& = \qmstateproduct{h_{\ell m}}{h_{\ell'm'}}_k\ , \\
+V_{k,\ell m,\ell' m'}(\lambda)& = \qmstateproduct{h_{\ell m}^*}{h_{\ell'm'}}_k  \ .
+\end{align}
+\end{subequations}
+Rewriting sums as matrix operations, the likelihood can be equivalently expressed as
+\begin{align}
+\ln {\cal L} &= \frac{D_{\rm ref}}{D} \text{Re}[ (F Y)^\dag Q]  \nonumber \\
+ & - \frac{D_{\rm ref}^2}{4 D^2} [ (FY)^\dag U FY + (FY)^TV FY]
+\label{eq:lnL:MatrixForm}
+\end{align}
+where $F,D$ are arrays over extrinsic parameters; $Q$ is an array over time and $(l,m)$; and $U,V$ ar matricies over
+$(l,m)$.
 
-* Refactoring code to use the efficient GPU loops
+\subsection{Accelerated evaluation via efficient multiplication}
+%* GPU code
+%* Refactoring code to use the efficient GPU loops
 
-** discuss tradeoffs: how to balance overall runtime.  When the cost of the waveform integration matters.
+At a fixed set of intrinsic parameters $\lambda$, \ILE{} will repeatedly evaluate the time-marginalized likelihood 
+for each candiate Monte Carlo set of extrinsic parameters $\theta$:
+\begin{eqnarray}
+{\cal L}_{\rm margT} \equiv  \int {\cal L} \frac{dt}{T}
+\label{eq:lnL:tmarg}
+\end{eqnarray}
+where $T$ is a small coincidence window consistent with the cadidate event time; here and previously,
+$T=0.3\unit{sec}$.   To accelerate the code, after precomputing the inner products $U,V,Q$, we simply shift them to the
+graphics card, then carry out all calculations necessary to implement Eqs. (\ref{eq:lnL:MatrixForm}, \ref{eq:lnL:tmarg})
+on the GPU.  To enable this implementation with portable code, we use \texttt{cupy}, a drop-in-replacement for equivalent
+\texttt{numpy} code used for the CPU-based version of ILE.    For the most costly part of the calculation -- the inner
+products necessary to evaluate $Q_{lm}(t)$ -- we use a custom CUDA kernel.  
+%
+With these changes alone, the likelihoood evaluation is roughly $60\times$ faster on equivalent hardware.  
+After this update, likelihood evaluation costs are not the performance- or cost-limiting feature of RIFT-based source
+parameter inference.
 
-\section{Results}
+\subsection{Exploiting low-cost likelihoods}
+Previously, each instance of ILE examined one intrinsic point $\lambda$.  The overall cost of this ILE evaluation
+involved two parts: a startup cost, a setup cost, and a Monte Carlo integration cost.  The  startup cost $\tau_{start}$ is associated with code setup, followed by reading and Fourier-transforming the
+data.  The setup cost $\tau_{setup}$ arises from waveform generation, followed by the inner products $U,V,Q$.  Finally, the Monte Carlo
+cost $\tau_{mc}=N_{it} \tau_{it}$ increases linearly with the number of Monte Carlo iterations $N_{it}$. 
+The choice of one $\lambda$ for each instance of ILE was eue to the substantial time  $\tau_{MC}$, which vastly
+dominated the overall computational cost.   For example, for a typical analysis of a short signal, \editremark{numbers}
+
+With the new low-cost likelihood evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now are
+comparable or in excess of the Monte Carlo evaluation time.   For this reason, we reorganize the workflow, so each
+instance of ILE loops over
+$N_{eval}$ different choices for $\lambda$. 
+%
+Additionally, particularly for massive binary black holes, we adopt two additional performance improvements.  First and
+foremost, we lower the sampling rate, thus lowering the startup cost $\tau_{start}$ and particularly setup cost
+$\tau_{setup}$.  Previously and out of an abundance of caution, \ILE{} employed a sampling rate $1/\Delta t = 16384
+\unit{Hz}$ for all calculations, but terminated all inner products in Eq. (\ref{eq:overlap}) at roughly $f_{\rm
+  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ decreases the  cost $\tau_{setup}$ by the
+same factor. 
+Also, to insure safe inner products over short data sets in the presence of very narrow spectral lines, we operated on a
+significantly-padded data buffer,
+using 
+``inverse spectrum truncation'' \editremark{citation}.  For binary black holes, the degree of padding was significantly
+in excess of the signal duration \editremark{numbers: say 64s for a BBH!}.  
+Instead and following LI, we adopt a much shorter inverse spectrum truncation length, a much shorter padding buffer, and
+a Tukey window applied to the data to remove discontinuities at the start and end of each segment.   
+Reducing the duration of data analyzed by a factor $s'$ will reduce the cost $\tau_{setup}$ by a factor $s'$.  
+%
+Combining these factors together, the overall computational cost of each marginalized likelihood
+evaluation is
+\begin{align}
+T_{ILE}/N_{\rm eval} = \tau_{start}/N_{\rm eval} + [ \tau_{setup}/ss' + N_{it} \tau_{it}]
+\end{align}
+\editremark{typical numbers in new configuration, using 16kHz, and also the $s,s'$ factors}
+
+\editremark{discussion}
+* discuss tradeoffs: how to balance overall runtime.  When the cost of the waveform integration matters.
+
+\section{Validation and Performance}
+\label{sec:demo}
 
@@
 * Benchmarking performance (new speed for BBHs; also show BNS with tides)
+
+\section{ Results (optional?)}
+
+* Add Jake as author.  Rerun all O2 BBHs with SEOBNRv3?
 
@@
 
-* Highly portable, generalizable to other codes
+* Highly portable, generalizable to other codes: code will smoothly tranlate between GPU-only and CPU-only mode
+
+* Not done yet: expect to transform MC and inner products, so all calculations remain on board, with just intermittent
+control I/O from CPU
 

### 2019-01-28 paper/paper.tex
paper.tex: updating draft

@@
 
+\def\RIT{Center for Computational Relativity and Gravitation, Rochester Institute of Technology, Rochester, New York 14623, USA}
+
 \begin{document}
@@
 \author{D. Wysocki}
+\affiliation{\RIT}
 \author{L. Fang}
+\affiliation{Brookhaven National Lab, Brookhaven NY}
 \author{R. O'Shaughnessy}
-\author(??Jake - if he does runs??}
+\affiliation{\RIT}
+\author{Jake - if he does runs}
+\affiliation{\RIT}
 \begin{abstract}
@@
 With these changes alone, the likelihoood evaluation is roughly $60\times$ faster on equivalent hardware.  
-After this update, likelihood evaluation costs are not the performance- or cost-limiting feature of RIFT-based source
-parameter inference.
+After this update, individual likelihood evaluation costs are not the performance- or cost-limiting feature of RIFT-based source
+parameter inference.  Instead, the overhead associated with the adaptive Monte Carlo and with the (CPU-based) inner
+product evaluations for $Q,U,V$ dominate our computational cost.
 
@@
 The choice of one $\lambda$ for each instance of ILE was eue to the substantial time  $\tau_{MC}$, which vastly
-dominated the overall computational cost.   For example, for a typical analysis of a short signal, \editremark{numbers}
+dominated the overall computational cost.   For example, for a typical analysis of a short signal --  a typical binary
+black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$ --  this version
+\editremark{cost table}
 
@@
 Combining these factors together, the overall computational cost of each marginalized likelihood
-evaluation is
+evaluation $T_{ILE}/N_{eval}$  is
 \begin{align}
-T_{ILE}/N_{\rm eval} = \tau_{start}/N_{\rm eval} + [ \tau_{setup}/ss' + N_{it} \tau_{it}]
+T_{ILE}/N_{\rm eval} &= \tau_{start}/N_{\rm eval} + [ \tau_{setup}/ss' + N_{it} \tau_{it}] 
 \end{align}
-\editremark{typical numbers in new configuration, using 16kHz, and also the $s,s'$ factors}
+For a typical binary black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$, this
+version requires $\tau_{start} \simeq 12\unit{s}$, $\tau_{setup}/ss' \simeq XX$, and $N_{it}\tau_{\it} \simeq XX$
+\editremark{numbers}.
+Typically we target $N_{\rm eval} \simeq 20$ so that startup costs are not a significant contributor to the overall
+runtime.   In this configuration, the run time per marginalized likelihood is around $T_{ILE}/N_{\rm eval} \simeq 20
+\unit{s}$ for a binary black hole \editremark{checkme - old number}.
 
-\editremark{discussion}
-* discuss tradeoffs: how to balance overall runtime.  When the cost of the waveform integration matters.
+For low-latency analyses, we need to assess the overall wallclock time needed to perform an analysis, often with a
+restricted number $N_{GPU}$ of available GPU-enabled machines.   To  completely $N_{\rm net}$ likelihood evaluations
+with these resources requires
+\begin{align}
+T_{net} &= \frac{N_{\rm net}}{N_{\rm GPU} N_{\rm eval}} T_{ILE} \nonumber \\
+& \simeq
+  \frac{N_{net}}{N_{GPU}}\left[\tau_{start}/N_{\rm    eval} + [ \tau_{setup}/ss' + N_{it} \tau_{it}]  \right]
+\end{align}
+Typically we target $N_{\rm eval}  \lesssim 3\times 10^4, N_{\rm GPU} \simeq 100$, or roughly $300\times$ the cost per
+individual marginalized likelihood evaluation.  For the binary black hole configuration described above, marginalized
+likelihood evaluations will complete in about $100 \unit{minutes}$ \editremark{checkme}. This number can be reduced to
+tens of minutes with a modestly larger GPU pool.
+This discussion ignores the latency introduced by the  gaussian process interpolation stage at the end.  We will revisit
+accelerated GP interpolation in subsequent work.
+
+The \ILE{} likelhood  generates waveform dynamics and $h_{lm}(t)$ once per evaluation $\lambda$.   Waveform generation can contribute significantly to the
+time needed to evaluate  $\ln {\cal L}_{\rm marg}$ for extremely costly
+waveform models which require $\tau_{wf} $ a significant fraction of $  \tau_{setup}$.   \editremark{give examples}
 
@@
 \label{sec:demo}
+
+\begin{figure}
+
+\end{figure}
 
@@
 
-\bibstyle{unsrt}
+\begin{acknowledgements}
+RO'S is supported by NSF PHY-1707965 and PHY-1607520.
+DW thanks the RIT COS and CCRG for support, and  NSF-1707965.
+\end{acknowledgements}
+
+%\bibstyle{unsrt}
 \bibliography{paperexport}

### 2019-01-28 paper/paper.tex
Leo affiliation, funding, and acknowledgment

@@
 \affiliation{\RIT}
-\author{L. Fang}
-\affiliation{Brookhaven National Lab, Brookhaven NY}
+\author{Yao-Lung L. Fang}
+%\author{Yao-Lung L. Fang (方耀龍)}
+\affiliation{Computational Science Initiative, Brookhaven National Laboratory, Upton, NY 11973, USA}
+\affiliation{National Synchrotron Light Source II, Brookhaven National Laboratory, Upton, NY 11973, USA}
 \author{R. O'Shaughnessy}
@@
 DW thanks the RIT COS and CCRG for support, and  NSF-1707965.
+Y-LLF was supported in part by BNL LDRD projects No.\,17-029 and No.\,19-002, 
+and New York State Urban Development Corporation, d/b/a Empire State Development, under contract No.\,AA289.
+We thank the Computational Science Initiative at BNL for hosting GPU Hackathon 2018, during which a part of this work was conducted.
+We also thank Jeff Layton at NVIDIA for bringing CuPy to our attention.
 \end{acknowledgements}

### 2019-01-28 paper/paper.tex
paper.tex: adding more text

--- 2019-01-28 paper/paper.tex
@@
 At a fixed set of intrinsic parameters $\lambda$, \ILE{} will repeatedly evaluate the time-marginalized likelihood 
-for each candiate Monte Carlo set of extrinsic parameters $\theta$:
+for each candidate Monte Carlo set of extrinsic parameters $\theta$:
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $T$ is a small coincidence window consistent with the cadidate event time; here and previously,
+where $T$ is a small coincidence window consistent with the candidate event time; here and previously,
 $T=0.3\unit{sec}$.   To accelerate the code, after precomputing the inner products $U,V,Q$, we simply shift them to the
@@
 %
-With these changes alone, the likelihoood evaluation is roughly $60\times$ faster on equivalent hardware.  
+With these changes alone, the likelihood evaluation is roughly $60\times$ faster on equivalent hardware.  
 After this update, individual likelihood evaluation costs are not the performance- or cost-limiting feature of RIFT-based source
@@
 black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$ --  this version
-\editremark{cost table}
+\editremark{cost table}, based on the assumption that the MC terminates after $N_{\rm it}\simeq 10^6$ iterations.  
+By contrast, the marginalized likelihood described above  converges incredibly rapidly if the candidate model is
+inconsistent with a signal (or the absence thereof) in the data, with only $N_{it} \simeq O(10^4)$  evaluations needed to identify the marginalized
+likleihood.
 
@@
 
+\subsection{Tradeoff between Monte Carlo integration and accuracy}
+
+Each marginalized likelihood evaluation has a relative uncertainty of order  $1/\sqrt{n_{\rm eff}}$, where the number of
+effective samples $n_{\rm eff}=\epsilon N_{\rm it}$ increases linearly with the total number $N_{\rm it}$ of Monte Carlo
+evaluations performed by \ILE{}.  Therefore, to increase (decrease) our accuracy for each likelihood evaluation by a factor of $A$
+will require a factor $A^2$ more (fewer) iterations.   
+%
+We adopt a fixed threshold on $n_{\rm eff}$.
+
+We do benefit slightly by re-using the adaptive Monte Carlo intergrator for each extrinsic point $\lambda$.  Since the
+integrator has already identified the likely range of sky locations and distances on the first iteration, each
+subsequent evaluation of the marginalized likelihood can converge marginally more quickly.
+
 \section{Validation and Performance}

### 2019-01-30 paper/paper.tex
paper.tex: adding table, connecting table and discussion to actual profiling results

@@
 Previously, each instance of ILE examined one intrinsic point $\lambda$.  The overall cost of this ILE evaluation
-involved two parts: a startup cost, a setup cost, and a Monte Carlo integration cost.  The  startup cost $\tau_{start}$ is associated with code setup, followed by reading and Fourier-transforming the
-data.  The setup cost $\tau_{setup}$ arises from waveform generation, followed by the inner products $U,V,Q$.  Finally, the Monte Carlo
-cost $\tau_{mc}=N_{it} \tau_{it}$ increases linearly with the number of Monte Carlo iterations $N_{it}$. 
+involved two parts: a startup cost, a setup cost, and a Monte Carlo integration cost.  The  startup cost $\tau_{start}$
+is associated with code setup followed by reading, data conditioning, and Fourier-transforming the
+data.  The setup cost $\tau_{setup}$ arises from waveform generation and  the inner products $U,V,Q$.  Finally, the Monte Carlo
+cost $\tau_{mc}=N_{ad} \tau_{ad}+N_{it}\tau_{eval}$ increases with the number of Monte Carlo iterations $N_{it}$ with
+and the number of times $N_{ad}$ the sampling prior used in adaptive Monte Carlo is regenerated, in proportion to their
+cost.   In the standard configuration, the adaptive sampling prior is regenerated every $n_{\rm chunk}$ iterations (i.e,
+ $N_{ad} \simeq N_{it}/n_{chunk}$), regenerating a sampling prior in two sky location parameters and distance.  
 The choice of one $\lambda$ for each instance of ILE was eue to the substantial time  $\tau_{MC}$, which vastly
 dominated the overall computational cost.   For example, for a typical analysis of a short signal --  a typical binary
-black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$ --  this version
-\editremark{cost table}, based on the assumption that the MC terminates after $N_{\rm it}\simeq 10^6$ iterations.  
-By contrast, the marginalized likelihood described above  converges incredibly rapidly if the candidate model is
-inconsistent with a signal (or the absence thereof) in the data, with only $N_{it} \simeq O(10^4)$  evaluations needed to identify the marginalized
-likleihood.
+black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$ --  this version has
+cost elements as shown in Table \ref{tab:CostBreakdown}, based on the assumption that the MC terminates after $N_{\rm it}\simeq 2\times 10^6$ iterations using
+$n_{\rm chunk}\simeq 10^4$.  
+The marginalized likelihood described above  converges incredibly rapidly to a small value if the candidate model is
+inconsistent with a signal (or the absence thereof) in the data, with only $N_{it} \simeq O(10^4)$  evaluations needed.
 
-With the new low-cost likelihood evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now are
-comparable or in excess of the Monte Carlo evaluation time.   For this reason, we reorganize the workflow, so each
+\begin{table}
+\begin{tabular}{lrr|cccc|r}
+Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it}$ & $\frac{T_{ILE}}{N_{eval}}$\\  %\hline 
+  &  & Hz & sec & sec & & $10^{-6} $sec & sec\\ \hline 
+CPU & 16384 & $(2,\pm 2) $ & 20 & & & &  \editremark{900}  \\ 
+       & 4096 & $(2,\pm 2) $ &   20 \\ \hline
+% LIGO-WA: 
+%  - old compute node: one marginalized likelihood in 2 minutes (!), 2x8
+GPU (a) & 16384 & $(2,\pm 2) $  & 20  & &   \\
+            & 4096 &$(2,\pm 2) $  &  20 &\\
+% LIGO-CIT, pcdev13
+%  - 20 iterations, about 450  to 500 sec to eval (i.e., about 25 sec per lnL)
+%  -  precompute: can time manually
+%   -  Overlap stages: PrecomputeLIkelihood + waveform, about 4
+%  - Histogram time: about (12+30) seconds overall (histogram + ODEint)
+%  - Actual likelihood evaluation: about 30 sec out of 20 eval, or about 1.5 sec
+   %%   1    0.061    0.061  506.265  506.265 integrate_likelihood_extrinsic_batchmode:8(<module>)
+   %%     20    0.194    0.010  498.581   24.929 integrate_likelihood_extrinsic_batchmode:690(analyze_event)
+   %%     20   80.397    4.020  464.263   23.213 mcsampler.py:327(integrate)
+   %%   2140   54.358    0.025  235.480    0.110 mcsampler.py:198(draw)
+   %%  23530    0.134    0.000  101.280    0.004 shape_base.py:286(hstack)
+   %%  24306  100.861    0.004  100.861    0.004 {numpy.concatenate}
+   %%  20136    0.896    0.000   84.698    0.004 function_base.py:2063(__call__)
+   %%  20136   33.608    0.002   83.802    0.004 function_base.py:2154(_vectorize_call)
+   %%   2140    1.403    0.001   45.791    0.021 integrate_likelihood_extrinsic_batchmode:810(likelihood_function)
+   %% 461726    0.790    0.000   36.137    0.000 polyint.py:62(__call__)
+   %%   2140    5.496    0.003   30.829    0.014 factored_likelihood.py:1676(DiscreteFactoredLogLikelihoodViaArrayVectorNoLoop)
+   %%    407    0.006    0.000   29.662    0.073 odepack.py:28(odeint)
+   %%    407    1.360    0.003   29.644    0.073 {scipy.integrate._odepack.odeint}
+   %%     20    0.052    0.003   28.785    1.439 factored_likelihood.py:92(PrecomputeLikelihoodTerms)
+   %% 461726    2.809    0.000   26.983    0.000 interpolate.py:657(_evaluate)
+   %%     20    0.048    0.002   24.454    1.223 lalsimutils.py:2582(hlmoft_SEOB_dict)
+   %%     20    0.030    0.002   24.406    1.220 lalsimutils.py:2743(complex_hoft)
+   %%     20   24.373    1.219   24.373    1.219 {lalsimulation._lalsimulation.SimInspiralChooseTDWaveform}
+GPU (b) & 16384 & & 20 \\
+       & 4096 & & 20 & $\simeq 1$ &   & 0.75  & 25\\ \hline
+GPU (b) & 16384 & $m=\pm 2, \pm 1$ &  \\
+       & 4096 & $m=\pm 2, \pm 1$ \\
+\end{tabular}
+% Compute node at LIGO-WA
+\caption{\label{tab:CostBreakdown}Evaluation costs on default hardware, for a two-mode system $(l,m)=(2,\pm 2)$.  Scenario (a) is a compute node on LIGO-WA; scenario (b) is more modern hardware. }
+\end{table}
+
+With the new low-cost likelihood evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now can be
+comparable or in excess of the total time used to evaluate the likelihood.   For this reason, we reorganize the workflow, so each
 instance of ILE loops over
@@
 \unit{Hz}$ for all calculations, but terminated all inner products in Eq. (\ref{eq:overlap}) at roughly $f_{\rm
-  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ decreases the  cost $\tau_{setup}$ by the
-same factor. 
+  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ decreases the  cost of all operations with
+timeseries -- they are shorter.  Depending on the relative cost of overhead versus array operations,  $\tau_{setup}$
+\emph{and} the cost per evaluation $\tau_{eval}$ decrease modestly because of this effect.  
 Also, to insure safe inner products over short data sets in the presence of very narrow spectral lines, we operated on a
@@
 \begin{align}
-T_{ILE}/N_{\rm eval} &= \tau_{start}/N_{\rm eval} + [ \tau_{setup}/ss' + N_{it} \tau_{it}] 
+T_{ILE}/N_{\rm eval} &= \frac{\tau_{start}}{N_{\rm eval}} 
+ + 
+ [ \tau_{setup}/s' + N_{ad} \tau_{ad} + N_{it}\tau_{it}
+ ] 
 \end{align}
 For a typical binary black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$, this
-version requires $\tau_{start} \simeq 12\unit{s}$, $\tau_{setup}/ss' \simeq XX$, and $N_{it}\tau_{\it} \simeq XX$
+version requires $\tau_{start} \simeq 12\unit{s}$, $\tau_{setup}/ss' \simeq XX$, $\tau_{ad}\simeq XXX$, and $N_{it}\tau_{\it} \simeq XX$
 \editremark{numbers}.  
-Typically we target $N_{\rm eval} \simeq 20$ so that startup costs are not a significant contributor to the overall
-runtime.   In this configuration, the run time per marginalized likelihood is around $T_{ILE}/N_{\rm eval} \simeq 20
-\unit{s}$ for a binary black hole \editremark{checkme - old number}.
+Because of the extremely high cost of adaptive overhead, we only adapt in two parameters, corresponding to sky location; 
+and only adapt during the first iteration which returns significant $\ln {\cal L}_{\rm marg}$.  As a result, $N_{ad}$
+(the number of adaptive stages \emph{per likelihood evaluation}) scales as $N_{it}/N_{chunk}/N_{\rm eval}$ and does not
+increase with the overall number of samples, so the overall evaluation time scales as
+\begin{align}
+T_{ILE,mod}/N_{\rm eval} &= \frac{\tau_{start} + \tau_{ad}N_{it}/n_{\rm chunk}}{N_{\rm eval}} 
+ \nonumber \\ &
+ + 
+ [ \tau_{setup}/s'  + N_{it}\tau_{it}
+ ] 
+\end{align}
+Typically we target $N_{\rm eval} \simeq 20$ so that startup and other one-time costs  are not a significant contributor to the overall
+runtime.   In this configuration, the run time per marginalized likelihood is around $T_{ILE}/N_{\rm eval} \simeq 25
+\unit{s}$ for a binary black hole \editremark{checkme - old number}.    Of this $25 \unit{s}$, only roughly
+$1.5\unit{s}$ corresponds to evaluating the likleihood, implying the cost $\tau_{it} \simeq 1.5/N_{\rm it} \simeq
+7.5\times 10^{-7}$..
 
 For low-latency analyses, we need to assess the overall wallclock time needed to perform an analysis, often with a
-restricted number $N_{GPU}$ of available GPU-enabled machines.   To  completely $N_{\rm net}$ likelihood evaluations
+restricted number $N_{GPU}$ of available GPU-enabled machines.   To  complete $N_{\rm net}$ likelihood evaluations
 with these resources requires
 \begin{align}
-T_{net} &= \frac{N_{\rm net}}{N_{\rm GPU} N_{\rm eval}} T_{ILE} \nonumber \\
-& \simeq
-  \frac{N_{net}}{N_{GPU}}\left[\tau_{start}/N_{\rm    eval} + [ \tau_{setup}/ss' + N_{it} \tau_{it}]  \right]
+T_{net,mod} &= \frac{N_{\rm net}}{N_{\rm GPU} N_{\rm eval}} T_{ILE,mod} \nonumber \\
+& =
+  \frac{N_{net}}{N_{GPU}}\left[
+\frac{\tau_{start} + \tau_{ad}N_{it}/n_{\rm chunk}}{N_{\rm eval}} 
+ \right. \nonumber \\
+& \left. + 
+ [ \tau_{setup}/ss'  + N_{it}\tau_{it}/s
+ ] 
+ \right]
 \end{align}
@@
 likelihood evaluations will complete in about $100 \unit{minutes}$ \editremark{checkme}. This number can be reduced to
-tens of minutes with a modestly larger GPU pool.
+tens of minutes with a modestly larger GPU pool, or by marginally more conservative convergence thresholds..
 This discussion ignores the latency introduced by the  gaussian process interpolation stage at the end.  We will revisit
@@
 time needed to evaluate  $\ln {\cal L}_{\rm marg}$ for extremely costly
-waveform models which require $\tau_{wf} $ a significant fraction of $  \tau_{setup}$.   \editremark{give examples}
+waveform models which require $\tau_{wf} $ a significant fraction of $  \tau_{setup}$.   That said, the tests described
+above used relatively costly waveform generation with $h_{lm}(t)$ allocation requiring $\tau_{wf} \simeq 1 \unit{s}$.
+
+\editremark{Highly heterogeneous GPU environment - very different performance depending on available resources}
 

### 2019-01-30 paper/paper.tex
paper.tex: minor additional profiling

@@
 \begin{table}
-\begin{tabular}{lrr|cccc|r}
-Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it}$ & $\frac{T_{ILE}}{N_{eval}}$\\  %\hline 
-  &  & Hz & sec & sec & & $10^{-6} $sec & sec\\ \hline 
-CPU & 16384 & $(2,\pm 2) $ & 20 & & & &  \editremark{900}  \\ 
+\begin{tabular}{lrr|ccccc|r}
+Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it,like}$ &$\tau_{it,rest}$ & $\frac{T_{ILE}}{N_{eval}}$\\  %\hline 
+  &  & Hz & sec & sec & & $\mu$sec & $\mu$sec\\ \hline 
+CPU & 16384 & $(2,\pm 2) $ & 20 & && & &  \editremark{900}  \\ 
        & 4096 & $(2,\pm 2) $ &   20 \\ \hline
@@
             & 4096 &$(2,\pm 2) $  &  20 &\\
-% LIGO-CIT, pcdev13
+%% % LIGo CIT, pcdev13, 16 kHz : basically indistinguishable from cost at 4kHz
+%%       1    0.027    0.027  560.572  560.572 integrate_likelihood_extrinsic_batchmode:8(<module>)
+%%        20    0.082    0.004  552.004   27.600 integrate_likelihood_extrinsic_batchmode:690(analyze_event)
+%%        20   76.306    3.815  503.342   25.167 mcsampler.py:327(integrate)
+%%      2345   57.019    0.024  241.951    0.103 mcsampler.py:198(draw)
+%%     25785    0.135    0.000   99.535    0.004 shape_base.py:286(hstack)
+%%     26561   99.102    0.004   99.102    0.004 {numpy.concatenate}
+%%     21981    0.933    0.000   87.819    0.004 function_base.py:2063(__call__)
+%%     21981   34.750    0.002   86.887    0.004 function_base.py:2154(_vectorize_call)
+%%      2345    1.391    0.001   86.327    0.037 integrate_likelihood_extrinsic_batchmode:810(likelihood_function)
+%%      2345    0.007    0.000   46.648    0.020 __init__.py:630(asnumpy)
+%%      2345   46.557    0.020   46.640    0.020 {method 'get' of 'cupy.core.core.ndarray' objects}
+%%        20    0.179    0.009   43.145    2.157 factored_likelihood.py:92(PrecomputeLikelihoodTerms)
+%%        20    0.174    0.009   37.498    1.875 lalsimutils.py:2582(hlmoft_SEOB_dict)
+%%        20    0.099    0.005   37.324    1.866 lalsimutils.py:2743(complex_hoft)
+%%        20   37.220    1.861   37.220    1.861 {lalsimulation._lalsimulation.SimInspiralChooseTDWaveform}
+%%    444365    0.761    0.000   35.303    0.000 polyint.py:62(__call__)
+%%      2345    5.260    0.002   33.956    0.014 factored_likelihood.py:1676(DiscreteFactoredLogLikelihoodViaArrayVectorNoLoop)
+%%       407    0.005    0.000   28.757    0.071 odepack.py:28(odeint)
+%%       407    1.327    0.003   28.740    0.071 {scipy.integrate._odepack.odeint}
+%%    444365    2.644    0.000   26.423    0.000 interpolate.py:657(_evaluate)
+%% 117270874   22.569    0.000   22.569    0.000 mcsampler.py:685(uniform_samp)
+% LIGO-CIT, pcdev13, 4kHz
 %  - 20 iterations, about 450  to 500 sec to eval (i.e., about 25 sec per lnL)
@@
    %%     20   24.373    1.219   24.373    1.219 {lalsimulation._lalsimulation.SimInspiralChooseTDWaveform}
-GPU (b) & 16384 & & 20 \\
-       & 4096 & & 20 & $\simeq 1$ &   & 0.75  & 25\\ \hline
-GPU (b) & 16384 & $m=\pm 2, \pm 1$ &  \\
-       & 4096 & $m=\pm 2, \pm 1$ \\
+GPU (b) & 16384 & & 20  & 1.8 & 1 & 0.85& 12 &28\\
+       & 4096 & & 20 & $1.2 $ &  1  & 0.75 &  & 25\\ \hline
+GPU (b) & 16384 & $\pm 2, \pm 1$ &  \\
+       & 4096 & $\pm 2, \pm 1$ \\
 \end{tabular}
@@
 With the new low-cost likelihood evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now can be
-comparable or in excess of the total time used to evaluate the likelihood.   For this reason, we reorganize the workflow, so each
+comparable or in excess of the total time used to evaluate the likelihood.   In fact, the total time per Monte Carlo
+evaluation $\tau_{it}=\tau_{manage}+\tau_{eval}$  spent in general-purpose overhead $\tau_{manage}$ will be much larger than the time $\tau_{eval}$ spent
+carrying out scientific calculations by evaluationg the likelihood.  For these reasons, we reorganize the workflow, so each
 instance of ILE loops over
@@
 %
-Additionally, particularly for massive binary black holes, we adopt two additional performance improvements.  First and
+Additionally, particularly for massive binary black holes, we investigate two additional performance improvements.  First and
 foremost, we lower the sampling rate, thus lowering the startup cost $\tau_{start}$ and particularly setup cost
@@
 \unit{Hz}$ for all calculations, but terminated all inner products in Eq. (\ref{eq:overlap}) at roughly $f_{\rm
-  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ decreases the  cost of all operations with
+  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ will the  cost of all operations with
 timeseries -- they are shorter.  Depending on the relative cost of overhead versus array operations,  $\tau_{setup}$
@@
 $1.5\unit{s}$ corresponds to evaluating the likleihood, implying the cost $\tau_{it} \simeq 1.5/N_{\rm it} \simeq
-7.5\times 10^{-7}$..
+7.5\times 10^{-7}$.
+Likewise, because overhead dominates over array manipulations, an analysis with  $16\unit{kHz}$ timeseries requires only
+marginally more time than the corresponding $4\unit{kHz}$ analysis.
 
@@
 waveform models which require $\tau_{wf} $ a significant fraction of $  \tau_{setup}$.   That said, the tests described
-above used relatively costly waveform generation with $h_{lm}(t)$ allocation requiring $\tau_{wf} \simeq 1 \unit{s}$.
+above used relatively costly waveform generation with $h_{lm}(t)$ allocation requiring $\tau_{wf} \simeq 1.2 \unit{s}$
+at 4kHz and $1.9\unit{s}$ at 16 kHz.
 

### 2019-01-30 paper/paper.tex
paper.tex work on BNS section

@@
 \affiliation{\RIT}
-\author{Jake - if he does runs}
+\author{Jake?}
 \affiliation{\RIT}
@@
 
-\subsection{Exploiting low-cost likelihoods}
+\subsection{Tradeoff between Monte Carlo integration and accuracy}
+
+Each marginalized likelihood evaluation has a relative uncertainty of order  $1/\sqrt{n_{\rm eff}}$, where the number of
+effective samples $n_{\rm eff}=\epsilon N_{\rm it}$ increases linearly with the total number $N_{\rm it}$ of Monte Carlo
+evaluations performed by \ILE{}.  Therefore, to increase (decrease) our accuracy for each likelihood evaluation by a factor of $A$
+will require a factor $A^2$ more (fewer) iterations.   
+%
+We adopt a fixed threshold on $n_{\rm eff}$.
+
+We do benefit slightly by re-using the adaptive Monte Carlo intergrator for each extrinsic point $\lambda$.  Since the
+integrator has already identified the likely range of sky locations and distances on the first iteration, each
+subsequent evaluation of the marginalized likelihood can converge marginally more quickly.
+
+\section{Validation and Performance}
+\label{sec:demo}
+
+We illustrate the code using two synthetic signals: a binary black hole with masses $m_{1,2}=35,30 M_\odot$ and a binary
+neutron star with masses $1.4 M_\odot$ and $1.35 M_\odot$.
+We generate  synthetic data for a two-detector LIGO configuration, assuming operation at  initial (BBH) or advanced (BNS) design  sensitivity.
+Both signals are chosen to have peak SNR $\simeq 14$ and are generated with SEOBNRv4.    We perform a multi-stage iterative RIFT analysis of these signal with the
+time-domain SEOBNRv4 (BBH) or TaylorT4 (BNS) approximation, under the simplifying assumption that both objects are point particles with zero
+spin.  We repeat this analysis on heterogeneous LVC collaboration computing resources, to assess our variable
+performance across architectures.  Notably, we investigate the following GPU options: (a) GTX 1050 Ti at LIGO-CIT,
+available in large numbers; (b) V100, available on selected high-performance machines; (c) and
+GTX 1050 at LIGO-LHO.    Unless otherwise noted in the text, we will discuss code configurations using CPU-only and GPU (b) using
+$4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.
+
+\subsection{Massive compact binary black holes}
 Previously, each instance of ILE examined one intrinsic point $\lambda$.  The overall cost of this ILE evaluation
@@
 
-\begin{table}
-\begin{tabular}{lrr|ccccc|r}
-Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it,like}$ &$\tau_{it,rest}$ & $\frac{T_{ILE}}{N_{eval}}$\\  %\hline 
-  &  & Hz & sec & sec & & $\mu$sec & $\mu$sec\\ \hline 
-CPU & 16384 & $(2,\pm 2) $ & 20 & && & &  \editremark{900}  \\ 
-       & 4096 & $(2,\pm 2) $ &   20 \\ \hline
-% LIGO-WA: 
+\begin{table*}
+\begin{tabular}{lrr|ccccc|rr}
+Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it,like}$ &$\tau_{it,rest}$ &
+$\frac{T_{ILE}}{N_{eval}}$ & GPU \\  %\hline 
+  &   Hz & m & sec & sec & & $\mu$sec & $\mu$sec  &sec  & use  \%\\ \hline 
+% ~/parse_report.sh profile_nogpu_pcdev13.log | more
+CPU & 16384 & $\pm 2 $ & 20 & &&532 & &  700  \\ 
+       & 4096 & $\pm 2 $ &   20 \\ \hline
+% ./parse_report.sh 20190130-profile_nogpu_HM_pcdev13.log  | more
+%    setup time: PrecomputeLikelihoodTerms, includes waveform generation. 
+%   evaluation: FactoredLogLikelihodTimeMarginalized Divide by actual number of calls, since not a block!
+%    
+CPU & 16384 & $\pm 2,\pm 1 $ & 20 & 1.5 && 680 & &  1000  \\ 
+       & 4096 & $ \pm 2, \pm 1 $ &   20 \\ \hline
+% LIGO-WA, or a compute node at CIT with a grpahics catd: 
 %  - old compute node: one marginalized likelihood in 2 minutes (!), 2x8
-GPU (a) & 16384 & $(2,\pm 2) $  & 20  & &   \\
-            & 4096 &$(2,\pm 2) $  &  20 &\\
+%  - CIT jobs, on compute nodes
+%        20 points at 16kHz take about 90 min
+%        20 points at kHz take anywhere from
+%        condor_userlog iteration_0_ile/logs/*.log
+GPU (a) & 16384 & $(2,\pm 2) $  & 20  & & && & 270 \\
+            & 4096 &$\pm 2 $  &  20 &  & & && 45 \\ \hline
 %% % LIGo CIT, pcdev13, 16 kHz : basically indistinguishable from cost at 4kHz
@@
    %%     20   24.373    1.219   24.373    1.219 {lalsimulation._lalsimulation.SimInspiralChooseTDWaveform}
-GPU (b) & 16384 & & 20  & 1.8 & 1 & 0.85& 12 &28\\
-       & 4096 & & 20 & $1.2 $ &  1  & 0.75 &  & 25\\ \hline
-GPU (b) & 16384 & $\pm 2, \pm 1$ &  \\
-       & 4096 & $\pm 2, \pm 1$ \\
+GPU (b) & 16384 & $\pm 2$ & 20  & 1.8 & 1 & 0.85& 12 &28 & 15\\
+       & 4096 & $\pm 2$  & 20 & $1.2 $ &  1  & 0.75 &  & 25\\ \hline
+% ~/parse_report.sh profile_gpu_HM_lowres_pcdev13.log | more
+GPU (b) & 16384 & $\pm 2, \pm 1$ & 20  \\
+       & 4096 & $\pm 2, \pm 1$ & 20 & 1&& 2.5  && 35 & \\ \hline
+%
+% LHO or LLO worker nodes
+%  ./parse_report.sh 20190130-profile_gpu_LHOworker_16kHz.log | more
+GPU (c) & 16384 & $\pm 2 $  & 20  & & & 18& &160 &  \\
+            & 4096 &$\pm 2 $  &  20 &  & & 22  & & 140 & $\simeq 50$ \\
 \end{tabular}
 % Compute node at LIGO-WA
-\caption{\label{tab:CostBreakdown}Evaluation costs on default hardware, for a two-mode system $(l,m)=(2,\pm 2)$.  Scenario (a) is a compute node on LIGO-WA; scenario (b) is more modern hardware. }
-\end{table}
+\caption{\label{tab:CostBreakdown}\textbf{Profiling performance: Binary black holes}: Evaluation costs for the
+  marginalized likelihood on default
+  hardware, for a two-mode system $(l,m)=\pm 2$ analysing $T=8\unit{s}$ of data with a massive binary black hole
+  $m_1=35 M_\odot,M_2=30 M_\odot$.  The last column indicates peak GPU utilization.}
+\end{table*}
 
@@
 runtime.   In this configuration, the run time per marginalized likelihood is around $T_{ILE}/N_{\rm eval} \simeq 25
-\unit{s}$ for a binary black hole \editremark{checkme - old number}.    Of this $25 \unit{s}$, only roughly
+\unit{s}$ for a binary black hole.    Of this $25 \unit{s}$, only roughly
 $1.5\unit{s}$ corresponds to evaluating the likleihood, implying the cost $\tau_{it} \simeq 1.5/N_{\rm it} \simeq
@@
 individual marginalized likelihood evaluation.  For the binary black hole configuration described above, marginalized
-likelihood evaluations will complete in about $100 \unit{minutes}$ \editremark{checkme}. This number can be reduced to
-tens of minutes with a modestly larger GPU pool, or by marginally more conservative convergence thresholds..
+likelihood evaluations will complete in about $100\; \unit{minutes}$ on a cluster of (b), or more realistically 200
+minutes on a cluster of (a). . This number can be reduced to
+tens of minutes with a modestly larger GPU pool, or by  marginally more conservative convergence thresholds. on $N_{\rm
+  eval}$ or $N_{it}$.  
 This discussion ignores the latency introduced by the  gaussian process interpolation stage at the end.  We will revisit
@@
 
-\editremark{Highly heterogeneous GPU environment - very different performance depending on available resources}
+The \ILE{} likelhood involves sums over modes, with the most expensive likelihood and setup operations (involving $Q$) growing linearly with
+the number of modes used.  This increased cost is most apparent in Table \ref{tab:CostBreakdown} when comparing the
+columns corresponding  $\tau_{it,like}$ for GPU (b) between the $\pm 2$ and $(\pm 2,\pm 1)$ rows.  The likelihood
+evaluation cost of runs with 4 modes is roughly twice that of runs with two modes, as expected.
 
-\subsection{Tradeoff between Monte Carlo integration and accuracy}
+In a heterogeneous computing environment, code performance varies substantially with available resources, as seen by
+contrasting corresponding results for (a,b,c) in Table \ref{tab:CostBreakdown}.      For
+the slowest hardware,  the overall runtime will be only a factor few smaller than the corresponding CPU-only runtime.
 
-Each marginalized likelihood evaluation has a relative uncertainty of order  $1/\sqrt{n_{\rm eff}}$, where the number of
-effective samples $n_{\rm eff}=\epsilon N_{\rm it}$ increases linearly with the total number $N_{\rm it}$ of Monte Carlo
-evaluations performed by \ILE{}.  Therefore, to increase (decrease) our accuracy for each likelihood evaluation by a factor of $A$
-will require a factor $A^2$ more (fewer) iterations.   
-%
-We adopt a fixed threshold on $n_{\rm eff}$.
+\subsection{Binary neutron stars}
+\begin{table*}
+\begin{tabular}{lrr|ccccc|rr}
+Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it,like}$ &$\tau_{it,rest}$ &
+$\frac{T_{ILE}}{N_{eval}}$ & GPU \\  %\hline 
+  &   Hz & m & sec & sec & & $\mu$sec & $\mu$sec  &sec  & use  \%\\ \hline 
+CPU & 16384 & $\pm 2,\pm 1 $ & 20 &  &&  & &    \\ 
+       & 4096 & $\pm 2, \pm 1$ & 20 & 1&&   &&  & \\ \hline
+GPU (b) & 16384 & $\pm 2, \pm 1$   & 20 & 26 & 2.4 && &35 &  \\
+       & 4096 & $ \pm 2, \pm 1 $        &   20  & 11.5 & 2.4 & & &24 & \\ \hline
+\end{tabular}
+\caption{\label{tab:CostBreakdown:BNS}\textbf{Profiling performance: Binary neutron stars}: Evaluation costs for the
+  marginalized likelihood on default
+  hardware,  analysing $T=8\unit{s}$ of data with a binary neutron stars
+  $m_1=1.4 M_\odot,M_2=1.35 M_\odot$, with convergence threshold $n_{\rm eff} > 50$.
+}
+\end{table*}
 
-We do benefit slightly by re-using the adaptive Monte Carlo intergrator for each extrinsic point $\lambda$.  Since the
-integrator has already identified the likely range of sky locations and distances on the first iteration, each
-subsequent evaluation of the marginalized likelihood can converge marginally more quickly.
+Table \ref{tab:CostBreakdown: BNS} shows the corresponding performance breakdown for our fiducial binary neutron star,
+which has signal SNR=32.  In this analysis, we have adopted the TaylorT4 model with $\ell=2$ and all $m=\pm 2,\pm 1$ modes as our
+fiducial analysis template.   Despite using more modes and longer signals, and despite an overall likelihood evaluation
+cost $\tau_{it}$ consistent BBH analyses,  likelihood evaluations for BNS are often
+\emph{as fast or faster} than the corresponding likelihood evaluations for BBH with otherwise comparable settings.
 
-\section{Validation and Performance}
-\label{sec:demo}
+Low-mass compact binaries like binary neutron stars have much longer inspiral times from the same starting frequency of
+$20\unit{Hz}$: roughly $160 \unit{s}$.   Nominally, the manipulation of a corresponding power-of-two duration data ($256\unit{s}$) involves
+$32\times$ more costly fourier transforms and time integrations than the BBH analysis described above.  
+%  a = np.zeros(16384*256,dtype=np.complex64); a.nbytes/1024/1024
+[An array of $256\unit{s}$ of data at $16\unit{kHz}$ corresponds to  32 Mb.]  Comparing $\tau_{setup}$ appearing in Tables
+ \ref{tab:CostBreakdown} and  \ref{tab:CostBreakdown:BNS}, we indeed find our BNS analysis requires  a factor of order   $32\times$
+ longer to set up all necessary inner products.  Unlike the BBH analysis, the inner product evaluation costs in
+ $\tau_{setup}$ now dominates the
+ overall evaluation time $T_{ILE}/N_{\rm eval}$.
 
-\begin{figure}
+Improved performance arises instead from re-using our adaptive integrator, which (after the first marginalized
+likelihood evaluation) can exploit tight localization afforded by the many BNS cycles available in band and the high
+amplitude of our fiducial BNS signal.  In other
+words, the Monte Carlo integral converges to our target accuracy in of order $10\times $ fewer steps than for BBH.  
+Because this improved performance relies on the sampler, not on our choice of hardware, we achieve a similar substantial
+improvement for all choices of GPU.
 
-\end{figure}
-
-* Comparison: agreement.  Sky localization example
-
-* Benchmarking performance (new speed for BBHs; also show BNS with tides)
+\editremark{test on other hardware}
 

### 2019-01-31 paper/paper.tex
paper.tex: update tables. set up conclusions

@@
    %%     20   24.373    1.219   24.373    1.219 {lalsimulation._lalsimulation.SimInspiralChooseTDWaveform}
-GPU (b) & 16384 & $\pm 2$ & 20  & 1.8 & 1 & 0.85& 12 &28 & 15\\
-       & 4096 & $\pm 2$  & 20 & $1.2 $ &  1  & 0.75 &  & 25\\ \hline
+% ./parse_report.sh 20190130-profile_gpu_lowres_pcdev13.log_welford | more
+% ./parse_report.sh 20190130-profile_gpu_lowres_pcdev13.log_16kHz | more
+%   'residual' time: take mcsampler, subtract likleihood, divide by number of calls and number of evals per chunk
+GPU (b) & 16384 & $\pm 2$ & 20  & 1.8 & 1 & \editremark{0.85}& \editremark{6} &28 & 15\\
+       & 4096 & $\pm 2$  & 20 & $1.2 $ &  1  & \editremark{0.75} & \editremark{6}  & 25\\ \hline
 % ~/parse_report.sh profile_gpu_HM_lowres_pcdev13.log | more
-GPU (b) & 16384 & $\pm 2, \pm 1$ & 20  \\
+%   - extra eval cost: (mcsampler - likelihood)/20/n_eval_likelihood/10^4  
+GPU (b) & 16384 & $\pm 2, \pm 1$ & 20 & 1 && 4.2 & \editremark{1}  & 38  \\
        & 4096 & $\pm 2, \pm 1$ & 20 & 1&& 2.5  && 35 & \\ \hline
@@
 
-With the new low-cost likelihood evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now can be
-comparable or in excess of the total time used to evaluate the likelihood.   In fact, the total time per Monte Carlo
+With the new low-cost ${\cal L}_{\rm margT}$ evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now can be
+comparable or in excess of the total time used to evaluate the marginalized likelihood ${\cal L}_{marg}$.   In fact, the total time per Monte Carlo
 evaluation $\tau_{it}=\tau_{manage}+\tau_{eval}$  spent in general-purpose overhead $\tau_{manage}$ will be much larger than the time $\tau_{eval}$ spent
@@
 %
-Combining these factors together, the overall computational cost of each marginalized likelihood
-evaluation $T_{ILE}/N_{eval}$  is
+Combining these factors together, the overall computational cost  $T_{ILE}/N_{eval}$  of each marginalized likelihood
+evaluation is
 \begin{align}
@@
 
-The \ILE{} likelhood involves sums over modes, with the most expensive likelihood and setup operations (involving $Q$) growing linearly with
+The \ILE{} likelhood $\ln {\cal L}_{\rm margT}$ involves sums over modes, with the most expensive likelihood and setup operations (involving $Q$) growing linearly with
 the number of modes used.  This increased cost is most apparent in Table \ref{tab:CostBreakdown} when comparing the
-columns corresponding  $\tau_{it,like}$ for GPU (b) between the $\pm 2$ and $(\pm 2,\pm 1)$ rows.  The likelihood
-evaluation cost of runs with 4 modes is roughly twice that of runs with two modes, as expected.
+columns corresponding  $\tau_{it,like}$ for GPU (b) between the $\pm 2$ and $(\pm 2,\pm 1)$ rows.  The cost of
+evaluating $\ln {\cal L}_{\rm margT}$   with 4 modes is roughly twice that of runs with two modes, as expected.
 
@@
   &   Hz & m & sec & sec & & $\mu$sec & $\mu$sec  &sec  & use  \%\\ \hline 
-CPU & 16384 & $\pm 2,\pm 1 $ & 20 &  &&  & &    \\ 
-       & 4096 & $\pm 2, \pm 1$ & 20 & 1&&   &&  & \\ \hline
-GPU (b) & 16384 & $\pm 2, \pm 1$   & 20 & 26 & 2.4 && &35 &  \\
-       & 4096 & $ \pm 2, \pm 1 $        &   20  & 11.5 & 2.4 & & &24 & \\ \hline
+% ./parse_report.sh 20190130-profile_bns_nogpu_16kHz_pcdev13.log | more
+%   - resampling time: take (mcsampler - likelihood), divide by 20=Neval, divide by number of adaptive steps (100)
+%   (11261 - 10974)/20/100. = 0.14 sec for 16 kHz
+CPU & 16384 & $\pm 2,\pm 1 $ & 35 & 26 & 0.14  & 680  & & 590    \\ 
+       & 4096 & $\pm 2, \pm 1$ & 35 & &  &   &&  & \\ \hline
+%   - resampling time: take (mcsampler - likelihood), divide by 20=Neval, divide by number of adaptive steps (100)
+%   (226.489 - 17)/20/100  : 0.1 sec for 4kHz
+%   (165 - 29.  )/20/100   : looks odd for 16 kHz
+GPU (b) & 16384 & $\pm 2, \pm 1$   & 35 & 26 & 0.07 &  2.4 & &35 &  \\
+       & 4096 & $ \pm 2, \pm 1 $        &  35  & 11.5 & 0.1 &  2.4  & &24 & \\ \hline
 \end{tabular}
@@
 
-Table \ref{tab:CostBreakdown: BNS} shows the corresponding performance breakdown for our fiducial binary neutron star,
+Table \ref{tab:CostBreakdown:BNS} shows the corresponding performance breakdown for our fiducial binary neutron star,
 which has signal SNR=32.  In this analysis, we have adopted the TaylorT4 model with $\ell=2$ and all $m=\pm 2,\pm 1$ modes as our
@@
 
+For the closed-form post-Newtonian approximation used in the study above, waveform costs are a modest contribution to
+overall runtime, contributing only 4.1 sec to $\tau_{setup}$ at 16 kHz.
+For other approximations not available in closed form, the waveform generation cost for binary neutron star inspirals
+can be much more substantial, particularly as $f_{\rm min}$ decreases below the 20\unit{Hz} used in this study.
+
 \section{ Results (optional?)}
@@
 \section{Conclusions}
+We have demonstrated that the marginalized likelihood $\ln {\cal L}_{\rm marg}(\lambda)$ appearing in the RIFT/rapidPE parameter
+inference calculation can be evaluated at fixed $\lambda$ in tens of seconds on average for
+both binary black holes and binary neutron stars.  This performance improvement could enable very low latency source
+parameter inference for compact binaries, which can be of use for targeting multimessenger followup observations via sky
+localization and  precise source characterization.     This prospect is particularly interesting because RIFT can
+often achieve this performance using computationally costly models for binary merger, as the waveform generation cost
+does not usually limit code performance.
 
-* Enable low-latency parameter inference in basically real time for GW candidate sources, even for relatively costly
-models with full physics
+The overall code performance and thus latency can be further decreased substantially, notably by converting the Monte
+Carlo random number generation and inner products to GPU-based operations.  In such a configuration, the marginalized
+likeliood would carry out almost all calculations (except waveform generation) on a GPU, with minimal communication off
+the board.  This configuration should further reduce the average time needed to compute $\ln {\cal L}_{\rm
+  marg}(\lambda)$ for both binary black holes (which are Monte Carlo limited) and binary neutron stars (which are
+inner-product limited).   We anticipate a further factor of roughly 10 reduction in overall evaluation time can be
+achieved soon.  At that level of performance, a single 8-GPU machine with contemporary hardware could perform parameter inference less than 10
+minutes.    Since binary compact objects intersect our past light cone only once every roughly 15 minutes, accounting
+for all past history, such a configuration would be able to address low-latency parameter inference for the duration of
+2nd-generation ground-based observing.
+% based on Neval ~ 10^4
 
-* Highly portable, generalizable to other codes: code will smoothly tranlate between GPU-only and CPU-only mode
-
-* Not done yet: expect to transform MC and inner products, so all calculations remain on board, with just intermittent
-control I/O from CPU
+The use of \texttt{cupy} enables our code to be highly portable across architectures and heterogeneous GPU environments,
+while transitioning smoothly between GPU and CPU mode.     The techniques we used here will be transported to other
+Bayesian inference modeling codes used to interpret GW observations.
+% Will swap version of code
 

### 2019-01-31 paper/paper.tex
paper.tex: spellcheck. References.

@@
 The properties of the sources responsible have been inferred via Bayesian comparison of  each source with candidate
-gravitational ave signals
+gravitational wave signals
 \cite{DiscoveryPaper,LIGO-O1-BBH,2017PhRvL.118v1101A,LIGO-GW170814,LIGO-GW170608,LIGO-GW170817-bns}
@@
 %overall computational cost and turnaround time of 
-computational cost of parameter estimation \cite{gwastro-pe-Brandon-STF2,gwastro-PE-AlternativeArchitectures,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,2016PhRvD..94d4031S,2017arXiv170302062V}.
+computational cost of parameter estimation \cite{gwastro-pe-Brandon-STF2,gwastro-PE-AlternativeArchitectures,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,2016PhRvD..94d4031S,2017CQGra..34k5006V}.
 % The computational cost of parameter estimation can be reduced 
@@
 The RIFT strategy described in \editremark{citations - include ROM} eschews these simplifications, performing
-embarassingly-parallel inferences even for costly models.   However, the method as previously implemented still had a
-significant net  computational cost and noticable startup time, limiting its diverse potential applications.
+embarrassingly-parallel inferences even for costly models.   However, the method as previously implemented still had a
+significant net  computational cost and noticeable startup time, limiting its diverse potential applications.
 
@@
 \end{align}
-where $F,D$ are arrays over extrinsic parameters; $Q$ is an array over time and $(l,m)$; and $U,V$ ar matricies over
+where $F,D$ are arrays over extrinsic parameters; $Q$ is an array over time and $(l,m)$; and $U,V$ are matrices over
 $(l,m)$.
@@
 
-We do benefit slightly by re-using the adaptive Monte Carlo intergrator for each extrinsic point $\lambda$.  Since the
+We do benefit slightly by re-using the adaptive Monte Carlo integrator for each extrinsic point $\lambda$.  Since the
 integrator has already identified the likely range of sky locations and distances on the first iteration, each
@@
 We generate  synthetic data for a two-detector LIGO configuration, assuming operation at  initial (BBH) or advanced (BNS) design  sensitivity.
-Both signals are chosen to have peak SNR $\simeq 14$ and are generated with SEOBNRv4.    We perform a multi-stage iterative RIFT analysis of these signal with the
+Both signals are generated with SEOBNRv4.    We perform a multi-stage iterative RIFT analysis of these signal with the
 time-domain SEOBNRv4 (BBH) or TaylorT4 (BNS) approximation, under the simplifying assumption that both objects are point particles with zero
@@
 GTX 1050 at LIGO-LHO.    Unless otherwise noted in the text, we will discuss code configurations using CPU-only and GPU (b) using
-$4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.
+$4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.  We also conservatively report computational cost for
+parameters $\lambda$ which are close to or within the posterior and therefore, due to their consistency with the data,
+have the best-determined extrinsic parameters and thus converge the most slowly.
+
+\editremark{report on distance to each signal, and in particular $d_{\rm max}$ as that might impact performance too}
+chosen to have peak SNR $\simeq 14$ and
 
@@
  $N_{ad} \simeq N_{it}/n_{chunk}$), regenerating a sampling prior in two sky location parameters and distance.  
-The choice of one $\lambda$ for each instance of ILE was eue to the substantial time  $\tau_{MC}$, which vastly
+The choice of one $\lambda$ for each instance of ILE was due to the substantial time  $\tau_{MC}$, which vastly
 dominated the overall computational cost.   For example, for a typical analysis of a short signal --  a typical binary
@@
   marginalized likelihood on default
-  hardware, for a two-mode system $(l,m)=\pm 2$ analysing $T=8\unit{s}$ of data with a massive binary black hole
+  hardware, for a two-mode system $(l,m)=\pm 2$ analyzing $T=8\unit{s}$ of data with a massive binary black hole
   $m_1=35 M_\odot,M_2=30 M_\odot$.  The last column indicates peak GPU utilization.}
@@
 evaluation $\tau_{it}=\tau_{manage}+\tau_{eval}$  spent in general-purpose overhead $\tau_{manage}$ will be much larger than the time $\tau_{eval}$ spent
-carrying out scientific calculations by evaluationg the likelihood.  For these reasons, we reorganize the workflow, so each
+carrying out scientific calculations by evaluating the likelihood.  For these reasons, we reorganize the workflow, so each
 instance of ILE loops over
@@
 \unit{s}$ for a binary black hole.    Of this $25 \unit{s}$, only roughly
-$1.5\unit{s}$ corresponds to evaluating the likleihood, implying the cost $\tau_{it} \simeq 1.5/N_{\rm it} \simeq
+$1.5\unit{s}$ corresponds to evaluating the likelihood, implying the cost $\tau_{it} \simeq 1.5/N_{\rm it} \simeq
 7.5\times 10^{-7}$.
@@
 
-The \ILE{} likelhood  generates waveform dynamics and $h_{lm}(t)$ once per evaluation $\lambda$.   Waveform generation can contribute significantly to the
+The \ILE{} likelihood  generates waveform dynamics and $h_{lm}(t)$ once per evaluation $\lambda$.   Waveform generation can contribute significantly to the
 time needed to evaluate  $\ln {\cal L}_{\rm marg}$ for extremely costly
@@
 
-The \ILE{} likelhood $\ln {\cal L}_{\rm margT}$ involves sums over modes, with the most expensive likelihood and setup operations (involving $Q$) growing linearly with
+The \ILE{} likelihood $\ln {\cal L}_{\rm margT}$ involves sums over modes, with the most expensive likelihood and setup operations (involving $Q$) growing linearly with
 the number of modes used.  This increased cost is most apparent in Table \ref{tab:CostBreakdown} when comparing the
@@
        & 4096 & $\pm 2, \pm 1$ & 35 & &  &   &&  & \\ \hline
+GPU (a) & 16384 & $\pm 2, \pm 1$   & 35 &  &&&&   \\
+       & 4096 & $ \pm 2, \pm 1 $        &  35  & &&&& 360 \\ \hline
 %   - resampling time: take (mcsampler - likelihood), divide by 20=Neval, divide by number of adaptive steps (100)
@@
   marginalized likelihood on default
-  hardware,  analysing $T=8\unit{s}$ of data with a binary neutron stars
+  hardware,  analyzing $T=8\unit{s}$ of data with a binary neutron stars
   $m_1=1.4 M_\odot,M_2=1.35 M_\odot$, with convergence threshold $n_{\rm eff} > 50$.
@@
 $20\unit{Hz}$: roughly $160 \unit{s}$.   Nominally, the manipulation of a corresponding power-of-two duration data ($256\unit{s}$) involves
-$32\times$ more costly fourier transforms and time integrations than the BBH analysis described above.  
+$32\times$ more costly Fourier transforms and time integrations than the BBH analysis described above.  
 %  a = np.zeros(16384*256,dtype=np.complex64); a.nbytes/1024/1024
@@
 
-\editremark{test on other hardware}
+\editremark{test on other hardware -- also problem, a lot of the speed improvement occurs because the initial grid is
+  crap/too wide}
 
@@
 Carlo random number generation and inner products to GPU-based operations.  In such a configuration, the marginalized
-likeliood would carry out almost all calculations (except waveform generation) on a GPU, with minimal communication off
+likelihood would carry out almost all calculations (except waveform generation) on a GPU, with minimal communication off
 the board.  This configuration should further reduce the average time needed to compute $\ln {\cal L}_{\rm
@@
 %\bibstyle{unsrt}
-\bibliography{paperexport}
+\bibliography{paperexport,LIGO-publications}
 \end{document}

### 2019-02-01 paper/paper.tex
paper.tex:edits, references

@@
 gravitational wave signals
-\cite{DiscoveryPaper,LIGO-O1-BBH,2017PhRvL.118v1101A,LIGO-GW170814,LIGO-GW170608,LIGO-GW170817-bns}
-\editremark{citations: real events, also LI and RIFT} \cite{gwastro-PE-AlternativeArchitectures,gwastro-PENR-RIFT}.  
+\cite{DiscoveryPaper,LIGO-O1-BBH,2017PhRvL.118v1101A,LIGO-GW170814,LIGO-GW170608,LIGO-GW170817-bns,LIGO-O2-Catalog,gwastro-PE-AlternativeArchitectures,gwastro-PENR-RIFT,gw-astro-PE-lalinference-v1}.
 Many more are expected as observatories reach design sensitivity \cite{2016LRR....19....1A}.  
 Both to handle the rapid pace of discovery and particularly to enable coordinated multimessenger followup, GW
-observatories should be able to reconstruct the full source parameters of coalescing binaries as fast as possible
-\cite{LIGO-whitepaper-2015} \editremark{update}.    
-However, particularly when using the best-available waveform models, these calculations can be very costly.  In
-practice, detailed inferences even using simplified waveforms take weeks for binary neutron stars, and can take months
-for costly time-domain effective-one-body models which incorporate the effects of precession.
+observatories should be able to reconstruct the evidence for a signal being present in the data along with the full source parameters of coalescing binaries as fast as possible
+\cite{LIGO-2016-WhitePaper,LIGO-Program-2018}.
+However, particularly when using the best-available waveform models, these calculations can be very costly.  For binary
+neutron stars, detailed inferences even using simplified waveforms take weeks.  Even for massive binary  black holes,
+these calculations can take  months for costly time-domain effective-one-body models which incorporate the effects of precession.
 
@@
 %
-The RIFT strategy described in \editremark{citations - include ROM} eschews these simplifications, performing
+The RIFT strategy described in \cite{gwastro-PE-AlternativeArchitectures,2017CQGra..34n4002O,gwastro-PENR-RIFT,NRPaper} eschews these simplifications, performing
 embarrassingly-parallel inferences even for costly models.   However, the method as previously implemented still had a
 significant net  computational cost and noticeable startup time, limiting its diverse potential applications.
-
-In this paper we describe a substantial improvement to the RIFT approach, enabled by translating its low-level
-algorithms to operate on graphics processing units (GPUs).
+%
+In this paper we describe a variant of the RIFT algorithm which  translates its low-level
+algorithms to operate on graphics processing units (GPUs), and adjusts its workflow to exploit the speed improvements
+this re-implementation affords.
 
 This paper is organized as follows.
-In Section \ref{Methods}, we review the underlying marginalized likelihood calculations used by RIFT; their updated GPU
-implementation; and the operating settings which enable substantially increased performance.
-In Section \ref{sec:demo}, we quantify the improved performance of our  GPU-accelerated code, and its agreement with the
-original CPU-only version.
+In Section \ref{sec:methods}, we review the underlying marginalized likelihood calculations used by RIFT and  their updated GPU
+implementation.  
+In Section \ref{sec:demo}, we quantify the improved performance of our  GPU-accelerated code,   while assessing
+operating settings which facilitate increased performance.
+In Section \ref{sec:end-to-end}, we describe the performance of the end-to-end pipeline on the synthetic nonspinning signals used in
+Section \ref{sec:demo}.  
+In Section \ref{sec:conclude}, we summarize our results and discuss their potential applications to future GW source and
+population inference.
 
 \section{Methods}
+\label{sec:methods}
 
@@
 For the remaining dimensions, to evaluate the likelihood in regions of high importance, we use an adaptive Monte Carlo as described in
-\cite{gwastro-PE-AlternativeArchitectures}.    As described in Pankow et al, this marginalized likelihood can be evaluated efficiently
-because, having generated the dynamics and outgoing radiation in all possible directions once and for all for fixed
-$\mathbf{\lambda}$, the likelihood can be evaluated as a function of $\theta$ at very low computational cost.
+\cite{gwastro-PE-AlternativeArchitectures}.
 
+This marginalized likelihood can be evaluated efficiently
+by  generating the dynamics and outgoing radiation in all possible directions once and for all for fixed
+$\mathbf{\lambda}$, using a spherical harmonic decomposition.  Using this cached information effectively,  the likelihood can be evaluated as a function of $\theta$ at very low computational cost.  
 % From ROM paper
@@
 \texttt{numpy} code used for the CPU-based version of ILE.    For the most costly part of the calculation -- the inner
-products necessary to evaluate $Q_{lm}(t)$ -- we use a custom CUDA kernel.  
+products necessary to evaluate $Q_{lm}(t)$, accounting for distinct time-of-flight-dependent time windows for each
+interferometer's data -- we use a custom CUDA kernel to perform the necessary matrix multiplication.  
 %
@@
 
-\section{Validation and Performance}
+\section{Analysis of marginalized likelihood evaluation cost}
 \label{sec:demo}
@@
 neutron star with masses $1.4 M_\odot$ and $1.35 M_\odot$.
-We generate  synthetic data for a two-detector LIGO configuration, assuming operation at  initial (BBH) or advanced (BNS) design  sensitivity.
-Both signals are generated with SEOBNRv4.    We perform a multi-stage iterative RIFT analysis of these signal with the
-time-domain SEOBNRv4 (BBH) or TaylorT4 (BNS) approximation, under the simplifying assumption that both objects are point particles with zero
-spin.  We repeat this analysis on heterogeneous LVC collaboration computing resources, to assess our variable
+ We perform this analysis on heterogeneous LVC collaboration computing resources, to assess our variable
 performance across architectures.  Notably, we investigate the following GPU options: (a) GTX 1050 Ti at LIGO-CIT,
@@
 GTX 1050 at LIGO-LHO.    Unless otherwise noted in the text, we will discuss code configurations using CPU-only and GPU (b) using
-$4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.  We also conservatively report computational cost for
-parameters $\lambda$ which are close to or within the posterior and therefore, due to their consistency with the data,
-have the best-determined extrinsic parameters and thus converge the most slowly.
+$4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.  In this section, we  conservatively report computational cost for
+parameters $\lambda$ which are close to or within the posterior.  Because of their consistency with the data, 
+the marginalized likelihood calculations converge the most slowly, as they have the best-determined extrinsic parameters.
 
-\editremark{report on distance to each signal, and in particular $d_{\rm max}$ as that might impact performance too}
-chosen to have peak SNR $\simeq 14$ and
+\subsection{Synthetic source generation and analysis settings}
+
+We generate  synthetic data for a two-detector LIGO configuration, assuming operation at  initial (BBH) or advanced (BNS) design  sensitivity.
+Both signals are generated with SEOBNRv4.   
+To qualitatively reproduce the noise power spectrum and amplitude of typical binary black hole observations in O1 and
+O2, the binary black hole signal has source distance $200 \unit{Mpc}$, chosen so SNR $\simeq 14$.  
+Similarly, to qualitatively reproduce the analysis of GW170817, our synthetic binary neutron star is placed $100
+\unit{Mpc}$ away, so the signal amplitude is SNR $\simeq 31$.
+
+ We perform a multi-stage iterative RIFT analysis of these signal with the
+time-domain SEOBNRv4 (BBH) or TaylorT4 (BNS) approximation, under the simplifying assumption that both objects are point particles with zero
+spin.   For both sources, we adopt the fiducial distance prior $p(D_L) \propto D_L^2$,   selecting  a maximum distance
+roughly five times larger than the known source distance: 1Gpc for the binary black hole and 500 Mpc for the neutron star.
 
@@
 cost $\tau_{mc}=N_{ad} \tau_{ad}+N_{it}\tau_{eval}$ increases with the number of Monte Carlo iterations $N_{it}$ with
-and the number of times $N_{ad}$ the sampling prior used in adaptive Monte Carlo is regenerated, in proportion to their
+and the number of times $N_{ad}$ the sampling prior used in adaptive Monte Carlo is regenerated from the most recent 
+$n_{\rm chunk}$ data points, in proportion to their
 cost.   In the standard configuration, the adaptive sampling prior is regenerated every $n_{\rm chunk}$ iterations (i.e,
@@
 \emph{and} the cost per evaluation $\tau_{eval}$ decrease modestly because of this effect.  
-Also, to insure safe inner products over short data sets in the presence of very narrow spectral lines, we operated on a
-significantly-padded data buffer,
-using 
-``inverse spectrum truncation'' \editremark{citation}.  For binary black holes, the degree of padding was significantly
-in excess of the signal duration \editremark{numbers: say 64s for a BBH!}.  
-Instead and following LI, we adopt a much shorter inverse spectrum truncation length, a much shorter padding buffer, and
+Also following Pankow et al, to insure safe inner products over short data sets in the presence of very narrow spectral lines, we operated on a
+significantly-padded data buffer and modified the noise power spectrum by truncating the inverse power spectrum to a
+finite response time.  For binary black holes, the degree of padding was significantly
+in excess of the signal duration.  
+Instead and following lalinference \cite{gw-astro-PE-lalinference-v1}, we adopt a much shorter inverse spectrum truncation length, a much shorter padding buffer, and
 a Tukey window applied to the data to remove discontinuities at the start and end of each segment.   
@@
  + 
- [ \tau_{setup}/s' + N_{ad} \tau_{ad} + N_{it}\tau_{it}
+ [ \tau_{setup} + N_{ad} \tau_{ad} + N_{it}\tau_{it}
  ] 
 \end{align}
-For a typical binary black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$, this
-version requires $\tau_{start} \simeq 12\unit{s}$, $\tau_{setup}/ss' \simeq XX$, $\tau_{ad}\simeq XXX$, and $N_{it}\tau_{\it} \simeq XX$
-\editremark{numbers}.  
+The first rows in table Table  \ref{tab:CostBreakdown} shows our estimated breakdown of these elements of the computational cost, for a typical binary black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$.
 Because of the extremely high cost of adaptive overhead, we only adapt in two parameters, corresponding to sky location; 
-and only adapt during the first iteration which returns significant $\ln {\cal L}_{\rm marg}$.  As a result, $N_{ad}$
-(the number of adaptive stages \emph{per likelihood evaluation}) scales as $N_{it}/N_{chunk}/N_{\rm eval}$ and does not
+and only adapt until a marginalied likelihood evaluation  returns significant $\ln {\cal L}_{\rm marg}$.  As a result, $N_{ad}$
+(the number of adaptive stages \emph{per likelihood evaluation}) scales as $N_{it}/n_{chunk}/N_{\rm eval}$ and does not
 increase with the overall number of samples, so the overall evaluation time scales as
@@
  + 
- [ \tau_{setup}/s'  + N_{it}\tau_{it}
+ [ \tau_{setup}  + N_{it}\tau_{it}
  ] 
@@
 & \left. + 
- [ \tau_{setup}/ss'  + N_{it}\tau_{it}/s
+ [ \tau_{setup}  + N_{it}\tau_{it}
  ] 
@@
 %   (165 - 29.  )/20/100   : looks odd for 16 kHz
-GPU (b) & 16384 & $\pm 2, \pm 1$   & 35 & 26 & 0.07 &  2.4 & &35 &  \\
+GPU (b) & 16384 & $\pm 2, \pm 1$   & 35 & 26 & 0.07 &  2.4 & &35 & 16 \\
        & 4096 & $ \pm 2, \pm 1 $        &  35  & 11.5 & 0.1 &  2.4  & &24 &  \\ \hline
+GPU (c) & 16384 & $\pm 2, \pm 1$   & 35 &  &&&& 105  \\
+       & 4096 & $ \pm 2, \pm 1 $        &  35  & 28 &&&& 55 \\ \hline
 \end{tabular}
@@
 
-\editremark{test on other hardware -- also problem, a lot of the speed improvement occurs because the initial grid is
-  crap/too wide}
-
 For the closed-form post-Newtonian approximation used in the study above, waveform costs are a modest contribution to
@@
 
+While GPU cards have finite memory, the modest size of our underlying data arrays and intemediate data products on-board ensures that we are unlikely to
+saturate this bound, even for 16 kHz data, unless we investigate significantly lower starting frequencies or
+vastly more modes.
+%\editremark{some issue: memory usage is about 1Gb for 16 kHz,  close to $4Gb$ limit on commodity cards now.  Will be hard to
+%  increase mode set substantially for BNS on the cheapest cards  }
+
+\section{Performance and validation demonstrations with full pipeline}
+\label{sec:end-to-end}
+
+To better illustrate code performance in realistic settings, we also describe end-to-end analyses with
+the original and modified code.     In a full analysis, many of the initially-proposed  evaluation points $\lambda$ are
+either not or marginally consistent with the data.  As a result, most marginalized likelihood evaluations in the first
+iteration proceed extremely rapidly.  As the high-cost and low-cost evaluations are generally well-mixed between
+different instances, the overall time to complete the full first iteration is typically much lower than subsequent
+iterations.   With a well-positioned and sufficiently large initial grid,  few follow-on iterations are needed.  
+%   Moreover, because the code progresses slowly only on candidate points which are most consistent with the data,
+
+%* more variable, because many evaluations early on \emph{are not} consistent, so only a small fraction of the evaluations
+%needed are consistent with the observations ... convergence can be faster as the signal amplitude decreases
+
+\subsection{Binary neutron star analysis}
+\label{sec:sub:BNSFull}
+
+\begin{figure*}
+\caption{\label{fig:BNS:MultiIterate}\textbf{Convergence of BNS analysis}: Results for marginal posterior distributions
+  of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
+  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
+  value of the underlying marginalized likelihood evaluations.   The corresponding dotted curves show an analysis using
+  only the $m=\pm 2$ modes \editremark{perform both} \emph{Left panel } Posterior distribution
+  over  $\mc$ and
+  $\delta=(m_1-m_2)/M$.    \emph{Center panel}: Marginal 1d CDFs of $\mc$, showing convergence.
+\emph{Right panel}: Marginal 1d empirical CDF of $\ln{\cal L}_{\rm marg}$, showing that at late times the proposed
+candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$.
+}
+\end{figure*}
+
+% Code configuration
+In this example, we use $100$ initial points spread over the two mass dimensions, adding $20$ evaluations per
+iteration.  With one instance analyzing each 20 evaluations,  we use five cores in the first iteration, followed by one
+core for remaining iterations.  We choose this simple configuration to facilitate visualization.    The left and center panels of Figure  \ref{fig:BNS:MultiIterate} show posterior disributions derived
+from the first several RIFT iterations.
+
+* show convergence diagnostic of $\ln {\cal L}$ range over the data: can see few outliers.  Easy way to assess agreement
+and stability
+
+** can also do L2 difference between likelihoods estimated with different data
+
+* COST OF OVERALL ANALYSIS, AVERAGED OVER WHOLE: average cost of $\ln {\cal L}$ is
+4096 Hz is X.  First iteration averages about 30 minutes in this configuration, second and subsequent iterations require
+about 2 h.
+
+Binary neutron star models with more parameters like spin and tides require correspondingly larger numbers of points in the
+  initial grid and per iteration.  Fortunately, the number of observationally significant and accessible dimensions is often
+  substantially less than the a prior necessary dimensionality.  In practice,we use roughly $20\times$ more points per
+  iteration for precessing massive BBH systems ($d=8$) or for spinning binary neutron stars with tides ($d=6$).
+
 \section{ Results (optional?)}
 
-* Add Jake as author.  Rerun all O2 BBHs with SEOBNRv3?
+*   Rerun all O2 BBHs with SEOBNRv3?
+
+* PP plot demo for zero-spin binaries, with profiling of full run?
+
+* reminder: HM for BNS break degeneracy, useful for source classification.  Show an example.
 
 \section{Conclusions}
+\label{sec:conclude}
 We have demonstrated that the marginalized likelihood $\ln {\cal L}_{\rm marg}(\lambda)$ appearing in the RIFT/rapidPE parameter
@@
 
+By allowing  models with higher-order modes to be used in low latency, our code can better-constrain the properties of low-mass
+binaries with suitable amplitudes and orientations \editremark{citations to old work}, as demonstrated in Section
+\ref{sec:sub:BNSFull}.  These tighter constraints could
+better  inform low-latency source classification and hence multimessenger followup observations of compact binary mergers.
+
+Beyond low-latency multimessenger astronomy, rapid parameter inference enables new applications.  For example, every
+parameter inference provide \emph{evidence} for a signal being present in the data; with rapid parameter inference, this
+evidence could be used as (the last stage in a hierarchical pipeline for) a detection statistic
+\editremark{citations:old idea,see Thrane/Rory papers recently} \cite{2018PhRvX...8b1019S}.    This approach can identify individual events and
+even a population.
+Alternatively and in many ways equivalently, one can identify a population of GW sources without assuming any one is
+real, by  applying parameer inference to more candidate events and self-consistently seperating foreground and
+background. 
+\cite{2015PhRvD..91b3005F,2019MNRAS.tmp..230G}.
+
 The use of \texttt{cupy} enables our code to be highly portable across architectures and heterogeneous GPU environments,

### 2019-02-02 paper/paper.tex
paper.tex: finish updating BNS sec

@@
 \begin{figure*}
+% COMMAND
+% plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
+
 \caption{\label{fig:BNS:MultiIterate}\textbf{Convergence of BNS analysis}: Results for marginal posterior distributions
@@
   over  $\mc$ and
-  $\delta=(m_1-m_2)/M$.    \emph{Center panel}: Marginal 1d CDFs of $\mc$, showing convergence.
-\emph{Right panel}: Marginal 1d empirical CDF of $\ln{\cal L}_{\rm marg}$, showing that at late times the proposed
+  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
+\emph{Bottom left}: Marginal 1d empirical CDF of $\ln{\cal L}_{\rm marg}$, showing that at late times the proposed
 candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$.
+\emph{Bottom right panel}: $\ln {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
+error budget early on, the Monte Carlo error is only shown for the last iteration.
 }
@@
 
+\begin{figure*}
+\caption{\label{fig:BNS:Comparison}\textbf{Comparison of GPU, CPU analysis}: \editremark{worthwhile}
+}
+\end{figure*}
+
+% 
+In our first test we compare two workflows, one with the original embarassingly-parallel RIFT code using $N_{\rm
+  eval}=1$ and one with the
+``batched'' GPU-accelerated code with $N_{\rm eval}=20$, analyzing data with $f_{\rm sample}=4096\unit{Hz}$.    
 % Code configuration
-In this example, we use $100$ initial points spread over the two mass dimensions, adding $20$ evaluations per
-iteration.  With one instance analyzing each 20 evaluations,  we use five cores in the first iteration, followed by one
-core for remaining iterations.  We choose this simple configuration to facilitate visualization.    The left and center panels of Figure  \ref{fig:BNS:MultiIterate} show posterior disributions derived
-from the first several RIFT iterations.
+In this example, we use $100$ initial points spread over the two mass dimensions $\mc_z,\delta$, adding $20$ evaluations per
+iteration.   We
+choose this simple configuration to facilitate visualization.
+The initial coarse grid covers a region $\mc_z\in[1.1962,1.1970] M_\odot$ and $\delta \in [0,0.25]$,  insuring the posterior was smaller than our initial
+coarse grid placement.     The top  panels of Figure  \ref{fig:BNS:MultiIterate} show posterior disributions derived
+from the first several RIFT iterations, while the bottom panels show convergence diagnostics.
+Figure \ref{fig:BNS:Comparison} compares results from the GPU and CPU codes on \editremark{identical test points: the
+  initial grid}.\footnote{This comparison of $\ln {\cal L}_{\rm marg}$ is limited by Monte Carlo integration error and
+  is provided  as a pedagogically convenient sanity check.  When comparing 
+  $\ln {\cal L}_{\rm margT}$ results on identical data and parameters, the GPU and CPU implementations agree to
+  significantly higher precision.}  The two results agree.
+Because our analysis uses a post-Newtonian model to interpret a signal generated with SEOBNRv4, all results are offset
+slightly from the synthetic signal's parameters
 
-* show convergence diagnostic of $\ln {\cal L}$ range over the data: can see few outliers.  Easy way to assess agreement
-and stability
+This specific workflow configuration  reduces overall core usage by maximizing GPU use per iteration: after the first
+iteration, only one GPU is active.   
+Specifically,  with the updated RIFT workflow used here, with one instance analyzing each 20 evaluations,  we use five
+core+GPU pairs in the
+first iteration, followed by one core+GPU for remaining iterations.
+By contrast, with the original CPU-only RIFT code analyzing each $\lambda$ in parallel, this process requires roughly 20 core-minutes
+per $\lambda$, using 100 cores in the first iteration and 20 cores in each
+subsequent evalution.   
+Note that  because   $N_{\rm eval}=20$ for the GPU is larger than the (hardware-and $f_{\rm sample}$ dependent)  speedup factor between the CPU and
+GPU implementation, the overall wallclock time needed for a end-to-end
+analysis analysis is larger for the GPU workflow.
+We can achieve comparable or smaller turnaround time for an otherwise identical analysis by adjusting $N_{\rm eval}$ appropriate to the available hardware.
 
-** can also do L2 difference between likelihoods estimated with different data
-
-* COST OF OVERALL ANALYSIS, AVERAGED OVER WHOLE: average cost of $\ln {\cal L}$ is
-4096 Hz is X.  First iteration averages about 30 minutes in this configuration, second and subsequent iterations require
-about 2 h.
+\editremark{modify output size: want more than 3000 samples, probably want 10k to make sure we are stable and to
+  minimize MC error}
 

### 2019-02-02 paper/paper.tex
paper.tex: add section with aligned spin, as example with production-scale analysis

@@
 roughly five times larger than the known source distance: 1Gpc for the binary black hole and 500 Mpc for the neutron star.
+For the binary black hole we generate the signal with $f_{\rm min}=8 \unit{Hz}$ and analyze the signal with  $f_{\rm
+  min}=10 \unit{Hz}$; for the binary neutron star, we generate the signal with $f_{\rm min}=20\unit{Hz}$ and anayze it
+with $f_{\rm min}=23\unit{Hz}$.
 
@@
 likelihood evaluations will complete in about $100\; \unit{minutes}$ on a cluster of (b), or more realistically 200
-minutes on a cluster of (a). . This number can be reduced to
+minutes on a cluster of (a).  This number can be reduced to
 tens of minutes with a modestly larger GPU pool, or by  marginally more conservative convergence thresholds. on $N_{\rm
@@
 
-\subsection{Binary neutron star analysis}
+\subsection{Binary neutron star analysis : Assuming  zero spin}
 \label{sec:sub:BNSFull}
@@
 
-\caption{\label{fig:BNS:MultiIterate}\textbf{Convergence of BNS analysis}: Results for marginal posterior distributions
+\caption{\label{fig:BNS:MultiIterate}\textbf{Convergence of BNS analysis: Zero spin}: Results for marginal posterior distributions
   of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
@@
 
+\subsection{Binary neutron star analysis : Assuming  nonprecessing spin}
+\label{sec:sub:BNS_spin}
+
 Binary neutron star models with more parameters like spin and tides require correspondingly larger numbers of points in the
@@
   substantially less than the a prior necessary dimensionality.  In practice,we use roughly $20\times$ more points per
-  iteration for precessing massive BBH systems ($d=8$) or for spinning binary neutron stars with tides ($d=6$).
+  iteration for precessing massive BBH systems ($d=8$) or for spinning binary neutron stars with tides ($d=6$).    As a
+  result, we can still achieve relatively rapid turnaround on a high-dimensional binary neutron star analysis.
+
+As a concrete demonstration of a low-latency analysis using the GPU-accelerated code, we reanalyze our fiducial binary
+neutron star signal, accounting for the possibility of nonzero (aligned) neutron star spins
+$\chi_{i,z}\in[-0.05,0.05]$.  
+Because  our fiducial post-Newtonian model (TaylorT4) as implemented in \texttt{lalsuite} does not include spin, we employ the \texttt{SEOBNRv4\_ROM} waveform approximation.
+%
+Figure \ref{fig:BNS:Spin} shows our results.
+\editremark{Key feature: runtime: LHO (GPU (c)), 40-ish GPU nodes.  Possibly also overlay SEOBNRv4 analysis as sanity check}
+
+\begin{figure*}
+% COMMAND
+% plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
+
+\caption{\label{fig:BNS:Spin}\textbf{Demonstration of low-latency spinning analysis}:XXX
+}
+\end{figure*}
 
@@
 
+\editremark{bonus mentions}: eccentricity
+
+Bonus code: lnLcut tapering, possibly neff tapering
+
+Puffball in DAG every few iterations, to insure stability around edges and not overcover the core
+
 \begin{acknowledgements}

### 2019-02-02 demos/convergence_data/parse_convergence.sh
parse_convergence.sh: useful intermediate script

@@ -1,0 +1,3 @@
+#! /bin/bash
+cat $1/*withpriorchange+annotation.dat | grep -v '#'  | awk '{print $1, $3}' > tmp.dat
+cat  -n tmp.dat

### 2019-02-02 demos/convergence_data/plot_convergence.py
plot_convergence.py: utility to plot lnL vs iteration number

@@ -1,0 +1,129 @@
+#! /usr/bin/env python
+
+# plot_posterior_corner.py
+#
+#  --posterior-file f1 --posterior-label n1 --posterior-file f2 --posterior-label n2 ...
+#  --parameter p1 --parameter p2 ...
+#
+# EXAMPLE
+#    python plot_posterior_corner.py --posterior-file downloads/TidalP4.dat --parameter lambda1 --parameter lambda2 --parameter mc
+#   python plot_posterior_corner.py --parameter mc --parameter eta --posterior-file G298048/production_C00_cleaned_TaylorT4/posterior-samples.dat  --parameter lambdat
+#
+# USAGE
+#    - hardcoded list of colors, used in order, for multiple plots
+#
+
+import lalsimutils
+import lal
+import numpy as np
+import argparse
+
+try:
+    import matplotlib
+    print " Matplotlib backend ", matplotlib.get_backend()
+    if matplotlib.get_backend() is 'agg':
+        fig_extension = '.png'
+        bNoInteractivePlots=True
+    else:
+	matplotlib.use('agg')
+        fig_extension = '.png'
+        bNoInteractivePlots =True
+    from matplotlib import pyplot as plt
+    bNoPlots=False
+except:
+    print " Error setting backend"
+
+import matplotlib.pyplot as plt
+from mpl_toolkits.mplot3d import Axes3D
+import matplotlib.lines as mlines
+import corner
+
+import our_corner
+try:
+    import bounded_kde
+except:
+    print " -No 1d kdes- "
+
+dpi_base=200
+legend_font_base=16
+rc_params = {'backend': 'ps',
+             'axes.labelsize': 11,
+             'axes.titlesize': 10,
+             'font.size': 11,
+             'legend.fontsize': legend_font_base,
+             'xtick.labelsize': 11,
+             'ytick.labelsize': 11,
+             #'text.usetex': True,
+             'font.family': 'Times New Roman'}#,
+             #'font.sans-serif': ['Bitstream Vera Sans']}#,
+plt.rcParams.update(rc_params)
+
+
+# Parse arguments
+parser = argparse.ArgumentParser()
+parser.add_argument("--convergence-file",action='append',help="filename of *.dat file [standard LI output]")
+parser.add_argument("--output",default='convergence',help="filename for plot")
+parser.add_argument("--pdf",action='store_true',help="filename for plot")
[…]

### 2019-02-03 demos/convergence_data/plot_mean_variance.py
plot scripts: update to be sensible

@@ -119,11 +119,11 @@ for pIndex in np.arange(len(convergence_list)):
     dat = np.zeros( (len(samples),3) )
     dat[:,0] = samples[:,0]
     dat[:,1] = samples[:,1]
-    dat[:,2] = 1/np.sqrt(samples[:,2])
+    dat[:,2] = samples[:,2]
     my_cmap_values = color_list[pIndex]
-    plt.plot(dat[:,0], dat[:,1],c=my_cmap_values,ls=linestyle_list[pIndex])
+    plt.errorbar(dat[:,0], dat[:,1],yerr=dat[:,2],c=my_cmap_values,ls=linestyle_list[pIndex])
 
 plt.xlabel("iteration")
 plt.xticks( dat[:,0])
-plt.ylabel(r'$\ln Z$')
+plt.ylabel(r'$\ln {\cal L}_{\rm marg}$')
 plt.savefig(opts.output+fig_extension)        # use more resolution, to make sure each image remains of consistent quality

### 2019-02-03 paper/paper.tex
paper.tex: add figure prototypes

@@
 parameters $\lambda$ which are close to or within the posterior.  Because of their consistency with the data, 
-the marginalized likelihood calculations converge the most slowly, as they have the best-determined extrinsic parameters.
+the marginalized likelihood calculations converge the most slowly, as they have the best-determined extrinsic
+parameters.
+
+Unless otherwise noted, all mass quantities described in this work are redshifted masses $m_{k,z}=(1+z)m_k$.
 
@@
 We generate  synthetic data for a two-detector LIGO configuration, assuming operation at  initial (BBH) or advanced (BNS) design  sensitivity.
-Both signals are generated with SEOBNRv4.   
+Both signals are generated with SEOBNRv4, with a zero-noise data realization.   
 To qualitatively reproduce the noise power spectrum and amplitude of typical binary black hole observations in O1 and
@@
 % plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
-
+\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_corner_mc_delta_mc.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_mc_cum.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_meanVar.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_converge.png}
 \caption{\label{fig:BNS:MultiIterate}\textbf{Convergence of BNS analysis: Zero spin}: Results for marginal posterior distributions
@@
   $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
-\emph{Bottom left}: Marginal 1d empirical CDF of $\ln{\cal L}_{\rm marg}$, showing that at late times the proposed
-candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$.
+\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
+first iteration the
+candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
 \emph{Bottom right panel}: $\ln {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
@@
 iteration.   We
-choose this simple configuration to facilitate visualization.
+choose this simple low-dimension, small-size, and slowly-converging configuration to facilitate visualization of the grid, posterior, and convergence.
 The initial coarse grid covers a region $\mc_z\in[1.1962,1.1970] M_\odot$ and $\delta \in [0,0.25]$,  insuring the posterior was smaller than our initial
-coarse grid placement.     The top  panels of Figure  \ref{fig:BNS:MultiIterate} show posterior disributions derived
+coarse grid spacing.     The top  panels of Figure  \ref{fig:BNS:MultiIterate} show posterior disributions derived
 from the first several RIFT iterations, while the bottom panels show convergence diagnostics.
@@
   significantly higher precision.}  The two results agree.
-Because our analysis uses a post-Newtonian model to interpret a signal generated with SEOBNRv4, all results are offset
-slightly from the synthetic signal's parameters
+Because our analysis uses a post-Newtonian model to interpret a signal generated with SEOBNRv4, the peak posterior
+density is slightly  offset
+ from the synthetic signal's parameters
 
@@
 We can achieve comparable or smaller turnaround time for an otherwise identical analysis by adjusting $N_{\rm eval}$ appropriate to the available hardware.
-
-\editremark{modify output size: want more than 3000 samples, probably want 10k to make sure we are stable and to
-  minimize MC error}
 

### 2019-02-03 paper/paper.tex
paper.tex: working on BNS section

@@
 \newcommand\ForInternalReference[1]{}
-%\newcommand\SkipForEarlyCirculation[1]{}
-\newcommand\SkipForEarlyCirculation[1]{#1}
-%\newcommand\SkipBoxing[1]{}
-\newcommand\SkipBoxing[1]{#1}
+\newcommand\SkipForEarlyCirculation[1]{}
+%\newcommand\SkipForEarlyCirculation[1]{#1}
 \newcommand\SkipPP[1]{}
@@
 
+\subsection{Binary black hole analysis}
+\label{sec:sub:BBHFull}
+
+\begin{figure*}
+% COMMAND
+% plot_posterior_corner.py --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --posterior-label 2 --posterior-label 1 --use-legend
+\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_m1_m2.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_mc_cum.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_meanVar.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_converge.png}
+\caption{\label{fig:BBH:MultiIterate}\textbf{Convergence of BBH analysis: Zero spin}: Results for marginal posterior distributions
+  of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
+  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
+  value of the underlying marginalized likelihood evaluations.   The corresponding dotted curves show an analysis using
+  only the $m=\pm 2$ modes \editremark{perform both} \emph{Left panel } Posterior distribution
+  over  $\mc$ and
+  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
+\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
+first iteration the
+candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
+\emph{Bottom right panel}: $\ln {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
+error budget early on, the Monte Carlo error is only shown for the last iteration.
+}
+\end{figure*}
+
+Modest-amplitude short-duration binary black holes empirically constitute the most frequent detection candidates
+for current ground-based GW observatories \cite{LIGO-O2-Catalog}.    Because of their brevity and hence broad posterior,
+the RIFT code's interpolation-based method converges  rapidly.  Combined with the low cost of each iteration,
+these sources require an exceptionally low committment of resources and can be performed in extremely low latency, as
+desired.
+To demonstrate this, we use the GPU-accelerated code with $N_{\rm eval}=20$, analyzing data with $f_{\rm
+  sample}=4096\unit{Hz}$.    We use 100 points in the first iteration, in a coarse mass  grid with astrophysically significant
+scale, followed by 20 points in each subsequent iteration.  Figure \ref{fig:BBH:MultiIterate} shows our results.   The
+final posterior is already well-explored with the initial gridpoints, and converged by the second iteration.  
+Conversely, on average all iterations required roughly 45 seconds per $\lambda$ to evaluate their grid.  This
+configuration therefore converged within 30 minutes.    We can achieve  smaller turnaround time for an otherwise identical analysis by adjusting $N_{\rm eval}$ appropriate to the available hardware.
+%
+While this low-dimensional problem does not capture all  fitting and automation challenges associated with high-dimensional
+ fully precessing binaries, it does capture the low cost and rapid response possible with RIFT.
+
 \subsection{Binary neutron star analysis : Assuming  zero spin}
@@
 
+Significant-amplitude  binary neutron star mergers are the most important scenario for rapid parameter inference, as low
+latency can enable multimessenger followup  \cite{LIGO-O2-Catalog}.    
 % 
-In our first test we compare two workflows, one with the original embarassingly-parallel RIFT code using $N_{\rm
+In our second test we compare two workflows, one with the original embarassingly-parallel RIFT code using $N_{\rm
   eval}=1$ and one with the
@@
 $\chi_{i,z}\in[-0.05,0.05]$.  
-Because  our fiducial post-Newtonian model (TaylorT4) as implemented in \texttt{lalsuite} does not include spin, we employ the \texttt{SEOBNRv4\_ROM} waveform approximation.
+Because  our fiducial post-Newtonian model (TaylorT4) as implemented in \texttt{lalsuite} does not include spin, we
+employ both the \texttt{SEOBNRv4\_ROM} and \texttt{SpinTaylorT4} waveform approximations for nonprecessing binaries,
+omitting higher-order modes.
 %
@@
 
+\SkipForEarlyCirculation{
 \section{ Results (optional?)}
@@
 * reminder: HM for BNS break degeneracy, useful for source classification.  Show an example.
+
+\editremark{bonus mentions}: eccentricity
+
+Bonus code: lnLcut tapering, possibly neff tapering
+
+Puffball in DAG every few iterations, to insure stability around edges and not overcover the core
+
+}
 
@@
 
-\editremark{bonus mentions}: eccentricity
-
-Bonus code: lnLcut tapering, possibly neff tapering
-
-Puffball in DAG every few iterations, to insure stability around edges and not overcover the core
-
 \begin{acknowledgements}

### 2019-02-03 paper/paper.tex
paper.tex: draft almost ready for P&P circulation

@@
 \affiliation{\RIT}
-\author{Jake?}
+\author{Jacob Lange}  % for all the convergence material we work on ... it comes from Jake's practice
 \affiliation{\RIT}
@@
 \begin{figure*}
+% BASED ON
+%   CIT: test_workflow_batch_gpu_lowlatency
 % COMMAND
-% plot_posterior_corner.py --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --posterior-label 2 --posterior-label 1 --use-legend
-\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_m1_m2.png}
-\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_mc_cum.png}
-\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_meanVar.png}
-\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_converge.png}
+%  plot_posterior_corner.py --posterior-file posterior_samples-4.dat  --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
+%plot_posterior_corner.py --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --posterior-label 2 --posterior-label 1 --use-legend
+\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_m2.png}
+\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_cum.png}
+\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_meanVar.png}
+\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_converge.png}
 \caption{\label{fig:BBH:MultiIterate}\textbf{Convergence of BBH analysis: Zero spin}: Results for marginal posterior distributions
@@
   show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
-  value of the underlying marginalized likelihood evaluations.   The corresponding dotted curves show an analysis using
-  only the $m=\pm 2$ modes \editremark{perform both} \emph{Left panel } Posterior distribution
+  value of the underlying marginalized likelihood evaluations.   
+%The corresponding dotted curves show an analysis using   only the $m=\pm 2$ modes \editremark{perform both} 
+\emph{Left panel } Posterior distribution
   over  $\mc$ and
@@
 candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
-\emph{Bottom right panel}: $\ln {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
-error budget early on, the Monte Carlo error is only shown for the last iteration.
+\emph{Bottom right panel}: The estimated evidence $Z = \int d\lambda {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
+error budget, Monte Carlo error is not shown.
 }
@@
 To demonstrate this, we use the GPU-accelerated code with $N_{\rm eval}=20$, analyzing data with $f_{\rm
-  sample}=4096\unit{Hz}$.    We use 100 points in the first iteration, in a coarse mass  grid with astrophysically significant
-scale, followed by 20 points in each subsequent iteration.  Figure \ref{fig:BBH:MultiIterate} shows our results.   The
+  sample}=4096\unit{Hz}$.    We use 100 points in the first iteration, in a very coarse mass  grid (i.e., spacing
+comparable to typical astrophysical mass scales), followed by 20 points in each subsequent iteration.  Figure \ref{fig:BBH:MultiIterate} shows our results.   The
 final posterior is already well-explored with the initial gridpoints, and converged by the second iteration.  
-Conversely, on average all iterations required roughly 45 seconds per $\lambda$ to evaluate their grid.  This
-configuration therefore converged within 30 minutes.    We can achieve  smaller turnaround time for an otherwise identical analysis by adjusting $N_{\rm eval}$ appropriate to the available hardware.
+Conversely, on average all iterations required roughly 45 seconds per $\lambda$ to evaluate their grid on GPU (a) hardware.  This
+configuration therefore converged within the 30 minutes needed for the first two iterations.
+We can achieve  smaller turnaround time for an otherwise identical analysis by adjusting $N_{\rm eval}$ appropriate to the available hardware.
 %
@@
   show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
-  value of the underlying marginalized likelihood evaluations.   The corresponding dotted curves show an analysis using
-  only the $m=\pm 2$ modes \editremark{perform both} \emph{Left panel } Posterior distribution
+  value of the underlying marginalized likelihood evaluations.  
+% The corresponding dotted curves show an analysis using   only the $m=\pm 2$ modes \editremark{perform both} 
+\emph{Left panel } Posterior distribution
   over  $\mc$ and
@@
 candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
-\emph{Bottom right panel}: $\ln {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
-error budget early on, the Monte Carlo error is only shown for the last iteration.
+\emph{Bottom right panel}: $Z = \int d\lambda {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
+error budget early on, Monte Carlo error is not shown.
 }
@@
 
-\begin{figure*}
-\caption{\label{fig:BNS:Comparison}\textbf{Comparison of GPU, CPU analysis}: \editremark{worthwhile}
-}
-\end{figure*}
+%% \begin{figure*}
+%% \caption{\label{fig:BNS:Comparison}\textbf{Comparison of GPU, CPU analysis}: \editremark{worthwhile}
+%% }
+%% \end{figure*}
 
@@
 from the first several RIFT iterations, while the bottom panels show convergence diagnostics.
-Figure \ref{fig:BNS:Comparison} compares results from the GPU and CPU codes on \editremark{identical test points: the
-  initial grid}.\footnote{This comparison of $\ln {\cal L}_{\rm marg}$ is limited by Monte Carlo integration error and
-  is provided  as a pedagogically convenient sanity check.  When comparing 
-  $\ln {\cal L}_{\rm margT}$ results on identical data and parameters, the GPU and CPU implementations agree to
-  significantly higher precision.}  The two results agree.
+%% Figure \ref{fig:BNS:Comparison} compares results from the GPU and CPU codes on \editremark{identical test points: the
+%%   initial grid}.\footnote{This comparison of $\ln {\cal L}_{\rm marg}$ is limited by Monte Carlo integration error and
+%%   is provided  as a pedagogically convenient sanity check.  When comparing 
+%%   $\ln {\cal L}_{\rm margT}$ results on identical data and parameters, the GPU and CPU implementations agree to
+%%   significantly higher precision.}  The two results agree.
 Because our analysis uses a post-Newtonian model to interpret a signal generated with SEOBNRv4, the peak posterior
@@
 
-As a concrete demonstration of a low-latency analysis using the GPU-accelerated code, we reanalyze our fiducial binary
-neutron star signal, accounting for the possibility of nonzero (aligned) neutron star spins
-$\chi_{i,z}\in[-0.05,0.05]$.  
+As a concrete demonstration of a realistic modest-latency analysis using the GPU-accelerated code, we reanalyze our fiducial binary
+neutron star signal, accounting for the possibility of nonzero (aligned) neutron star spins.  Our initial grid consists of 5000 points, spread uniformly across a 4-dimensional
+hypercube $\mc_z\in[1.1962,1.1970] M_\odot$, $\delta \in[0,0.25]$ and $\chi_{i,z}\in[-0.05,0.05]$.  Subsequent
+iterations use 500 random draws from the estimated posterior samples produced from the previous iteration.  
 Because  our fiducial post-Newtonian model (TaylorT4) as implemented in \texttt{lalsuite} does not include spin, we
 employ both the \texttt{SEOBNRv4\_ROM} and \texttt{SpinTaylorT4} waveform approximations for nonprecessing binaries,
-omitting higher-order modes.
+omitting higher-order modes.   For SEOBNRv4\_ROM, we use 48 GPU (c) enabled nodes; for SpinTaylorT4, we use 100 GPU (a) nodes.
 %
 Figure \ref{fig:BNS:Spin} shows our results.
-\editremark{Key feature: runtime: LHO (GPU (c)), 40-ish GPU nodes.  Possibly also overlay SEOBNRv4 analysis as sanity check}
+%
+
+Due to our self-imposed resource constraints on $N_{\rm eval}=20$ and the number of GPUs, the first iteration is
+resource-limited and requires
+of order $(5000/20/100)\simeq
+2.5$ to $(5000/20/48)\simeq
+ 5$ times as long as the first iteration of the zero-spin BNS analysis described above.  Each marginalized likelihood
+ evaluation in the first requires roughly 60 seconds (SpinTaylorT4) or  (SEOBNRv4ROM), on average.   Subsequent iterations proceed
+ in  \editremark{timing}
 
@@
 % plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
+\includegraphics[width=0.45\textwidth]{figures/bns_withspin_corner_mc_delta_mc_xi.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_withspin_mc_cum.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_meanVar.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_converge.png}
+\caption{\label{fig:BNS:Spin}\textbf{Demonstration of low-latency spinning analysis}:
+Results for marginal posterior distributions
+  of our fiducial synthetic neutron star, assuming the binary has nonprecessing spins.  Solid contours show credible intervals; solid one-dimensional distributions
+  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
+  value of the underlying marginalized likelihood evaluations.  The solid curves show an analysis with SEOBNRv4\_ROM;
+  the dotted curves show an analysis with SpinTaylorT4.   \emph{Left panel } Posterior distribution
+  over  $\mc$,
+  $\delta=(m_1-m_2)/M$, and $\chi_{\rm eff}$.    \emph{Right panel}: Marginal 1d CDFs of $\delta$, showing convergence.
+\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
+first iteration the
+candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
+\emph{Bottom right panel}: $Z = \int d\lambda {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
+error budget early on, Monte Carlo error is not shown.
 
-\caption{\label{fig:BNS:Spin}\textbf{Demonstration of low-latency spinning analysis}:XXX
 }

### 2019-02-03 paper/paper.tex
paper.tex: minor edits

@@
 
-Unless otherwise noted, all mass quantities described in this work are redshifted masses $m_{k,z}=(1+z)m_k$.
-
 \subsection{Synthetic source generation and analysis settings}
@@
 spin.   For both sources, we adopt the fiducial distance prior $p(D_L) \propto D_L^2$,   selecting  a maximum distance
-roughly five times larger than the known source distance: 1Gpc for the binary black hole and 500 Mpc for the neutron star.
+roughly five times larger than the known source distance: 1Gpc for the binary black hole and 500 Mpc for the neutron
+star.  We adopt a uniform prior on the component (redshifted) masses $m_{i,z}$ (and, when appropriate, spins
+$\chi_{i,z}$).
+Unless otherwise noted, all mass quantities described in this work are redshifted masses $m_{k,z}=(1+z)m_k$.
 For the binary black hole we generate the signal with $f_{\rm min}=8 \unit{Hz}$ and analyze the signal with  $f_{\rm
@@
   iteration for precessing massive BBH systems ($d=8$) or for spinning binary neutron stars with tides ($d=6$).    As a
-  result, we can still achieve relatively rapid turnaround on a high-dimensional binary neutron star analysis.
+  result, we can still achieve relatively rapid turnaround on a high-dimensional binary neutron star analysis even in
+  resource-constrained environments.
 
@@
  5$ times as long as the first iteration of the zero-spin BNS analysis described above.  Each marginalized likelihood
- evaluation in the first requires roughly 60 seconds (SpinTaylorT4) or  (SEOBNRv4ROM), on average.   Subsequent iterations proceed
- in  \editremark{timing}
+ evaluation in the first requires roughly 60 seconds for both SpinTaylorT4 and SEOBNRv4ROM, on average.   Subsequent
+ iterations are not resource-limited and complete in roughly 20 minutes, depending on hardware \editremark{timing}
 
@@
 % based on Neval ~ 10^4
+Alternatively, if larger resource pools are available in low latency, both the original and now GPU-accelerated RIFT  can
+perform extremely rapid parameter inference if large iterations are performed completely in parallel (i.e., $N_{\rm
+  eval} \simeq 1$).
 
 By allowing  models with higher-order modes to be used in low latency, our code can better-constrain the properties of low-mass
-binaries with suitable amplitudes and orientations \editremark{citations to old work}, as demonstrated in Section
-\ref{sec:sub:BNSFull}.  These tighter constraints could
+binaries with suitable amplitudes and orientations, when circumstances allow.  These tighter constraints could
 better  inform low-latency source classification and hence multimessenger followup observations of compact binary mergers.
@@
 \begin{acknowledgements}
+DW thanks the RIT COS and CCRG for support, and  NSF-1707965.
+JL is supported by NSF PHY-1707965.  
 RO'S is supported by NSF PHY-1707965 and PHY-1607520.
-DW thanks the RIT COS and CCRG for support, and  NSF-1707965.
 Y-LLF was supported in part by BNL LDRD projects No.\,17-029 and No.\,19-002, 

### 2019-02-03 demos/convergence_data/parse_composite_convergence.py
parse_composite_convergence.py

@@ -1,0 +1,11 @@
+#! bin/bash
+
+import sys
+import numpy as np
+
+indx=1
+col_lnL=9
+for fname in sys.argv:
+    dat = np.loadtxt(fname)
+    print indx, np.mean(dat[:,col_lnL]), np.std(dat[:,col_lnL])
+    indx+=1

### 2019-02-03 paper/paper.tex
paper.tex: closing out BNS aligned spin example

@@
  evaluation in the first requires roughly 60 seconds for both SpinTaylorT4 and SEOBNRv4ROM, on average.   Subsequent
- iterations are not resource-limited and complete in roughly 20 minutes, depending on hardware \editremark{timing}
+marginalized likelihood iterations are not resource-limited and complete in roughly 35 minutes, depending on hardware.
 
@@
 % COMMAND
-% plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
-\includegraphics[width=0.45\textwidth]{figures/bns_withspin_corner_mc_delta_mc_xi.png}
+% plot_posterior_corner.py  --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc  --parameter chi_eff --mc-range  '[1.19655,1.197]' --chi-max 0.025 --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
+\includegraphics[width=0.45\textwidth]{figures/bns_withspin_corner_mc_delta_xi.png}
 \includegraphics[width=0.45\textwidth]{figures/bns_withspin_mc_cum.png}
@@
   value of the underlying marginalized likelihood evaluations.  The solid curves show an analysis with SEOBNRv4\_ROM;
-  the dotted curves show an analysis with SpinTaylorT4.   \emph{Left panel } Posterior distribution
+  the dotted curves show an analysis with SpinTaylorT4.  \editremark{currenly only v4ROM shown}  \emph{Left panel } Posterior distribution
   over  $\mc$,
@@
 error budget early on, Monte Carlo error is not shown.
-
 }
@@
 Carlo random number generation and inner products to GPU-based operations.  In such a configuration, the marginalized
-likelihood would carry out almost all calculations (except waveform generation) on a GPU, with minimal communication off
+likelihood code would perform almost all calculations (except waveform generation) on a GPU, with minimal communication off
 the board.  This configuration should further reduce the average time needed to compute $\ln {\cal L}_{\rm

### 2019-02-04 paper/paper.tex
paper.tex: mention Brandon and Pankow previous work

@@
 %
-In this paper we describe a variant of the RIFT algorithm which  translates its low-level
-algorithms to operate on graphics processing units (GPUs), and adjusts its workflow to exploit the speed improvements
+Previously, B. Miller et al \editremark{cite private communication} developed a custom-coded  implementation of the
+relevant likelihood on graphics processing units (GPUs), suggesting substantial performance improvements.  
+In this paper,  we describe a variant of the RIFT algorithm which flexibly translates its low-level
+algorithms to operate on GPUs, and adjusts its workflow to exploit the speed improvements
 this re-implementation affords.

### 2019-02-05 paper/paper.tex
paper.tex: lots of minor tweaks to profile data. **must automate** this

--- 2019-02-03 paper/paper.tex
@@
 %
-The RIFT strategy described in \cite{gwastro-PE-AlternativeArchitectures,2017CQGra..34n4002O,gwastro-PENR-RIFT,NRPaper} eschews these simplifications, performing
+The RIFT/rapidPE strategy described in \cite{gwastro-PE-AlternativeArchitectures,2017CQGra..34n4002O,gwastro-PENR-RIFT,NRPaper} eschews these simplifications, performing
 embarrassingly-parallel inferences even for costly models.   However, the method as previously implemented still had a
@@
 %
-In this paper we describe a variant of the RIFT algorithm which  translates its low-level
-algorithms to operate on graphics processing units (GPUs), and adjusts its workflow to exploit the speed improvements
+Previously, B. Miller  developed a custom-coded  implementation of the
+relevant likelihood on graphics processing units (GPUs), suggesting substantial performance improvements were possible \cite{Thesis-Brandon-2016,GPU-NU-work}.  
+In this paper,  we describe a variant of the RIFT approach which flexibly translates one of its
+algorithms to operate on GPUs, and adjusts its workflow to exploit the speed improvements
 this re-implementation affords.
@@
 graphics card, then carry out all calculations necessary to implement Eqs. (\ref{eq:lnL:MatrixForm}, \ref{eq:lnL:tmarg})
-on the GPU.  To enable this implementation with portable code, we use \texttt{cupy}, a drop-in-replacement for equivalent
+on the GPU.  To enable this implementation with portable code, we use \texttt{cupy} \cite{cupy_learningsys2017}, a drop-in-replacement for equivalent
 \texttt{numpy} code used for the CPU-based version of ILE.    For the most costly part of the calculation -- the inner
@@
 We generate  synthetic data for a two-detector LIGO configuration, assuming operation at  initial (BBH) or advanced (BNS) design  sensitivity.
-Both signals are generated with SEOBNRv4, with a zero-noise data realization.   
+Both signals are generated with SEOBNRv4 effective-one-body waveform approximation \cite{2017PhRvD..95d4028B}, with a zero-noise data realization.   
 To qualitatively reproduce the noise power spectrum and amplitude of typical binary black hole observations in O1 and
@@
  We perform a multi-stage iterative RIFT analysis of these signal with the
-time-domain SEOBNRv4 (BBH) or TaylorT4 (BNS) approximation, under the simplifying assumption that both objects are point particles with zero
+time-domain SEOBNRv4 (BBH) or TaylorT4 \cite{gw-astro-PN-Comparison-AlessandraSathya2009}( BNS) approximation, under the simplifying assumption that both objects are point particles with zero
 spin.   For both sources, we adopt the fiducial distance prior $p(D_L) \propto D_L^2$,   selecting  a maximum distance
@@
 % ~/parse_report.sh profile_nogpu_pcdev13.log | more
-CPU & 16384 & $\pm 2 $ & 20 & &&532 & &  700  \\ 
+CPU & 16384 & $\pm 2 $ & 20 & 2.4 &&540 & 20 &  700  \\ 
        & 4096 & $\pm 2 $ &   20 \\ \hline
 % ./parse_report.sh 20190130-profile_nogpu_HM_pcdev13.log  | more
+% /parse_report.sh ./profile_nogpu_HM_lowres_pcdev13.log 
 %    setup time: PrecomputeLikelihoodTerms, includes waveform generation. 
@@
 %        condor_userlog iteration_0_ile/logs/*.log
-GPU (a) & 16384 & $(2,\pm 2) $  & 20  & & && & 270 \\
+GPU (a) & 16384 & $\pm 2 $  & 20  & & && & 270 \\
             & 4096 &$\pm 2 $  &  20 &  & & && 45 \\ \hline
@@
        & 4096 & $\pm 2$  & 20 & $1.2 $ &  1  & \editremark{0.75} & \editremark{6}  & 25\\ \hline
+% ./parse_report.sh 20190130-profile_gpu_HM_16kHz_pcdev13.log | more
+% ./parse_report.sh 20190130-profile_gpu_HM_lowres_pcdev13.log  | more
 % ~/parse_report.sh profile_gpu_HM_lowres_pcdev13.log | more
@@
 GPU (b) & 16384 & $\pm 2, \pm 1$ & 20 & 1 && 4.2 & \editremark{1}  & 38  \\
-       & 4096 & $\pm 2, \pm 1$ & 20 & 1&& 2.5  && 35 & \\ \hline
+       & 4096 & $\pm 2, \pm 1$ & 20 & 1&& 2.5  & \editremark{1} & 35 & \\ \hline
 %
@@
 %  ./parse_report.sh 20190130-profile_gpu_LHOworker_16kHz.log | more
-GPU (c) & 16384 & $\pm 2 $  & 20  & & & 18& &160 &  \\
-            & 4096 &$\pm 2 $  &  20 &  & & 22  & & 140 & $\simeq 50$ \\
+% ./parse_report.sh 20190130-profile_gpu_LHOworker_lowres.log | more
+%   * worker iteration cose: (3027 - 687)/20/3851/10^4 * 10^6 // N
+GPU (c) & 16384 & $\pm 2 $  & 20  &6  & & 18& 3 &160 &  \\
+            & 4096 &$\pm 2 $  &  20 & 3.7 & & 22  & 3 & 140 & $\simeq 50$ \\
 \end{tabular}
@@
   hardware, for a two-mode system $(l,m)=\pm 2$ analyzing $T=8\unit{s}$ of data with a massive binary black hole
-  $m_1=35 M_\odot,M_2=30 M_\odot$.  The last column indicates peak GPU utilization.}
+  $m_1=35 M_\odot,M_2=30 M_\odot$.  The last column indicates peak GPU utilization.
+\editremark{Comment: will update last number based on bulk profiling of a whole run; other numbers are from deep dive into individual run}
+}
 \end{table*}
@@
  [ \tau_{setup}  + N_{it}\tau_{it}
- ] 
+ ]  \\
+% (20 \[Tau]s1 + \[Tau]ad 1 2*10^6/10^4)/   20 + (\[Tau]start*1 + 2*10^6 \[Tau]it1 *10^(-6)) // Expand
+& \simeq (\tau_{\rm start}/20\unit{s}) +  10 (\tau_{ad}/\mu s) 
+ \nonumber \\ &
++ 2  [(\tau_{it,liike}/\mu s) + \tau_{it,rest}/\mu s]
 \end{align}
-Typically we target $N_{\rm eval} \simeq 20$ so that startup and other one-time costs  are not a significant contributor to the overall
+where in the latter expression we substitute the usual values of $N_{it} \simeq 2\times 10^6$,  $n_{\rm chunk}\simeq
+10^4$, and $N_{\rm eval}\simeq 20$ conservtively adopted
+for problems involving many degrees of freedom $d$ (between 6 to 8).  We target $N_{\rm eval} \simeq 20$ so that startup and other one-time costs  are not a significant contributor to the overall
 runtime.   In this configuration, the run time per marginalized likelihood is around $T_{ILE}/N_{\rm eval} \simeq 25
@@
 marginally more time than the corresponding $4\unit{kHz}$ analysis.
+%
+Conversely, the overhead of  performing the Monte Carlo (including both $\tau_{ad}$ and $\tau_{it,rest}$) is a quite substantial contribution to the overall runtime for the best-available hardware.
 
@@
        & 4096 & $\pm 2, \pm 1$ & 35 & &  &   &&  &  \\ \hline
-GPU (a) & 16384 & $\pm 2, \pm 1$   & 35 &  &&&&   \\
-       & 4096 & $ \pm 2, \pm 1 $        &  35  & &&&& 360 \\ \hline
+GPU (a) & 16384 & $\pm 2, \pm 1$   & 35 &  &&&  &   \\
+       & 4096 & $ \pm 2, \pm 1 $        &  35  & &&&& 60--450  \\ \hline
+% ./parse_report.sh 20190205-profile_bns_gpu_lowres_pcdev13_iteration3.log | more
 %   - resampling time: take (mcsampler - likelihood), divide by 20=Neval, divide by number of adaptive steps (100)
@@
        & 4096 & $ \pm 2, \pm 1 $        &  35  & 11.5 & 0.1 &  2.4  & &24 &  \\ \hline
-GPU (c) & 16384 & $\pm 2, \pm 1$   & 35 &  &&&& 105  \\
-       & 4096 & $ \pm 2, \pm 1 $        &  35  & 28 &&&& 55 \\ \hline
+% Runtime: Based on *production* data for the LHO result, with ADVANCED ITERATIONS to insure quality
+%   demos/test_workflow_batch_bns_gpu_lowlatency_LHO_v0 : about 2.5 hours for 20 points
+% Profiling: actual individual run,
+%    - it, rest: based on mcsampler time - likelihood, divide by 20, divide by 10^4
+GPU (c) & 16384 & $\pm 2, \pm 1$   & 35 & 71 &&20 & \editremark{38} &  105   \\
+       & 4096 & $ \pm 2, \pm 1 $        &  35  & 28 && && 60  \\ \hline
 \end{tabular}
@@
   $m_1=1.4 M_\odot,M_2=1.35 M_\odot$, with convergence threshold $n_{\rm eff} > 50$.
+\editremark{Comment: will update net runtime number based on bulk profiling of a whole run; other numbers are from deep dive into individual run}
 }
@@
 which has signal SNR=32.  In this analysis, we have adopted the TaylorT4 model with $\ell=2$ and all $m=\pm 2,\pm 1$ modes as our
-fiducial analysis template.   Despite using more modes and longer signals, and despite an overall likelihood evaluation
-cost $\tau_{it}$ consistent BBH analyses,  likelihood evaluations for BNS are often
-\emph{as fast or faster} than the corresponding likelihood evaluations for BBH with otherwise comparable settings.
+fiducial analysis template.   This source has significantly longer duration and higher amplitude, both of which contribute to slower convergence and longer runtime. 
+%% Despite using more modes and longer signals, and despite an overall likelihood evaluation
+%% cost $\tau_{it}$ consistent BBH analyses,  likelihood evaluations for BNS are often
+%% \emph{as fast or faster} than the corresponding likelihood evaluations for BBH with otherwise comparable settings.
 
@@
 
-Improved performance arises instead from re-using our adaptive integrator, which (after the first marginalized
+Improved performance arises in part from re-using our adaptive integrator, which (after the first marginalized
 likelihood evaluation) can exploit tight localization afforded by the many BNS cycles available in band and the high
-amplitude of our fiducial BNS signal.  In other
-words, the Monte Carlo integral converges to our target accuracy in of order $10\times $ fewer steps than for BBH.  
-Because this improved performance relies on the sampler, not on our choice of hardware, we achieve a similar substantial
-improvement for all choices of GPU.
+amplitude of our fiducial BNS signal.   In some cases, the Monte Carlo integral converges to our target accuracy in of order $10\times $ fewer steps than for BBH.  
+Additionally, particularly for first-stage RIFT grids, the improved performance arises from our choice of initial grid: many points fit poorly and identified as such well before the fiducial $2\times 10^6$ Monte Carlo iterations.
+% This latter benefit can occur more frequently as we move to higher dimensions.
+Both sources of improved performance relative to the BBH analysis are independent of choice of hardware.
 
@@
 saturate this bound, even for 16 kHz data, unless we investigate significantly lower starting frequencies or
-vastly more modes.
+employ vastly more angular modes $h_{lm}$.
 %\editremark{some issue: memory usage is about 1Gb for 16 kHz,  close to $4Gb$ limit on commodity cards now.  Will be hard to
@@
 % 
-In our second test we compare two workflows, one with the original embarassingly-parallel RIFT code using $N_{\rm
+In our second test we compare two workflows, one with the original embarrassingly-parallel RIFT code using $N_{\rm
   eval}=1$ and one with the
@@
 The initial coarse grid covers a region $\mc_z\in[1.1962,1.1970] M_\odot$ and $\delta \in [0,0.25]$,  insuring the posterior was smaller than our initial
-coarse grid spacing.     The top  panels of Figure  \ref{fig:BNS:MultiIterate} show posterior disributions derived
+coarse grid spacing.     The top  panels of Figure  \ref{fig:BNS:MultiIterate} show posterior distributions derived
 from the first several RIFT iterations, while the bottom panels show convergence diagnostics.
@@
 per $\lambda$, using 100 cores in the first iteration and 20 cores in each
-subsequent evalution.   
+subsequent evaluation.   
 Note that  because   $N_{\rm eval}=20$ for the GPU is larger than the (hardware-and $f_{\rm sample}$ dependent)  speedup factor between the CPU and
@@
 iterations use 500 random draws from the estimated posterior samples produced from the previous iteration.  
-Because  our fiducial post-Newtonian model (TaylorT4) as implemented in \texttt{lalsuite} does not include spin, we
+Because  our fiducial post-Newtonian model (TaylorT4) as implemented in \texttt{lalsuite}\cite{lalsuite} does not include spin, we
 employ both the \texttt{SEOBNRv4\_ROM} and \texttt{SpinTaylorT4} waveform approximations for nonprecessing binaries,
@@
  5$ times as long as the first iteration of the zero-spin BNS analysis described above.  Each marginalized likelihood
- evaluation in the first requires roughly 60 seconds for both SpinTaylorT4 and SEOBNRv4ROM, on average.   Subsequent
+ evaluation in the first iteration (as well as subsequent iterations) requires roughly 60 seconds for both SpinTaylorT4 and SEOBNRv4ROM, on average.   Subsequent
 marginalized likelihood iterations are not resource-limited and complete in roughly 35 minutes, depending on hardware.
+As previously, we can achieve comparable or smaller turnaround time for an otherwise identical analysis by adjusting $N_{\rm eval}$ appropriate to the available hardware.
 
@@
 % COMMAND
-% plot_posterior_corner.py  --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc  --parameter chi_eff --mc-range  '[1.19655,1.197]' --chi-max 0.025 --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
+% plot_posterior_corner.py  --posterior-file posterior_samples-5.dat  --posterior-file posterior_samples-4.dat   --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc  --parameter chi_eff --mc-range  '[1.19655,1.197]' --chi-max 0.025 --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None  --posterior-label 5 --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend --posterior-linestyle '-' --posterior-linestyle '-' --posterior-linestyle '-' --posterior-linestyle '-' --posterior-linestyle ':'  --posterior-color 'k' --posterior-color 'r' --posterior-color 'g' --posterior-color 'b' --posterior-color 'k' --lnL-cut 50
 \includegraphics[width=0.45\textwidth]{figures/bns_withspin_corner_mc_delta_xi.png}
@@
   show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
-  value of the underlying marginalized likelihood evaluations.  The solid curves show an analysis with SEOBNRv4\_ROM;
-  the dotted curves show an analysis with SpinTaylorT4.  \editremark{currenly only v4ROM shown}  \emph{Left panel } Posterior distribution
+  value of the underlying marginalized likelihood evaluations.
+Only evaluations with marginalized log-likelihoods within $30$ of the maximum are shown, to increase contrast.
+The solid curves show an analysis with SEOBNRv4\_ROM.
+%  the dotted curves show an analysis with SpinTaylorT4.  \editremark{currenly only v4ROM shown} 
+ \emph{Left panel } Posterior distribution
   over  $\mc$,
@@
 
-By allowing  models with higher-order modes to be used in low latency, our code can better-constrain the properties of low-mass
-binaries with suitable amplitudes and orientations, when circumstances allow.  These tighter constraints could
+By allowing  models with higher-order modes to be used in low latency, our code can exploit the tighter constraints which higher modes can enable on  the properties of low-mass
+binaries with suitable amplitudes and orientations.  These tighter constraints could
 better  inform low-latency source classification and hence multimessenger followup observations of compact binary mergers.
@@
 evidence could be used as (the last stage in a hierarchical pipeline for) a detection statistic
-\editremark{citations:old idea,see Thrane/Rory papers recently} \cite{2018PhRvX...8b1019S}.    This approach can identify individual events and
+%\editremark{citations:old idea,see Thrane/Rory papers recently} 
+\cite{2018PhRvX...8b1019S}.    
+This approach can identify individual events and
 even a population.
 Alternatively and in many ways equivalently, one can identify a population of GW sources without assuming any one is
-real, by  applying parameer inference to more candidate events and self-consistently seperating foreground and
+real, by  applying parameter inference to more candidate events and self-consistently separating foreground and
 background. 
@@
 while transitioning smoothly between GPU and CPU mode.     The techniques we used here will be transported to other
-Bayesian inference modeling codes used to interpret GW observations.
+Bayesian inference modeling codes used to interpret GW observations \cite{gwastro-PopulationReconstruct-Parametric-Wysocki2018,gwastro-PopulationReconstruct-Code-PopModels}.
 % Will swap version of code

### 2019-02-08 paper/paper.tex
paper.tex: add more good ideas to conclusion:

@@
 
+When suitable surrogate models are available, the overall code performance and thus latency could be yet again further
+reduced by eliminating the iteration and fitting stages entirely, performing one Monte Carlo at once
+\cite{gwastro-PE-AlternativeArchitecturesROM}.  This approach exploits a linear representation of  $h_{lm}(t)$  via
+basis functions, to enable rapid likelihood evaluation as a function of both extrinsic parameters and  $\lambda$.
+Though not necessarily or compactly available for all surrogate models, particularly for the small basis sizes necessary
+to fit onboard GPUs, this approach could enable exceedingly low latency at small computational cost.  This alternative
+architecture would be exceptionally well-suited to the alternative applications of low-latency PE described above.
+
 The use of \texttt{cupy} enables our code to be highly portable across architectures and heterogeneous GPU environments,

### 2019-02-12 paper/paper.tex
paper.tex: respond to Veitch. Create prototype demo for SEOBNRv3

@@
 % From RIFT paper
-\ILE{}  -- a specific algorithm to ``integrate over extrinsic parameters'' -- provides a straightforward and efficient mechanism to compare any specific candidate gravitational wave source with
+\ILE{}  -- a specific algorithm to ``Integrate (the Likeilhood) over Extrinsic parameters'' -- provides a straightforward and efficient mechanism to compare any specific candidate gravitational wave source with
 real or synthetic data   \cite{gwastro-PE-AlternativeArchitectures,NRPaper,2017PhRvD..96j4041L,2017CQGra..34n4002O},
@@
 orientation of the binary relative to the earth.  
-Specifically the likelihood of the data given gaussian noise has the form  (up to normalization)
+Specifically the likelihood of the data given gaussian noise, relative to gaussian noise, has the form  (up to normalization)
 \begin{equation}
 \label{eq:lnL}
-\ln {\cal L}(\bm{\lambda} ;\theta )=-\frac{1}{2}\sum\limits_{k}\langle h_{k}(\bm{\lambda} ,\theta )-d_{k} |h_{k}(\bm{\lambda} ,\theta )-d_{k}\rangle _{k}-\langle d_{k}|d_{k}\rangle _{k},
+\ln {\cal L}(\bm{\lambda} ,\theta )=-\frac{1}{2}\sum\limits_{k}\langle h_{k}(\bm{\lambda} ,\theta )-d_{k} |h_{k}(\bm{\lambda} ,\theta )-d_{k}\rangle _{k}-\langle d_{k}|d_{k}\rangle _{k},
 \end{equation}
@@
 \end{equation}
-via direct Monte Carlo integration, where $p(\theta)$ is uniform in 4-volume and source orientation.  
-The integral in time is performed at a low level, by direct quadrature. 
+via direct Monte Carlo integration over almost all parameters $\theta$, where $p(\theta)$ is uniform in 4-volume and source orientation.  
+For the event time parameter, we marginalize by direct quadrature, for each choice of the remaining Monte Carlo parameters.
 For the remaining dimensions, to evaluate the likelihood in regions of high importance, we use an adaptive Monte Carlo as described in
@@
 for a plane wave propagating along $\hat{n}$ \cite{gwastro-PE-AlternativeArchitectures}.
-We then substitute these expressions for $h_k$ into the likelihood function~\eqref{eq:loglikelihood}
+We then substitute these expressions for $h_k$ into the likelihood function~\eqref{eq:lnL}
 thereby generating~\cite{gwastro-PE-AlternativeArchitectures}
@@
 % COMMAND
-%  plot_posterior_corner.py --posterior-file posterior_samples-4.dat  --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
-%plot_posterior_corner.py --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --posterior-label 2 --posterior-label 1 --use-legend
+% ../plot_last_iterations_with.sh --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --use-legend --posterior-color 'b' --posterior-color 'k' --posterior-linestyle ':' --posterior-linestyle ':'
+% ../plot_last_iterations_with.sh --parameter mc --parameter delta_mc   --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --use-legend --posterior-color 'b' --posterior-color 'k' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':'
+%
 \includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_m2.png}
@@
 \caption{\label{fig:BBH:MultiIterate}\textbf{Convergence of BBH analysis: Zero spin}: Results for marginal posterior distributions
-  of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
+  of our fiducial synthetic binary black hole.  Solid contours show credible intervals; solid one-dimensional distributions
   show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
@@
 \SkipForEarlyCirculation{
+\subsection{Precessing binary black hole analysis}
+\label{sec:sub:BBH_v3}
+
+In this section, we reanalyze our fiducial (zero spin) binary black hole signal without any restrictions on spin
+magnitude or alignment, using a precessing effective--one-body approximation (SEOBNRv3 \editremark{cite}).   
+We start with an initial  \emph{three-dimensional} grid of 5000 points in $\mc \in [20,35] M_\odot$, $\eta
+\in[0.2,0.25]$ and $\chi_{\rm eff} \in [-0.5,0.5]$.  We very conservatively perform 10 iterations, adding 500 points
+candidate evaluation points each iteration, using a volumetric spin prior on both dimensionless spins.
+
+We break our  pipeline into two stages.  In the first five
+iterations, the likelihood is assumed to depend only on  $\mc,\eta,\chi_{\rm eff}$.   By construction, during these
+stages our algorithm proposes random choices for the perpendicular spins $\chi_{i,\perp}$  and the remaining
+antisymmetric degree of freedom for the two candidate spins.    As demonstrated in Figure \ref{fig:BBH:Spin}, well
+before the first five iterations complete, we've thoroughly explored the $\mc,\eta,\chi_{\rm eff}$ distribution and
+converged to a (biased) posterior.
+
+In the second stage, we perform a gaussian process fit which includes  all binary parameters ($\mc,\eta,\chi_{\rm
+  eff},\chi_-,\chi_{i,x},\chi_{i,y}$).  As demonstrated in Figure \ref{fig:BBH:Spin}, we quickly converge to a posterior
+distribution.
+
+Throughout this calculation, each individual marginalized likelihood required roughly 1 minute to evaluate.  With 100
+available GPU+cores, this analysis could require as little as 1.7 hours to evaluate the marginalized likelihood for all
+10 iterations.  In practice, by using $N_{\rm eval}=20$ and therefore restricting to 25 cores for the iterative stages, we insured each iteration after the first required no less than
+20 minutes to complete.  As a result of this self-imposed restriction, the wallclock time necessary to complete all 10 iterations was several hours.
+That said, in this particular case substantially fewer iterations were required to reach convergence.
+
+\begin{figure*}
+% COMMAND
+% ../plot_last_iterations_with.sh --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --use-legend --posterior-color 'b' --posterior-color 'k' --posterior-linestyle ':' --posterior-linestyle ':'
+% ../plot_last_iterations_with.sh --parameter mc --parameter delta_mc --parameter chi_eff  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --use-legend --posterior-color 'b' --posterior-color 'k' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':'
+%  ../plot_last_iterations_with.sh --parameter a1z --parameter a2z --parameter chi1_perp --parameter chi2_perp  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --use-legend --posterior-color 'b' --posterior-color 'k' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle '-' --posterior-linestyle '-' --posterior-color 'g' --posterior-color 'magenta'
+
+\includegraphics[width=0.45\textwidth]{figures/bns_withspin_mc_cum.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_meanVar.png}
+\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_converge.png}
+\caption{\label{fig:BBH:Spin}\textbf{Demonstration precessing binary black hole analysis}:
+Results for marginal posterior distributions
+  of our fiducial binary black hole, assuming the binary could have precessing spins.  Solid contours show credible intervals; solid one-dimensional distributions
+  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
+  value of the underlying marginalized likelihood evaluations.
+Only evaluations with marginalized log-likelihoods within $30$ of the maximum are shown, to increase contrast.
+The solid curves show an analysis with SEOBNRv3.
+%  the dotted curves show an analysis with SpinTaylorT4.  \editremark{currenly only v4ROM shown} 
+ \emph{Left panel } Posterior distribution
+  over  $\mc$,
+  $\delta=(m_1-m_2)/M$, and $\chi_{\rm eff}$.    \emph{Right panel}: Marginal 1d CDFs of $\delta$, showing convergence.
+\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
+first iteration the
+candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
+\emph{Bottom right panel}: $Z = \int d\lambda {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
+error budget early on, Monte Carlo error is not shown.
+}
+\end{figure*}
+}
+
+\SkipForEarlyCirculation{
 \section{ Results (optional?)}

### 2019-02-13 paper/paper.tex
paper: updates to profiling using automation rather than horrible humans

@@
 \affiliation{\RIT}
+\author{R. O'Shaughnessy}
+\affiliation{\RIT}
 \author{Yao-Lung L. Fang}
@@
 \affiliation{National Synchrotron Light Source II, Brookhaven National Laboratory, Upton, NY 11973, USA}
-\author{R. O'Shaughnessy}
-\affiliation{\RIT}
 \author{Jacob Lange}  % for all the convergence material we work on ... it comes from Jake's practice
@@
 %overall computational cost and turnaround time of 
-computational cost of parameter estimation \cite{gwastro-pe-Brandon-STF2,gwastro-PE-AlternativeArchitectures,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,2016PhRvD..94d4031S,2017CQGra..34k5006V}.
+computational cost of parameter estimation \cite{gwastro-pe-Brandon-STF2,gwastro-PE-AlternativeArchitectures,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,2016PhRvD..94d4031S,2017CQGra..34k5006V,2018arXiv180608792Z}.
 % The computational cost of parameter estimation can be reduced 
@@
 adopting a sparse representation to reduce the computational cost of data
-handling \cite{antil2013two,gwastro-mergers-PE-ReducedOrder-2013,2016PhRvD..94d4031S,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,gwastro-PE-AlternativeArchitectures}.
+handling \cite{antil2013two,gwastro-mergers-PE-ReducedOrder-2013,2016PhRvD..94d4031S,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,gwastro-PE-AlternativeArchitectures,2010arXiv1007.4820C,2018arXiv180608792Z}.
 % 2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C
@@
 % ~/parse_report.sh profile_nogpu_pcdev13.log | more
-CPU & 16384 & $\pm 2 $ & 20 & 2.4 &&540 & 20 &  700  \\ 
-       & 4096 & $\pm 2 $ &   20 \\ \hline
+CPU & 16384 & $\pm 2 $ & 20 & 2.4 &&540 & 20 &  690  \\ 
+       & 4096 & $\pm 2 $ &   20  &&&& 20 \\ \hline
 % ./parse_report.sh 20190130-profile_nogpu_HM_pcdev13.log  | more
@@
 %    
-CPU & 16384 & $\pm 2,\pm 1 $ & 20 & 1.5 && 680 & &  1000  \\ 
-       & 4096 & $ \pm 2, \pm 1 $ &   20 \\ \hline
+CPU & 16384 & $\pm 2,\pm 1 $ & 20 & 1.5 && 680 & 20 &  1060  \\ 
+       & 4096 & $ \pm 2, \pm 1 $ &   20 &&&& 20  \\ \hline
 % LIGO-WA, or a compute node at CIT with a grpahics catd: 
@@
 GPU (a) & 16384 & $\pm 2 $  & 20  & & && & 270 \\
-            & 4096 &$\pm 2 $  &  20 &  & & && 45 \\ \hline
+            & 4096 &$\pm 2 $  &  20 &  & & & & 45 \\ \hline
 %% % LIGo CIT, pcdev13, 16 kHz : basically indistinguishable from cost at 4kHz
@@
 %   'residual' time: take mcsampler, subtract likleihood, divide by number of calls and number of evals per chunk
-GPU (b) & 16384 & $\pm 2$ & 20  & 1.8 & 1 & \editremark{0.85}& \editremark{6} &28 & 15\\
-       & 4096 & $\pm 2$  & 20 & $1.2 $ &  1  & \editremark{0.75} & \editremark{6}  & 25\\ \hline
+GPU (b) & 16384 & $\pm 2$ & 20  & 1.8 & 1 & 0.85& 20 &28 & 15\\
+       & 4096 & $\pm 2$  & 20 & $1.2 $ &  1  & 0.75 & 20  & 25\\ \hline
 % ./parse_report.sh 20190130-profile_gpu_HM_16kHz_pcdev13.log | more
@@
 %   - extra eval cost: (mcsampler - likelihood)/20/n_eval_likelihood/10^4  
-GPU (b) & 16384 & $\pm 2, \pm 1$ & 20 & 1 && 4.2 & \editremark{1}  & 38  \\
-       & 4096 & $\pm 2, \pm 1$ & 20 & 1&& 2.5  & \editremark{1} & 35 & \\ \hline
+GPU (b) & 16384 & $\pm 2, \pm 1$ & 20 & 1 && 4.2 & 20  & 38  \\
+       & 4096 & $\pm 2, \pm 1$ & 20 & 1&& 2.5  & 20 & 35 & \\ \hline
 %
@@
 %   * worker iteration cose: (3027 - 687)/20/3851/10^4 * 10^6 // N
-GPU (c) & 16384 & $\pm 2 $  & 20  &6  & & 18& 3 &160 &  \\
-            & 4096 &$\pm 2 $  &  20 & 3.7 & & 22  & 3 & 140 & $\simeq 50$ \\
+GPU (c) & 16384 & $\pm 2 $  & 20  &6  & & 18&  58&160 &  \\
+            & 4096 &$\pm 2 $  &  20 & 3.7 & & 11  & 58  & 140 & $\simeq 50$ \\
 \end{tabular}
@@
   $m_1=35 M_\odot,M_2=30 M_\odot$.  The last column indicates peak GPU utilization.
-\editremark{Comment: will update last number based on bulk profiling of a whole run; other numbers are from deep dive into individual run}
+%\editremark{Comment: will update last number based on bulk profiling of a whole run; other numbers are from deep dive into individual run}
 }
@@
 %   (11261 - 10974)/20/100. = 0.14 sec for 16 kHz
-CPU & 16384 & $\pm 2,\pm 1 $ & 35 & 26 & 0.14  & 680  & & 590    \\ 
-       & 4096 & $\pm 2, \pm 1$ & 35 & &  &   &&  &  \\ \hline
+CPU & 16384 & $\pm 2,\pm 1 $ & 35 & 26 & 0.14  & 680  & 25 & 590     \\ 
+       & 4096 & $\pm 2, \pm 1$ & 35 & &  &   &&  25 &  \\ \hline
 GPU (a) & 16384 & $\pm 2, \pm 1$   & 35 &  &&&  &    \\
@@
 %    - it, rest: based on mcsampler time - likelihood, divide by 20, divide by 10^4
-GPU (c) & 16384 & $\pm 2, \pm 1$   & 35 & 71 &&20 & \editremark{38} &  105   \\
-       & 4096 & $ \pm 2, \pm 1 $        &  35  & 28 && && 60  \\ \hline
+GPU (c) & 16384 & $\pm 2, \pm 1$   & 35 & 71 &&20 & 38 &  105   \\
+       & 4096 & $ \pm 2, \pm 1 $        &  35  & 28 && 12 && 60  \\ \hline
 \end{tabular}
@@
   $m_1=1.4 M_\odot,M_2=1.35 M_\odot$, with convergence threshold $n_{\rm eff} > 50$.
-\editremark{Comment: will update net runtime number based on bulk profiling of a whole run; other numbers are from deep dive into individual run}
 }
@@
 localization and  precise source characterization.     This prospect is particularly interesting because RIFT can
-often achieve this performance using computationally costly models for binary merger, as the waveform generation cost
-does not usually limit code performance.
+often achieve this performance using computationally costly models for binary merger with rich physics like higher modes
+or eccentricity, as the waveform generation cost does not usually limit code performance.
 
@@
   eval} \simeq 1$).
+%
 
@@
 real, by  applying parameter inference to more candidate events and self-consistently separating foreground and
-background. 
+background
 \cite{2015PhRvD..91b3005F,2019MNRAS.tmp..230G}.

### 2019-03-19 paper/paper.tex
paper.tex: starting referee response changes

@@
 \newcommand\SkipForEarlyCirculation[1]{}
+\newcommand\AddedResponse[1]{\textbf{#1}}
 %\newcommand\SkipForEarlyCirculation[1]{#1}
@@
 \begin{equation}
+\label{eq:lnLmarg}
  {\cal L}_{\rm marg}\equiv\int  {\cal L}(\bm{\lambda} ,\theta )p(\theta )d\theta
@@
 by  generating the dynamics and outgoing radiation in all possible directions once and for all for fixed
-$\mathbf{\lambda}$, using a spherical harmonic decomposition.  Using this cached information effectively,  the likelihood can be evaluated as a function of $\theta$ at very low computational cost.  
+$\bm{\lambda}$, using a spherical harmonic decomposition.  Using this cached information effectively,  the likelihood can be evaluated as a function of $\theta$ at very low computational cost.  
 % From ROM paper
@@
 \begin{align} \label{eq:strain}
-h(t,\vartheta,\phi;\lambda) =  h_+(t,\vartheta,\phi;\lambda) - 
-                                i h_\times (t,\vartheta,\phi;\lambda) \, ,
+h(t,\vartheta,\phi;\bm{\lambda}) =  h_+(t,\vartheta,\phi;\bm{\lambda}) - 
+                                i h_\times (t,\vartheta,\phi;\bm{\lambda}) \, ,
 \end{align}
@@
 \begin{align} \label{eq:strain_mode}
-h(t,\vartheta,\phi;\lambda) = 
-\sum_{\ell=2}^{\infty} \sum_{m=-\ell}^{\ell} \frac{D_{\rm ref}}{D} h^{\ell m}(t;\lambda) \Y{-2}_{\ell m} \left(\vartheta, \phi \right) \, ,
+h(t,\vartheta,\phi;\bm{\lambda}) = 
+\sum_{\ell=2}^{\infty} \sum_{m=-\ell}^{\ell} \frac{D_{\rm ref}}{D} h^{\ell m}(t;\bm{\lambda}) \Y{-2}_{\ell m} \left(\vartheta, \phi \right) \, ,
 \end{align}
-where the sum includes all available harmonic modes $h^{\ell m}(t;\pmb{\lambda})$ made available by the model;  where
+where the sum includes all available harmonic modes $h^{\ell m}(t;\pmb{\bm{\lambda}})$ made available by the model;  where
 $D_{\rm ref}$ is a fiducial reference distance; and where $D$, the luminosity distance to the  source, is one of the
@@
 \begin{align}
-\ln {\cal L}(\lambda, \theta) 
-&= (D_{\rm ref}/D) \text{Re} \sum_k \sum_{\ell m}(F_k \Y{-2}_{\ell m})^* Q_{k,lm}(\lambda,t_k)\nonumber \\
+\ln {\cal L}(\bm{\lambda}, \theta) 
+&= (D_{\rm ref}/D) \text{Re} \sum_k \sum_{\ell m}(F_k \Y{-2}_{\ell m})^* Q_{k,lm}(\bm{\lambda},t_k)\nonumber \\
 &   -\frac{(D_{\rm ref}/D)^2}{4}\sum_k \sum_{\ell m \ell' m'}
@@
 {
-|F_k|^2 [\Y{-2}_{\ell m}]^*\Y{-2}_{\ell'm'} U_{k,\ell m,\ell' m'}(\lambda)
+|F_k|^2 [\Y{-2}_{\ell m}]^*\Y{-2}_{\ell'm'} U_{k,\ell m,\ell' m'}(\bm{\lambda})
 }
@@
 \begin{align}
-Q_{k,\ell m}(\lambda,t_k) &\equiv \qmstateproduct{h_{\ell m}(\lambda,t_k)}{d}_k \nonumber\\
-&= 2 \int_{|f|>f_{\rm low}} \frac{df}{S_{n,k}(|f|)} e^{2\pi i f t_k} \tilde{h}_{\ell m}^*(\lambda;f) \tilde{d}(f)\ , \\
-{ U_{k,\ell m,\ell' m'}(\lambda)}& = \qmstateproduct{h_{\ell m}}{h_{\ell'm'}}_k\ , \\
-V_{k,\ell m,\ell' m'}(\lambda)& = \qmstateproduct{h_{\ell m}^*}{h_{\ell'm'}}_k  \ .
+Q_{k,\ell m}(\bm{\lambda},t_k) &\equiv \qmstateproduct{h_{\ell m}(\bm{\lambda},t_k)}{d}_k \nonumber\\
+&= 2 \int_{|f|>f_{\rm low}} \frac{df}{S_{n,k}(|f|)} e^{2\pi i f t_k} \tilde{h}_{\ell m}^*(\bm{\lambda};f) \tilde{d}(f)\ , \\
+{ U_{k,\ell m,\ell' m'}(\bm{\lambda})}& = \qmstateproduct{h_{\ell m}}{h_{\ell'm'}}_k\ , \\
+V_{k,\ell m,\ell' m'}(\bm{\lambda})& = \qmstateproduct{h_{\ell m}^*}{h_{\ell'm'}}_k  \ .
 \end{align}
 \end{subequations}
-Rewriting sums as matrix operations, the likelihood can be equivalently expressed as
-\begin{align}
-\ln {\cal L} &= \frac{D_{\rm ref}}{D} \text{Re}[ (F Y)^\dag Q]  \nonumber \\
- & - \frac{D_{\rm ref}^2}{4 D^2} [ (FY)^\dag U FY + (FY)^TV FY]
-\label{eq:lnL:MatrixForm}
-\end{align}
-where $F,D$ are arrays over extrinsic parameters; $Q$ is an array over time and $(l,m)$; and $U,V$ are matrices over
-$(l,m)$.
 
-\subsection{Accelerated evaluation via efficient multiplication}
-%* GPU code
-%* Refactoring code to use the efficient GPU loops
+\AddedResponse{In RIFT, the marginalization in Eq. (\ref{eq:lnLmarg}) over extrinsic parameters is performed by evaluating this likelihood $\ln {\cal
+    L}(\bm{\lambda},\theta)$ on long
+  arrays $\theta_\alpha$ of extrinsic parameters, including $D_\alpha$ but excluding time.    Treating the block of quantities that arise
+  in one sequence of evaluations together,  the expressions $Q,F,U,V$ and $D$ can be considered as multi-dimensional
+  arrays.  For example, $F_k(\theta_\alpha)$ is a matrix indexed over $k$ (detectors) and $\alpha$
+  (extrinsic parameters); $\Y{-2}_{lm}(\theta_k)$ is a matrix of shape (modes)$\times$(extrinsic); and $D_\alpha$ is a one-dimensional array over $\alpha$.  By contrast, the  $U,V$ are arrays of shape
+  (detectors)$\times$ (modes)$^2$ are independent of extrinsic parameters.  While $Q$ depends explicitly on time, and
+  varies rapidly on short timescales, $F$ varies on the earth's rotation timescale, so we  approximate $F$ as  time-independent. }
 
-At a fixed set of intrinsic parameters $\lambda$, \ILE{} will repeatedly evaluate the time-marginalized likelihood 
+At a fixed set of intrinsic parameters $\bm{\lambda}$, \ILE{} will repeatedly evaluate the time-marginalized likelihood 
 for each candidate Monte Carlo set of extrinsic parameters $\theta$:
@@
 where $T$ is a small coincidence window consistent with the candidate event time; here and previously,
-$T=0.3\unit{sec}$.   To accelerate the code, after precomputing the inner products $U,V,Q$, we simply shift them to the
+$T=0.3\unit{sec}$.  
+%
+\AddedResponse{
+Because we treat the  parameter  asymmetrically when marginalizing over all extrinsic parameters, we likewise organize
+a multidimensional array representation of $Q$ to emphasize the special role of time: for each $k,l,m$ we construct a uniformly-sampled timeseries
+$Q_{k,lm}(\bm{\lambda},\tau)$ versus $\tau$, truncating it to a narrow coincidence window.   In effect, we represent $Q$
+by a matrix 
+  $Q_{k,lm}(\bm{\lambda},t+\hat{n}\cdot x_k)$ with  shape (detectors)$\times$(modes)$\times$(extrinsic)$\times$(time).
+}
+%
+\AddedResponse{In terms of these multidimensional arrays, r}ewriting sums as matrix operations, the likelihood can be equivalently expressed as
+\begin{align}
+\ln {\cal L} &= \frac{D_{\rm ref}}{D} \text{Re}[ (F Y)^\dag Q]  \nonumber \\
+ & - \frac{D_{\rm ref}^2}{4 D^2} [ (FY)^\dag U FY + (FY)^TV FY]
+\label{eq:lnL:MatrixForm}
+\end{align}
+\AddedResponse{where this symbolic expression employs an implicit index-summation convention such that all naturally paired
+  indices are contracted.  The result is an array of shape (time)$\times$(extrinsic).}
+%where $F,D$ are arrays over extrinsic parameters; $Q$ is an array over time and $(l,m)$; and $U,V$ are matrices over
+%$(l,m)$.
+
+\subsection{Accelerated evaluation via efficient multiplication}
+%* GPU code
+%* Refactoring code to use the efficient GPU loops
+
+%% At a fixed set of intrinsic parameters $\bm{\lambda}$, \ILE{} will repeatedly evaluate the time-marginalized likelihood 
+%% for each candidate Monte Carlo set of extrinsic parameters $\theta$:
+%% \begin{eqnarray}
+%% {\cal L}_{\rm margT} \equiv  \int {\cal L} \frac{dt}{T}
+%% \label{eq:lnL:tmarg}
+%% \end{eqnarray}
+%% where $T$ is a small coincidence window consistent with the candidate event time; here and previously,
+%% $T=0.3\unit{sec}$. 
+  To accelerate the code, after precomputing the inner products $U,V,Q$, we simply shift them to the
 graphics card, then carry out all calculations necessary to implement Eqs. (\ref{eq:lnL:MatrixForm}, \ref{eq:lnL:tmarg})
-on the GPU.  To enable this implementation with portable code, we use \texttt{cupy} \cite{cupy_learningsys2017}, a drop-in-replacement for equivalent
+on the GPU.  
+\AddedResponse{These arrays are only a few kilobytes.  We then construct blocks of $10^4$ random extrinsic parameters $\theta_\alpha$
+  with the CPU; transfer them to the board; and use on-board code to construct $10^4$ values for $\lnLmarg$. }
+To enable this implementation with portable code, we use \texttt{cupy} \cite{cupy_learningsys2017}, a drop-in-replacement for equivalent
 \texttt{numpy} code used for the CPU-based version of ILE.    For the most costly part of the calculation -- the inner
@@
 
-We do benefit slightly by re-using the adaptive Monte Carlo integrator for each extrinsic point $\lambda$.  Since the
+We do benefit slightly by re-using the adaptive Monte Carlo integrator for each extrinsic point $\bm{\lambda}$.  Since the
 integrator has already identified the likely range of sky locations and distances on the first iteration, each
@@
 $4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.  In this section, we  conservatively report computational cost for
-parameters $\lambda$ which are close to or within the posterior.  Because of their consistency with the data, 
+parameters $\bm{\lambda}$ which are close to or within the posterior.  Because of their consistency with the data, 
 the marginalized likelihood calculations converge the most slowly, as they have the best-determined extrinsic
@@
 \subsection{Massive compact binary black holes}
-Previously, each instance of ILE examined one intrinsic point $\lambda$.  The overall cost of this ILE evaluation
+Previously, each instance of ILE examined one intrinsic point $\bm{\lambda}$.  The overall cost of this ILE evaluation
 involved two parts: a startup cost, a setup cost, and a Monte Carlo integration cost.  The  startup cost $\tau_{start}$
@@
  $N_{ad} \simeq N_{it}/n_{chunk}$), regenerating a sampling prior in two sky location parameters and distance.  
-The choice of one $\lambda$ for each instance of ILE was due to the substantial time  $\tau_{MC}$, which vastly
+The choice of one $\bm{\lambda}$ for each instance of ILE was due to the substantial time  $\tau_{MC}$, which vastly
 dominated the overall computational cost.   For example, for a typical analysis of a short signal --  a typical binary
@@
 instance of ILE loops over
-$N_{eval}$ different choices for $\lambda$. 
+$N_{eval}$ different choices for $\bm{\lambda}$. 
 %
@@
 
-The \ILE{} likelihood  generates waveform dynamics and $h_{lm}(t)$ once per evaluation $\lambda$.   Waveform generation can contribute significantly to the
+The \ILE{} likelihood  generates waveform dynamics and $h_{lm}(t)$ once per evaluation $\bm{\lambda}$.   Waveform generation can contribute significantly to the
 time needed to evaluate  $\ln {\cal L}_{\rm marg}$ for extremely costly
@@
 To better illustrate code performance in realistic settings, we also describe end-to-end analyses with
-the original and modified code.     In a full analysis, many of the initially-proposed  evaluation points $\lambda$ are
+the original and modified code.     In a full analysis, many of the initially-proposed  evaluation points $\bm{\lambda}$ are
 either not or marginally consistent with the data.  As a result, most marginalized likelihood evaluations in the first
@@
   of our fiducial synthetic binary black hole.  Solid contours show credible intervals; solid one-dimensional distributions
-  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
+  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
   value of the underlying marginalized likelihood evaluations.   
@@
 candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
-\emph{Bottom right panel}: The estimated evidence $Z = \int d\lambda {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
+\emph{Bottom right panel}: The estimated evidence $Z = \int d\bm{\lambda} {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
 error budget, Monte Carlo error is not shown.
@@
 final posterior is already well-explored with the initial gridpoints, and converged by the second iteration.  
-Conversely, on average all iterations required roughly 45 seconds per $\lambda$ to evaluate their grid on GPU (a) hardware.  This
+Conversely, on average all iterations required roughly 45 seconds per $\bm{\lambda}$ to evaluate their grid on GPU (a) hardware.  This
 configuration therefore converged within the 30 minutes needed for the first two iterations.
@@
   of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
-  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
+  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
   value of the underlying marginalized likelihood evaluations.  
@@
 candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
-\emph{Bottom right panel}: $Z = \int d\lambda {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
+\emph{Bottom right panel}: $Z = \int d\bm{\lambda} {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
 error budget early on, Monte Carlo error is not shown.
@@
 first iteration, followed by one core+GPU for remaining iterations.
-By contrast, with the original CPU-only RIFT code analyzing each $\lambda$ in parallel, this process requires roughly 20 core-minutes
-per $\lambda$, using 100 cores in the first iteration and 20 cores in each
+By contrast, with the original CPU-only RIFT code analyzing each $\bm{\lambda}$ in parallel, this process requires roughly 20 core-minutes
+per $\bm{\lambda}$, using 100 cores in the first iteration and 20 cores in each
 subsequent evaluation.   
@@
   of our fiducial synthetic neutron star, assuming the binary has nonprecessing spins.  Solid contours show credible intervals; solid one-dimensional distributions
-  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
+  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
   value of the underlying marginalized likelihood evaluations.
@@
   of our fiducial binary black hole, assuming the binary could have precessing spins.  Solid contours show credible intervals; solid one-dimensional distributions
-  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\lambda$ and
+  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
   value of the underlying marginalized likelihood evaluations.
@@
 \label{sec:conclude}
-We have demonstrated that the marginalized likelihood $\ln {\cal L}_{\rm marg}(\lambda)$ appearing in the RIFT/rapidPE parameter
-inference calculation can be evaluated at fixed $\lambda$ in tens of seconds on average for
+We have demonstrated that the marginalized likelihood $\ln {\cal L}_{\rm marg}(\bm{\lambda})$ appearing in the RIFT/rapidPE parameter
+inference calculation can be evaluated at fixed $\bm{\lambda}$ in tens of seconds on average for
 both binary black holes and binary neutron stars.  This performance improvement could enable very low latency source
@@
 the board.  This configuration should further reduce the average time needed to compute $\ln {\cal L}_{\rm
-  marg}(\lambda)$ for both binary black holes (which are Monte Carlo limited) and binary neutron stars (which are
+  marg}(\bm{\lambda})$ for both binary black holes (which are Monte Carlo limited) and binary neutron stars (which are
 inner-product limited).   We anticipate a further factor of roughly 10 reduction in overall evaluation time can be
@@
 \cite{gwastro-PE-AlternativeArchitecturesROM}.  This approach exploits a linear representation of  $h_{lm}(t)$  via
-basis functions, to enable rapid likelihood evaluation as a function of both extrinsic parameters and  $\lambda$.
+basis functions, to enable rapid likelihood evaluation as a function of both extrinsic parameters and  $\bm{\lambda}$.
 Though not necessarily or compactly available for all surrogate models, particularly for the small basis sizes necessary

### 2019-03-19 paper/referee/response-v0.tex
paper: referee. Draft response

\documentclass[onecolumn]{revtex4}
\usepackage{hyperref}
\usepackage{color}
\usepackage{xspace}
\usepackage{bm}
\newcommand\editremark[1]{{\color{red} #1}}

\newcommand{\IMRPD}{\textsc{IMRPhenomD}\xspace}
\newcommand{\IMRPDT}{\textsc{IMRPhenomD\_NRTidal}\xspace}
\newcommand{\IMRP}{\textsc{IMRPhenomPv2}\xspace}
\newcommand{\SEOBP}{\textsc{SEOBNRv3}\xspace}
\newcommand{\SEOBA}{\textsc{SEOBNRv4}\xspace}
\newcommand{\SEOBAROM}{\textsc{SEOBNRv4\_ROM}\xspace}
\newcommand{\NRSur}{NRSur7dq2\xspace}
\newcommand{\TEOB}{SEOBNRv4T\xspace}
\newcommand{\Resum}{TEOBResumS\xspace}
\newcommand\RIFT{RIFT}

\begin{document}

We appreciate the referee's thoughtful suggestions, which will improve the paper.

\begin{enumerate}
\item * The introduction should be expanded to (briefly) discuss
hardware-based acceleration techniques in GW data analysis. For
example, there are significant efforts on the search side to use GPUs.
Within parameter estimation communities, much effort has gone into
optimizing LALInference including the use of multi-core environments
(using OpenMP, for example).

\noindent \textbf{Response}:

\item * Due to the extra $<d | d >$ normalization factor in Eq 1, Eqs 1 and 7
look inconsistent to me.

\noindent \textbf{Response}:  Eq. (1) and (7) from Pankow et al are consistent.  The extra factor of $<d|d>$ eliminates the term in the likelihood which is independent of
the proposed signal $h(\lambda,\theta)$.

\item * Please write an equation showing how Q, U, V, F, and D are related
to quantities appearing in Eq 7 and Eq 8. Based on the text alone I
wasn't able to figure out the corresponding formula. Presumably, the
authors did not include this in the current draft due to the
expression being potentially messy. If this is the case, then perhaps
at least one of these matrices could be written down explicitly from
which the others could be worked out by the reader.

\noindent \textbf{Response}: We have significantly expanded the discussion between Eq. (8) and the start of section II.B
to emphasize how $Q,U,V,F,D$ are encoded into multidimensional arrays.

\item * What is the unbolded Lambda appearing on page 2 and how is it
related to the bolded version? Please define this variable in the
paper.

\noindent \textbf{Response}: We have now consistently rendered $\lambda\rightarrow \bm{\lambda}$; we apologize for the
confusion this duplicated symbol introduced.

\item * I did not quite understand what was meant by "are arrays over
extrinsic parameters" (page 2) since it appears that these parameters
are continuous variables (and not yet sampled).

\noindent \textbf{Response}: Our response to \# 3 should clarify this discussion.  We originally muddled the distinction
between evaluating the likelihood in general and on a large array of discrete  Monte Carlo samples

\item * Just to clarify: the GPU accelerated part is the
time-marginalization and not the marginalization over all of the
extrinsic parameters (Eq 4)? If this is true, then maybe state that
explicitly because it's not obvious and, additionally, the reader will
probably not be aware of the fact that this 1-d integral (rhw time
marginalization) is what looks to be the slowest part of rapidPE.

\noindent \textbf{Response}: The GPU-accelerated parts are (a) all trigonometric operations appearing in $Y$ and $F$, (b)  matrix operations
necessary to compute $Q$, including the time-of-flight array shift; and (c) the one-dimensional time marginalization.

\item * One of the main criticisms of GPU computing is the data transfer
latencies to and from the device. There should be some discussion of
what data is being transferred back and forth, at what stage of the
algorithm does this occur, and does it need to be done repeatedly.
Also, is any special care taken to reduce to the number of data
transfers? Or perhaps the problem is so computationally heavy this
isn't really an issue?

\noindent \textbf{Response}: In our response to \# 3, we now clarify the data transferred to the board (the initial
$Q,U,V$, once,  but also the samples $\theta_\alpha$, each iteration).

In a future version of the code, described in the second paragraph of our conclusions, we intend to perform all Monte
Carlo random number generation on the GPU.  In this configuration we will not have significant data transfer on and off
the board.

\item * Relatedly, how much memory is required to store these matrices?
Since GPUs have smaller RAM sizes, is any special care or tuning
required to get the problem to fit onto the device? Later on the
authors use the ell=2 modes as example -- since the size of the
matrices grows with the number of modes, does the code need to account
for this in any way?

\noindent \textbf{Response}: While the memory footprint of our operations loosely grows linearly with the number of
modes involved, we have noticed no significant limitations on our memory footprint.  We have colleagues who use models with modes
$l \le 8$ without reporting memory footprint problems.

The memory footprint of the initially transferred data is dominated by the $Q_{k,lm}$ time arrays is no larger than $16 kHz
\times 0.15  \lesssim 2500$ floating point numbers per mode per detector.  Even with several hundred modes, these are
quite small.
Potentially more problematic are our intermediate data products, notably the matrix of time-shifted arrays $Q$ built
from $Q_{k,lm}(\lambda,\tau)$ .  Our code has been carefully
organized to loop over detectors so the memory footprint of $Q$ is only (modes)$\times$ (extrinsic)$\times$ (time),
which is roughly $10^4$ times the footprint of each individual $Q_{k,lm}(\lambda,\tau)$ timeseries.  So many modes can
fit well within a typical 4Gb GPU memory space, even allowing for substantial overhead.

\item * What is epsilon appearing in 2c? Is it possible to quote typical
values? Is it problem-specific?

\noindent \textbf{Response}:   The factor $\epsilon$ connecting the number of effective samples to the number of Monte
Carlo trials is indeed highly problem dependent, and reflects the convergence rate of the Monte Carlo integral.
Approximately equivalently, the stanard deviation of the Monte Carlo error estimate scales as $1/\sqrt{N_{it}}$, with a
prefactor that  depends significantly on the specific integrand.

\item * So the reader can more easily compare the performance diagnostics,
please state the most relevant hardware specs of the GPU/CPUs used in
Sec 3 (processor speed, number of cores on the GPU, and memory size)

\noindent \textbf{Response}:

\item * Sec 3b introduces many new variables, a handful of which are
undefined. Please make sure all variables have been defined. Better
still, if possible, I believe the readability of this section could be
improved if all of the variables their definitions are placed into a
table.

\noindent \textbf{Response}:

\item * In table 1: why does the setup time reduce when more modes are
included? (comparing the first two rows of this table)

\noindent \textbf{Response}:

\item * The authors focus on computing posteriors of intrinsic parameters.
But for low-latency studies, the extrinsic parameters are arguably
more important as they can be used to direct optical telescopes. Can
the authors please comment on what changes (if any) are needed to do
low-latency studies for producing, for example, sky maps? Is this a
"trivial" change to the code or will it require more significant
re-working? I ask because fully Bayesian low-latency sky maps would be
very useful, so if this code can currently do that then it would be
worth mentioning it more prominently. Otherwise, it would be worth
discussing what additional work will be needed to realize this
promise.

\noindent \textbf{Response}:

\item * Fig 1's caption: what does "on the evaluation points" mean? What are
the evaluation points -- maybe this could be explained in a bit more
detail in the main body of the paper.

\noindent \textbf{Response}:

\item * Figure 1: is the quantity being plotted in the lower left corner
actually the mean of $-log(L_{marg})$? (not the minus sign).

\noindent \textbf{Response}:

\item * Figure 2: same questions I have for figure 1 apply here.

\noindent \textbf{Response}:

\item * Page 7: "Our initial grid consists of 5000 points, spread uniformly
across a 4-dimensional hypercube"... do the authors mean to say $9^4 =
6561$ points?

\noindent \textbf{Response}:

\item * To give the timing numbers (ie the walltime) a bit more context,
please consider adding the time it would take to do a similar analysis
(same recovery model, set of modes, and similar CPU hardware) with
LALInference. Even rough estimates would be very helpful towards
understanding the full impact of the GPU-accelerated rapidPE code. I
feel like this is one of the most important points, as it is the key
to the impact of this work.

\noindent \textbf{Response}:

\item * The fonts in some of the figures, and especially figure 3, are small
and in almost unreadable. Please improve the readability of these
figures.

\noindent \textbf{Response}:

\item * I would appreciate it if the authors could explain to me why the
evidence converges much more quickly in figure 3 as compared to
figures 1 and 2. Or said in a different way, way are the shapes of
evidence vs iteration so different. They may optionally wish to
comment on this in the paper too.

\noindent \textbf{Response}:

\end{enumerate}

We also fixed the following proofreading issues identified by the referee
\begin{itemize}

\item * "Reducing the sampling rate by a factor s will the cost of all
operations with timeseries – they are shorter."

\noindent \textbf{Response}:

\item * "or by marginally more conservative convergence thresholds. on N
eval or N it"

\noindent \textbf{Response}:

\item *"Fortunately, the number of observationally significant and
accessible dimensions is often substantially less than the a prior
necessary dimensionality."

\noindent \textbf{Response}:

\end{itemize}
\end{document}

### 2019-03-20 paper/referee/response-v0.tex
response: continuing

@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}:  We added a sentence about pycbc, a search/parameter inference code which makes extensive
+use of GPUs.    We defer a discussion of non-GPU changes to how lalinference's parallelization is implemented (e.g., MPI in the original
+lalinference paper vs OpenMP) since we feel it would distract the reader to discuss the details how another code harnesses more cores
+simultaneously, particularly since we are not experts in those details (which seem not to be published yet?) and since doing so doesn't reduce its overall comptuational cost.
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}:    \editremark{ADD ME}
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}:  We perform our timing estimates on individual runs on busy nodes and clusters on which we
+have shared access, not
+dedicated hardware.  We anticipate small timing discrepancies from run to run.
+
+GO BACK AND CHECK THESE NUMBERS
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}:  We have added a brief comment in our conclusions to emphasize that extrinsic information
+has been produced with a similar pipeline to the one used here and will be made rapidly available in an automated mode very soon.
+
+We point out however that our inference strategy does not account for the effect of calibration error, which can have a
+notable impact on sky localization: it's not immediately clear our strategy would necessarily provide a dramatic change
+over the Bayestar skymaps, except insofar as we perform inference with a fully precessing signal model (and/or a model
+containing higher modes).
 

### 2019-03-20 paper/referee/response-v0.tex
referee response/paper: minor edits

@@
 organized to loop over detectors so the memory footprint of $Q$ is only (modes)$\times$ (extrinsic)$\times$ (time),
-which is roughly $10^4$ times the footprint of each individual $Q_{k,lm}(\lambda,\tau)$ timeseries.  So many modes can
+which (because we perform evaluations in batches of  $10^4$ extrinsic parameters at a time) is roughly $10^4$ times the
+footprint of all the underlying $Q_{k,lm}(\lambda,\tau)$ timeseries.  So many modes can
 fit well within a typical 4Gb GPU memory space, even allowing for substantial overhead.
@@
 
-GO BACK AND CHECK THESE NUMBERS
+\editremark{GO BACK AND CHECK THESE NUMBERS}
 
@@
 \noindent \textbf{Response}:  We have added a brief comment in our conclusions to emphasize that extrinsic information
-has been produced with a similar pipeline to the one used here and will be made rapidly available in an automated mode very soon.
+has previously been produced with a similar pipeline to the one benchmarked  here,  and that information will be
+made rapidly available in an automated mode by this pipeline very soon.
 
-We point out however that our inference strategy does not account for the effect of calibration error, which can have a
+We point out however that our inference strategy does not presently account for the effect of calibration error, which can have a
 notable impact on sky localization: it's not immediately clear our strategy would necessarily provide a dramatic change
 over the Bayestar skymaps, except insofar as we perform inference with a fully precessing signal model (and/or a model
-containing higher modes).
+containing higher modes).     By contrast, LI inference can account for the impact of calibration error on sky localization.
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}: We modified the caption to define the set whose mean and variance is being evaluated.
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}: No, the sign is correct.  Due to the normalization convention adopted (see response \# 2), the peak value
+of our likelihood (ratio) is greater than zero.
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}: The sign is correct.
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}: We modified the text to add the word ``approximately'' before uniformly.
+
+In practice, our grid layout code returns precisely the number of points requested.  It attempts to choose the sizes of
+each edge of the hypercube to best conform to but be strictly larger than the target size, and then truncates the set of
+candidate points.
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}:  In the conclusions, we \editremark{will} have added some anecdotal apples-to-apples timing
+information on runs of practical interest.
+
+That said, a direct apples-to-apples comparison between RIFT and LI using \emph{identical} waveforms and settings to
+what we benchmark here can be grossly unfair to LI, even though it is an apples-to-apples comparison.  LI developers
+have highly prioritized the use of fast approximations, and reduced-order-quadratures that further accelerate the use of
+these approximations.     To the extent these approximations don't impact results, LI can often perform comparable
+calculations much more quickly than a configuration which used apples-to-apples settings.
+
+Of course, these approximations are approximations.  Existing approximations  have been demonstrated by example to have limitations /
+introduce biases that, if left unchecked, would be increasingly significant as observaitons accumulate (e.g., Lange et
+al 2018).  RIFT's rapid performance  and low cost can enable systematics investigations which would otherwise be inaccessible.
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}:  \editremark{fix me: rcparams}
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}:  The convergence rate of the evidence depends on the system studied and, critically, the
+initial grid: if we have a very well-chosen initial starting grid, the fit  converges rapidly, potentially after just
+one iteration.  The configuration used
+in Fig 3 has a different dimensionality (4d, not 2d) and a larger starting  grid than our lower-dimensionality investigation.
+
+Qualitatively speaking and based on our large but anecdotoally-built experience, we anticipate that  due to strong degeneracy in
+spin -- only one spin degree of freedom is well-constrained -- the seemingly-sparse grid in $\mc$ and $\delta$ is very
+informative because several redundant configurations with similar $\chi_{\rm eff}$ but different $\chi_{i,z}$ inform our
+fit.
+
+A new  student here has begun systematically investigating parameters that impact the convergence rate, using controlled
+circumstances. The results of these investigations are too preliminary to report here.
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}:added ``reduce''
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}: Changed ``. on $N_{eval}$'' to ``$n_{\rm eff}$''
 
@@
 
-\noindent \textbf{Response}:
+\noindent \textbf{Response}: removed ``the'' and ``necessary'' in the last few words
 

### 2019-03-20 paper/paper.tex
referee response/paper: minor edits

@@
 \newcommand\SkipForEarlyCirculation[1]{}
-\newcommand\AddedResponse[1]{\textbf{#1}}
+\newcommand\AddedResponse[1]{{\color{blue} {#1}}}
 %\newcommand\SkipForEarlyCirculation[1]{#1}
@@
 Previously, B. Miller  developed a custom-coded  implementation of the
-relevant likelihood on graphics processing units (GPUs), suggesting substantial performance improvements were possible \cite{Thesis-Brandon-2016,GPU-NU-work}.  
+relevant likelihood on graphics processing units (GPUs), suggesting substantial performance improvements were possible
+\cite{Thesis-Brandon-2016,GPU-NU-work}.  
+\AddedResponse{The pycbc search code\cite{2016CQGra..33u5004U} and its pycbc-inference extension
+  \cite{2019PASP..131b4503B} can also make use of several GPU
+  hardware-accelerated operations.
+}
 In this paper,  we describe a variant of the RIFT approach which flexibly translates one of its
@@
 \unit{Hz}$ for all calculations, but terminated all inner products in Eq. (\ref{eq:overlap}) at roughly $f_{\rm
-  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ will the  cost of all operations with
+  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ will \AddedResponse{reduce} the  cost of all operations with
 timeseries -- they are shorter.  Depending on the relative cost of overhead versus array operations,  $\tau_{setup}$
@@
 minutes on a cluster of (a).  This number can be reduced to
-tens of minutes with a modestly larger GPU pool, or by  marginally more conservative convergence thresholds. on $N_{\rm
-  eval}$ or $N_{it}$.  
+tens of minutes with a modestly larger GPU pool, or by  marginally more conservative convergence thresholds
+ $n_{\rm   eff}$ or $N_{it}$.  
 This discussion ignores the latency introduced by the  gaussian process interpolation stage at the end.  We will revisit
@@
   $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
-\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
+\emph{Bottom left}: Mean and variance of  \AddedResponse{the array $\ln{\cal L}_{\rm marg}(\bm{\lambda}_j)$  for
+$j=1,2,\ldots N_{\rm eval}$ indexing all candidate sets of intrinsic parameters $\bm{\lambda}_j$ performed in that iteration},  showing that after the
 first iteration the
@@
   initial grid and per iteration.  Fortunately, the number of observationally significant and accessible dimensions is often
-  substantially less than the a prior necessary dimensionality.  In practice,we use roughly $20\times$ more points per
+  substantially less than the  prior  dimensionality.  In practice,we use roughly $20\times$ more points per
   iteration for precessing massive BBH systems ($d=8$) or for spinning binary neutron stars with tides ($d=6$).    As a
@@
 As a concrete demonstration of a realistic modest-latency analysis using the GPU-accelerated code, we reanalyze our fiducial binary
-neutron star signal, accounting for the possibility of nonzero (aligned) neutron star spins.  Our initial grid consists of 5000 points, spread uniformly across a 4-dimensional
+neutron star signal, accounting for the possibility of nonzero (aligned) neutron star spins.  Our initial grid consists
+of 5000 points, spread \AddedResponse{approximately} uniformly across a 4-dimensional
 hypercube $\mc_z\in[1.1962,1.1970] M_\odot$, $\delta \in[0,0.25]$ and $\chi_{i,z}\in[-0.05,0.05]$.  Subsequent
@@
 
+\AddedResponse{In this work, we have focused exclusively on a simplified pipeline  to produce posterior distributions
+  for detector-frame intrinsic parameters.   The  code  also produces
+reliable extrinsic parameter distributions \cite{gwastro-PE-AlternativeArchitectures}.  With some postprocessing,
+naturally provides joint posterior distributions for all intrinsic and extrinsic parameter distributions together;
+examples of these distributions have been published \cite{LIGO-O2-Catalog}.  
+Presently, rather than harvest extrinsic information from every iteration, we harvest joint intrinsic and
+extrinsic information with a single final iteration, which we will implement shortly in our production pipeline.
+}
+
 \begin{acknowledgements}

### 2019-03-21 paper/referee/response-v0.tex
referee: response editing

@@
 \newcommand\editremark[1]{{\color{red} #1}}
-
+\newcommand\E[1]{{\left< #1 \right> }}
 \newcommand{\IMRPD}{\textsc{IMRPhenomD}\xspace}
@@
 \noindent \textbf{Response}:  We added a sentence about pycbc, a search/parameter inference code which makes extensive
-use of GPUs.    We defer a discussion of non-GPU changes to how lalinference's parallelization is implemented (e.g., MPI in the original
-lalinference paper vs OpenMP) since we feel it would distract the reader to discuss the details how another code harnesses more cores
-simultaneously, particularly since we are not experts in those details (which seem not to be published yet?) and since doing so doesn't reduce its overall comptuational cost.
+use of GP, and another sentence highlighting code optimizations in lalinference.
+
+We regret that a more extended discussion of non-GPU changes to how lalinference's parallelization is implemented (e.g., MPI in the original
+lalinference paper vs OpenMP)  would distract the reader,  and that a suitable citation to that project specifically
+does not seem available (after consulting with two of the developers responsible).
 
@@
 Approximately equivalently, the stanard deviation of the Monte Carlo error estimate scales as $1/\sqrt{N_{it}}$, with a
-prefactor that  depends significantly on the specific integrand.
+prefactor  that  depends significantly on the specific integrand $f$ (e.g., $\sigma^2= \frac{1}{N}\int (f^2 - \E{f}^2) dx$)
 

### 2019-03-21 paper/referee/response-v0.tex
paper: more on response, providing profiling anecdotes as requested

@@
 \newcommand\editremark[1]{{\color{red} #1}}
-\newcommand\E[1]{{\left< #1 \right> }}
+\newcommand{\E}[1]{{\left\langle #1 \right\rangle }}
 \newcommand{\IMRPD}{\textsc{IMRPhenomD}\xspace}
@@
 Approximately equivalently, the stanard deviation of the Monte Carlo error estimate scales as $1/\sqrt{N_{it}}$, with a
-prefactor  that  depends significantly on the specific integrand $f$ (e.g., $\sigma^2= \frac{1}{N}\int (f^2 - \E{f}^2) dx$)
+prefactor  that  depends significantly on the specific integrand $f$ (e.g., $\sigma^2= \frac{1}{N}\int (f^2 - \E{f}^2) p_s(x)dx$)
 
@@
 
-\noindent \textbf{Response}:    \editremark{ADD ME}
+\noindent \textbf{Response}:   We add an Appendix to describe the hardware used in our study.
 
@@
 
-\noindent \textbf{Response}:  In the conclusions, we \editremark{will} have added some anecdotal apples-to-apples timing
-information on runs of practical interest.
+\noindent \textbf{Response}:  In the conclusions, we  have added anecdotal  timing
+information on an analysis of practical interest.
 
-That said, a direct apples-to-apples comparison between RIFT and LI using \emph{identical} waveforms and settings to
-what we benchmark here can be grossly unfair to LI, even though it is an apples-to-apples comparison.  LI developers
+That said, this anecdote does not employ identical settings in both codes.  A direct apples-to-apples comparison between RIFT and LI using \emph{identical} waveforms and settings to
+what we benchmark here can be grossly unfair to LI.  LI developers
 have highly prioritized the use of fast approximations, and reduced-order-quadratures that further accelerate the use of
 these approximations.     To the extent these approximations don't impact results, LI can often perform comparable
-calculations much more quickly than a configuration which used apples-to-apples settings.
+calculations much more quickly than a LI configuration which used apples-to-apples settings.
 
@@
 Qualitatively speaking and based on our large but anecdotoally-built experience, we anticipate that  due to strong degeneracy in
-spin -- only one spin degree of freedom is well-constrained -- the seemingly-sparse grid in $\mc$ and $\delta$ is very
+spin -- only one spin degree of freedom is well-constrained -- the seemingly-sparse grid in ${\cal M}_c$ and $\delta$ is very
 informative because several redundant configurations with similar $\chi_{\rm eff}$ but different $\chi_{i,z}$ inform our

### 2019-03-21 paper/paper.tex
referee: response editing
paper: more on response, providing profiling anecdotes as requested

@@
 % 2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C
+\AddedResponse{Not published but also important are code optimization projects that improve    infrastructure,
+  such as better parallelization for \texttt{lalinference} \cite{gw-astro-PE-lalinference-v1,lalsuite}.}
  Some methods, however, achieve rapid turnaround through simplifying approximations.  
@@
 neutron star with masses $1.4 M_\odot$ and $1.35 M_\odot$.
- We perform this analysis on heterogeneous LVC collaboration computing resources, to assess our variable
+ We perform this analysis on heterogeneous LVC collaboration computing resources \AddedResponse{described in more detail
+   in the Appendix}, to assess our variable
 performance across architectures.  Notably, we investigate the following GPU options: (a) GTX 1050 Ti at LIGO-CIT,
@@
 
+\AddedResponse{In addition to producing results rapidly, RIFT results can be produced with a noticably smaller overall resource footprint than
+  loosely similar LI analyses, even without tuning to optimize  RIFT pipeline settings.  As a concrete and non-optimized
+  example,  for all five of the iterations of the spinning binary neutron star parameter inference with \texttt{SEOBNRv4\_ROM} from
+  $20\unit{Hz}$ described in this work, our  RIFT analysis expended  roughly 14 core-days.  By contrast, a TaylorF2 analysis of GW170817
+  with LI in MCMC mode starting from $23\unit{Hz}$ required 228 core-days.  % C02-1_TF2LS
+  For precessing binary black holes using SEOBNRv3, the improvement is equally substantial: 10 core-days for a
+  10-iteration investigation of a synthetic GW150914-like source with RIFT, versus 291 core-days for a LI analysis of
+  GW170729.  
+  We defer detailed relative benchmarking using comparably-converged parameter inference to future work.
+}
+
 The overall code performance and thus latency can be further decreased substantially, notably by converting the Monte
@@
 
+\appendix
+\section{Properties of resources used}
+% https://ldas-gridmon.ligo.caltech.edu/condor/
+%
+% > CIT:
+%% > 560 nodes with single GTX 1050 Ti
+%% > 32 nodes with dual K10 PCI cards (shows up as 4 GPU devices) and dual GTX 750 Ti
+%% > 16 nodes with dual GTX 750 Ti (and dual Xeon Phi co-processor)
+%% > 
+%% > LHO and LLO each:
+%% > 40 nodes with GTX 1050 Ti
+%% > single DGX-1 system with 8 tightly coupled V100
+
+\AddedResponse{
+LIGO-CIT worker nodes, denoted by (a)  in the text, are principally S6 nodes with 2-CPU $\times 8$ core Opteron
+2.3 GHz machines with 16 Gb of RAM, with GTX 1050 Ti cards with 4 Gb of RAM.  For LIGO-CIT, profiling reports reflect
+performance averaged over the whole cluster and hence lacks the detailed reporting produced for the other configurations.
+% 
+LIGO-LHO worker nodes with GPUs, denoted by (c) in the text, are a heterogeneous configuration mostly consisting of
+(a).  Unlike profiling at LIGO-CIT, the profiling for LIGO-LHO reflects controlled tests on a single node.
+% nvidia-smi
+The V100 machine (ldas-pcdev13, denoted by (b) in the text) is  a 24-core ES-2650 v4 machine with 4 GPUs, only one of
+which is active in our tests: Tesla V100 with 16 Gb of RAM.
+All non-GPU profiling was  also performed on this machine.  
+}
+
 %\bibstyle{unsrt}

### 2019-03-25 paper/referee/response-v0.tex
referee: basic

@@
 
-\noindent \textbf{Response}:  We perform our timing estimates on individual runs on busy nodes and clusters on which we
+\noindent \textbf{Response}:  
+With only 20 setups (waveform generation plus overlaps) per profiling run,  we anticpate that $\tau_{setup}$ is among the
+least-well-constrained pieces of information in our tables.  Moreover, we perform the vast majority of our timing estimates on individual runs on busy nodes and clusters on which we
 have shared access, not
-dedicated hardware.  We anticipate small timing discrepancies from run to run.
+dedicated hardware, so we can dump (and analyze) detailed profiling output.  We have seen small timing discrepancies from run to run for the setup due to resource
+availability; for example, we also find setup time decreased when going from $\pm 2$ to $\pm 2, \pm 1$ for GPU (b).
+More broadly, the CPU and GPU(b) tests were performed on the same machine and reflect the same hardware and logic
+path, so almost all the variation seen in the $\tau_{setup}$ reflects variability in timing on a single machine.
 
-\editremark{GO BACK AND CHECK THESE NUMBERS}
+As this aspect now has a more significant impact on our performance,  we will delve much more deeply into waveform generation and overlap timing in the near future in later work, as we
+optimize these aspects of the code.
 

### 2019-03-25 demos/convergence_data/plot_mean_variance.py
plot tools: adjust rcparams

@@ -45,14 +45,14 @@ except:
     print " -No 1d kdes- "
 
 dpi_base=200
-legend_font_base=16
+legend_font_base=30
 rc_params = {'backend': 'ps',
-             'axes.labelsize': 11,
-             'axes.titlesize': 10,
-             'font.size': 11,
+             'axes.labelsize': 16,
+             'axes.titlesize': 16,
+             'font.size': 30,
              'legend.fontsize': legend_font_base,
-             'xtick.labelsize': 11,
-             'ytick.labelsize': 11,
+             'xtick.labelsize': 16,
+             'ytick.labelsize': 16,
              #'text.usetex': True,
              'font.family': 'Times New Roman'}#,
              #'font.sans-serif': ['Bitstream Vera Sans']}#,

### 2019-04-06 paper/paper.tex
referee: basic

@@
 \includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_cum.png}
+% python plot_mean_variance.py --convergence-file 20190203-bbh-zerospin-batch_gpu_lowlatency_meanVar.dat  --output bbh_zerospin_lnL_meanVar
 \includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_meanVar.png}
+% python plot_convergence.py --convergence-file 20190203-bbh-zerospin-batch_gpu_lowlatency.dat
 \includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_converge.png}
@@
 \includegraphics[width=0.45\textwidth]{figures/bns_zerospin_mc_cum.png}
+% python plot_mean_variance.py --convergence-file 20190201-bns-zerospin-nobatch_meanVar.dat  --output bns_zerospin_lnL_meanVar
 \includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_meanVar.png}
+% python plot_convergence.py --convergence-file 20190201-bns-zerospin-nobatch.dat --output bns_zerospin_lnL_converge
 \includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_converge.png}
@@
 \includegraphics[width=0.45\textwidth]{figures/bns_withspin_mc_cum.png}
+% python plot_mean_variance.py --convergence-file 20190203-bns-withspin-batch_gpu_lowlatency_noHM_v4ROM_meanVar.dat  --output bns_withspin_lnL_meanVar
 \includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_meanVar.png}
+% python plot_convergence.py --convergence-file 20190203-bns-withspin-batch_gpu_lowlatency_noHM_v4ROM.dat  --output bns_withspin_lnL_converge
 \includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_converge.png}
@@
 
-\AddedResponse{In this work, we have focused exclusively on a simplified pipeline  to produce posterior distributions
+\AddedResponse{In this work, we have focused exclusively on profiling a simplified pipeline  to produce posterior distributions
   for detector-frame intrinsic parameters.   The  code  also produces
 reliable extrinsic parameter distributions \cite{gwastro-PE-AlternativeArchitectures}.  With some postprocessing,
-naturally provides joint posterior distributions for all intrinsic and extrinsic parameter distributions together;
-examples of these distributions have been published \cite{LIGO-O2-Catalog}.  
+this pipeline  provides joint posterior distributions for all intrinsic and extrinsic parameter distributions together;
+examples of these distributions have been published elsewhere \cite{LIGO-O2-Catalog}.  
 Presently, rather than harvest extrinsic information from every iteration, we harvest joint intrinsic and
@@
 \begin{acknowledgements}
+We thank our anonymous referee for helpful feedback.
 DW thanks the RIT COS and CCRG for support, and  NSF-1707965.
